---
title: "scSidekick - Getting Started: PrepObject, ShowColors, GetColors"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{scSidekick - Getting Started: PrepObject, ShowColors, GetColors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 7,
  fig.height = 5
)
options(cli.width = 80)
```

## Overview

scSidekick is built around one idea: **configure your project once, and every
downstream step inherits that configuration automatically** - not just colors
and layouts, but a full reproducibility trail.

After a single setup call, the following come for free as byproducts of ordinary
plotting and analysis:

- **Consistent, publication-ready figures** - one palette and factor ordering
  shared across every plot, single-cell and spatial alike, with no repeated
  `scale_color_manual()` or `group.by` arguments.
- **A plain-text `.legend` sidecar** written next to every saved figure,
  describing exactly what was plotted - so you never reverse-engineer a figure
  at paper time (see *Reporting & Utilities*).
- **A draft methods paragraph** reconstructed from the Seurat command history
  with `ExtractMethods()`.
- **A recoverable session log** - `start_session_logger()` captures the commands
  and figures of a session so an interrupted analysis can be rebuilt, and
  `summarize_r_session()` turns it into a clean report and a runnable R script.
- **A one-call slide deck** - `create_analysis_pptx()` assembles your saved
  figures into a sectioned PowerPoint for the next lab meeting.

`PrepObject()` is the engine that makes all of this possible. Call it once after
clustering and it stores a color palette, factor-level ordering, an output
directory, and default plotting parameters directly inside the Seurat object at
`seurat_object@misc$nk_settings`. Every downstream scSidekick function reads from
that slot when its own arguments are left `NULL` - so you set your project's look
and behavior a single time, and never repeat it. Learn `PrepObject()` here; the
rest of the package builds on it.

**In short:** run `PrepObject()` once -> consistent figures, auto-written
legends, a methods paragraph, a recoverable session log, and a slide deck all
follow automatically.

---

## Installation and demo data

```{r install, eval=FALSE}
if (!require("devtools", quietly = TRUE)) install.packages("devtools")
devtools::install_github("nourabdelfattah/scSidekick")
```

```{r load, echo=FALSE}
library(scSidekick)
library(Seurat)
library(SeuratData)
library(dplyr)
```

We use `bmcite`, a bone marrow CITE-seq dataset from SeuratData. It is small
(~30 k cells) and already comes with rich metadata - `lane`, `donor`,
`celltype.l1`, and `celltype.l2` - so we can demonstrate multi-variable
workflows without manufacturing fake columns.

```{r demo-data, include=FALSE}
data("bmcite")
bmcite <- UpdateSeuratObject(bmcite)
```

---

## 1. Basic `PrepObject` call

`PrepObject()` registers one or more metadata variables and assigns each level
a distinct color from scSidekick's built-in palette. Pass a character vector
to handle all your grouping columns at once.

```{r prep-show, eval=FALSE}
bmcite <- PrepObject(bmcite,
  variables = c("lane", "donor", "celltype.l1", "celltype.l2"))
```

```{r prep-run, include=FALSE}
bmcite <- PrepObject(bmcite,
  variables = c("lane", "donor", "celltype.l1", "celltype.l2"))
```

After this call `bmcite@misc$nk_settings` holds a color map for each
variable. Each variable gets its own independent palette and levels are
stored as ordered factors, so axes and legends always appear in a consistent
order. Inspect them with `ShowColors()` or `GetColors()` (see below).

---

## 2. Setting default plotting parameters

`group.by`, `split.by`, `output_dir`, `object_name`, and `subset_name` store
package-wide defaults that every plotting function inherits when those
arguments are left `NULL`.

```{r prep-defaults, eval=FALSE}
bmcite <- PrepObject(bmcite,
  variables   = c("lane", "donor", "celltype.l1", "celltype.l2"),
  group.by    = "celltype.l1",
  split.by    = "donor",
  output_dir  = "./bmcite_Analysis/Figures",
  object_name = "bmcite",
  subset_name = "AllCells")
```

With these stored, a bare `PlotDimPlots(bmcite)` will group by cell type,
split by donor, and save to the specified folder - no repeated arguments.

---

## 3. Standard Seurat processing

You can add new variables (like `seurat_clusters`) to `PrepObject` later,
after running the clustering pipeline. Here we process the object and use
`Determine_nDims()` to pick the number of dimensions automatically.

```{r standard-setup, include=FALSE}
bmcite <- bmcite %>%
  NormalizeData() %>%
  FindVariableFeatures() %>%
  ScaleData() %>%
  RunPCA()
```

`Determine_nDims()` applies two heuristics to the PCA variance explained and
returns the more conservative (lower) of the two. Blue numbers are the
selected dimensions; orange are discarded.

```{r determine-ndims}
result <- Determine_nDims(bmcite, reduction = "pca")
result$plot
```

```{r cluster, include=FALSE}
bmcite <- RunUMAP(bmcite, dims = 1:result$n_dims)
bmcite <- FindNeighbors(bmcite, dims = 1:result$n_dims)
bmcite <- FindClusters(bmcite, resolution = 0.5)

# Register the new clusters variable
bmcite <- PrepObject(bmcite, variables = "seurat_clusters")
```

---

## 4. Inspect registered colors with `ShowColors()`

`ShowColors()` renders a swatch grid for every variable stored in
`nk_settings` - one panel per variable, levels on the y-axis. This is the
fastest sanity check before generating figures.

```{r show-colors, fig.height=8}
ShowColors(bmcite)
```

---

## 5. Custom colors

Override any variable's palette with a named vector. Any level you omit falls
back to the auto-palette, so you only specify the colors you care about.

```{r custom-colors, eval=FALSE}
my_cols <- c("CD4 T" = "#E41A1C", "CD8 T" = "#377EB8", "NK" = "#4DAF4A",
             "B"     = "#984EA3", "Mono"  = "#FF7F00")

bmcite <- PrepObject(bmcite,
  variables     = "celltype.l1",
  custom_colors = list(celltype.l1 = my_cols))
```

---

## 6. Controlling factor level order

`var_levels` sets the display order for any variable, controlling legend
entries, axis ticks, and split panels. Levels you don't list are appended at
the end.

```{r level-order, eval=FALSE}
bmcite <- PrepObject(bmcite,
  variables  = c("celltype.l1", "donor"),
  var_levels = list(
    celltype.l1 = c("CD4 T", "CD8 T", "NK", "B", "Mono", "DC", "other T"),
    donor       = c("d1", "d2", "d3", "d4", "d5", "d6", "d7", "d8")
  ))
```

---

## 7. Retrieving colors programmatically with `GetColors()`

`GetColors()` returns the named color vector for a variable, ready to pass
into `scale_color_manual()` or any scSidekick `colors` argument.

```{r get-colors}
# Single variable
ct_cols <- GetColors(bmcite, "celltype.l1")
ct_cols

# All registered variables at once
all_cols <- GetColors(bmcite)
names(all_cols)
```

---

## 8. Forcing a rebuild

Once `PrepObject()` has run, subsequent calls skip re-computation unless you
pass `force = TRUE`. Use this after adding new metadata or changing cluster
assignments.

```{r force-rebuild, eval=FALSE}
bmcite$Assignment <- dplyr::case_when(
  bmcite$celltype.l1 == "CD4 T" ~ "Helper T",
  bmcite$celltype.l1 == "CD8 T" ~ "Cytotoxic T",
  TRUE ~ bmcite$celltype.l1
)

bmcite <- PrepObject(bmcite,
  variables = c("celltype.l1", "celltype.l2", "seurat_clusters", "Assignment"),
  force     = TRUE)
```

---

## 9. Recommended full pipeline

This is a reference workflow for a real multi-sample project. Every step uses
a scSidekick function; the Seurat processing steps in the middle are standard.
Nothing here runs - it is a template you copy and adapt.

```{r full-pipeline, eval=FALSE}
library(scSidekick)
library(Seurat)
library(harmony)

# ── 1. Load raw 10x data ─────────────────────────────────────────────────────
SeuratObj <- LoadSamplesRNA(
  sample_dirs  = list.files("path/to/10x_dirs", full.names = TRUE),
  sample_names = c("HC1", "HC2", "AD1", "AD2"),
  metadata     = data.frame(
    Sample = c("HC1", "HC2", "AD1", "AD2"),
    Group  = c("HC",  "HC",  "AD",  "AD"),
    Batch  = c(1, 1, 2, 2)
  )
)

# ── 2. QC ─────────────────────────────────────────────────────────────────────
PlotQCMetrics(SeuratObj, group.by = "Sample", output_dir = "./Figures/QC")
# Inspect the QC plots, then filter:
SeuratObj <- subset(SeuratObj, subset = DoubletStatus == "singlet" & miQC.keep == "keep")

# ── 3. Normalize & reduce ─────────────────────────────────────────────────────
SeuratObj <- NormalizeData(SeuratObj) %>%
  FindVariableFeatures(nfeatures = 2000) %>%
  ScaleData(vars.to.regress = c("S.Score", "G2M.Score", "percent.mt")) %>%
  RunPCA(npcs = 50)

# ── 4. (Optional) Harmony batch correction ────────────────────────────────────
SeuratObj <- harmony::RunHarmony(SeuratObj, group.by.vars = c("Sample", "Batch"))

# ── 5. Choose dimensions ──────────────────────────────────────────────────────
dims <- Determine_nDims(SeuratObj, reduction = "harmony")  # or "pca"
dims$plot   # inspect the elbow

SeuratObj <- RunUMAP(SeuratObj,       dims = 1:dims$n_dims, reduction = "harmony")
SeuratObj <- FindNeighbors(SeuratObj, dims = 1:dims$n_dims, reduction = "harmony")
SeuratObj <- FindClusters(SeuratObj,  resolution = 0.5)

# ── 6. Register colors & defaults ─────────────────────────────────────────────
SeuratObj <- PrepObject(SeuratObj,
  variables   = c("Sample", "Group", "seurat_clusters"),
  group.by    = "seurat_clusters",
  split.by    = "Sample",
  output_dir  = "./Figures",
  object_name = "MyProject")
ShowColors(SeuratObj)

# ── 7. Explore & visualize ────────────────────────────────────────────────────
PlotDimPlots(SeuratObj,
  group.by   = c("seurat_clusters", "Sample", "Group"),
  show_trend = TRUE)

# ── 8. Marker genes & annotation ──────────────────────────────────────────────
GroupHeatmap(SeuratObj, group.by = "seurat_clusters")

markers_df <- data.frame(
  Genes    = c("CD3E", "CD4", "CD8A", "NKG7", "CD79A", "LYZ", "S100A8"),
  CellType = c("T",    "T",   "T",    "NK",   "B",     "Mono","Mono")
)
SplitDotPlot(SeuratObj, markers_df = markers_df, group.by = "seurat_clusters")

de <- RunWilcoxAUC(SeuratObj, group.by = "seurat_clusters")
SeuratObj <- CellTypeAssignmentHelper(SeuratObj,
  markers        = de,
  cluster_column = "seurat_clusters",
  sctype_tissues = "Immune system",
  output_dir     = "./Figures/Annotation")

# ── 9. Finalize labels, update PrepObject ─────────────────────────────────────
SeuratObj$CellType <- dplyr::recode(SeuratObj$seurat_clusters,
  `0` = "CD4 T", `1` = "Monocyte", `2` = "NK", `3` = "B", ...)

SeuratObj <- PrepObject(SeuratObj,
  variables = c("seurat_clusters", "Sample", "Group", "CellType"),
  group.by  = "CellType",
  force     = TRUE)

PlotDimPlots(SeuratObj, group.by = c("CellType", "Sample", "Group"), show_trend = TRUE)

# ── 10. Feature maps ──────────────────────────────────────────────────────────
PlotFeaturePlots(SeuratObj,
  features = c("CD3E", "CD4", "CD8A", "NKG7", "CD79A", "LYZ"),
  split.by = "Sample")

# ── 11. Composition ───────────────────────────────────────────────────────────
PlotTrendLabeled(SeuratObj,
  stat.by  = "CellType",
  group.by = "Group",
  split.by = "Sample")

# ── 12. Pathway analysis ──────────────────────────────────────────────────────
RunGSEA(SeuratObj,
  group.by   = "CellType",
  split.by   = NULL,
  output_dir = "./Figures/Pathways")

# ── 13. Cell-cell communication ───────────────────────────────────────────────
cc_results <- RunCellChat(SeuratObj,
  cell.by    = "CellType",
  group.by   = "Group",
  groups     = c("HC", "AD"),
  output_dir = "./Figures/CellChat")

CompareCellChat(
  cellchat_list = cc_results,
  output_dir    = "./Figures/CellChat/Comparison")

# ── 14. Assemble slide deck ───────────────────────────────────────────────────
create_analysis_pptx(output_dir = "./Figures", object_name = "MyProject")
```

---

## Parameter reference

| Parameter | Type | Default | Description |
|-----------|------|---------|-------------|
| `variables` | character vector | required | `meta.data` column names to register |
| `palettes` | character | `"auto"` | Palette strategy; `"auto"` uses `Nour18` |
| `reverse` | logical | `FALSE` | Reverse color order for auto-assigned palettes |
| `custom_colors` | named list | `list()` | Named color vectors per variable |
| `var_levels` | named list | `list()` | Level orderings per variable |
| `group.by` | character | `NULL` | Default grouping variable |
| `split.by` | character | `NULL` | Default split variable |
| `output_dir` | character | `NULL` | Default output directory |
| `object_name` | character | `NULL` | Label used in saved filenames |
| `subset_name` | character | `NULL` | Sub-label used in saved filenames |
| `force` | logical | `FALSE` | Re-run even if `nk_settings` already exists |

---

## See also

- Vignette 02: `Determine_nDims()` in the context of dimensionality reduction
- Vignette 03: `PlotDimPlots()` - UMAP visualization using colors from `PrepObject`
- Vignette 07: Spatial Transcriptomics - the same settings drive combined
  single-cell and spatial figures (`PlotMasterMaps`, `GenerateMasterGeneMaps`)
- Vignette 12: Reporting & Utilities - the reproducibility layer in full:
  `ExtractMethods()`, the auto-written `.legend` sidecars, the session logger,
  `create_analysis_pptx()`, and the `Nour_pal()` / `Nour_cols()` palette utilities

---

## Session info

```{r session-info}
sessionInfo()
```
