---
title: "scSidekick — Dimensionality Reduction: Determine_nDims"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Dimensionality Reduction: Determine_nDims}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 7,
  fig.height = 5
)
options(cli.width = 80)
```

## Overview

Choosing how many dimensions to retain after PCA (or LSI, ICA, etc.) is one of
the most consequential decisions in a single-cell pipeline. Too few and you lose
biological signal; too many and you add noise.

`Determine_nDims()` automates this using two complementary heuristics and
returns the more **conservative** (lower) of the two — a safe default that you
can override if the plot suggests otherwise.

---

```{r pkgs, include=FALSE}
library(scSidekick)
library(Seurat)
library(SeuratData)
library(dplyr)
```

```{r load-data, include=FALSE}
data("bmcite")
bmcite <- UpdateSeuratObject(bmcite)
bmcite <- PrepObject(bmcite,
  variables   = c("lane", "donor", "celltype.l1", "celltype.l2"),
  group.by    = "celltype.l1",
  object_name = "bmcite")
bmcite <- bmcite %>%
  NormalizeData() %>%
  FindVariableFeatures() %>%
  ScaleData() %>%
  RunPCA(npcs = 50)
```

## 1. Basic call

```{r determine-ndims}
dims <- Determine_nDims(bmcite, reduction = "pca")
```

The console output shows both heuristics and the final selection. The plot
displays every PC as a number — blue = selected, orange = discarded. Vertical
lines mark the two cutoffs.

```{r show-plot}
dims$plot
```

---

## 2. Accessing the result

```{r access-result}
# Recommended number of dimensions
dims$n_dims

# Heuristic 1 — cumulative variance plateau
dims$cum_thresh

# Heuristic 2 — elbow
dims$elbow_thresh
```

`dims$n_dims` is a plain integer: pass it to `dims = 1:dims$n_dims` in all
downstream Seurat calls to guarantee consistency.

---

## 3. Using the result downstream

```{r downstream}
bmcite <- RunUMAP(bmcite,       dims = 1:dims$n_dims)
bmcite <- FindNeighbors(bmcite, dims = 1:dims$n_dims)
bmcite <- FindClusters(bmcite,  resolution = 0.5)
```

After this, `bmcite` has a UMAP embedding and cluster assignments ready for
all visualization functions in vignettes 03–05.

---

## 4. Adjusting thresholds

The defaults (`threshold_cum_pct = 90`, `threshold_pct = 5`) work well for
most datasets. Adjust them if the elbow plot looks unusual.

```{r custom-thresholds, eval=FALSE}
# Stricter: require 85 % cumulative variance, per-dim contribution < 3 %
dims_strict <- Determine_nDims(bmcite,
  threshold_cum_pct = 85,
  threshold_pct     = 3)
dims_strict$n_dims

# More permissive: 95 % cumulative, 1 % marginal drop
dims_loose <- Determine_nDims(bmcite,
  threshold_cum_pct = 95,
  threshold_pct     = 1)
dims_loose$n_dims
```

---

## 5. Using with any linear reduction

The function works with any named reduction slot — PCA, LSI (ATAC), ICA, or
Harmony.

```{r other-reduction, eval=FALSE}
# After RunHarmony():
library(harmony)
bmcite <- RunHarmony(bmcite, group.by.vars = "donor")
dims_h <- Determine_nDims(bmcite, reduction = "harmony")
```

---

## 6. Saving the elbow plot

`dims$plot` is a standard `ggplot2` object.

```{r save-plot, eval=FALSE}
ggplot2::ggsave(
  filename = "bmcite_elbow.pdf",
  plot     = dims$plot,
  width    = 6, height = 4)
```

---

## Parameter reference

| Parameter | Type | Default | Description |
|-----------|------|---------|-------------|
| `seurat_object` | Seurat | required | Object with a linear reduction computed |
| `reduction` | character | `"pca"` | Name of the reduction slot |
| `threshold_pct` | numeric | `5` | Max per-dim variance (%) for a selected dimension |
| `threshold_cum_pct` | numeric | `90` | Target cumulative variance (%) |
| `threshold_diff_pct` | numeric | `0.1` | Min consecutive drop to count as elbow |
| `elbow_plot` | logical | `TRUE` | Return annotated elbow plot |

## Return value

| Element | Type | Description |
|---------|------|-------------|
| `$n_dims` | integer | Recommended number of dimensions |
| `$cum_thresh` | integer | Dimension from heuristic 1 (cumulative plateau) |
| `$elbow_thresh` | integer | Dimension from heuristic 2 (elbow) |
| `$plot` | ggplot | Annotated elbow plot |

---

## See also

- Vignette 01: `PrepObject` — run after clustering to register colors
- Vignette 03: `PlotDimPlots` — UMAP visualization after `RunUMAP(dims = 1:dims$n_dims)`

---

## Session info

```{r session-info}
sessionInfo()
```
