---
title: "scSidekick — UMAP Visualization: PlotDimPlots"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{UMAP Visualization: PlotDimPlots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 10,
  fig.height = 5
)
options(cli.width = 80)
```

## Overview

`PlotDimPlots()` generates a grid of UMAP (or any reduction) scatter plots.

- **`group.by`** controls which metadata column colors each cell
- **`split.by`** creates one column of panels per level (one panel per donor,
  sample, condition, etc.)
- **`row.by`** creates one row of panels per level
- Pass a **vector** to `group.by` and scSidekick iterates, producing one
  complete figure per variable

Colors come from `PrepObject` automatically — no need to pass them manually.

---

```{r pkgs, include=FALSE}
library(scSidekick)
library(Seurat)
library(SeuratData)
library(dplyr)
```

```{r build-bmcite, include=FALSE}
data("bmcite")
bmcite <- UpdateSeuratObject(bmcite)
bmcite <- PrepObject(bmcite,
  variables   = c("lane", "donor", "celltype.l1", "celltype.l2"),
  group.by    = "celltype.l1",
  split.by    = "donor",
  object_name = "bmcite")
bmcite <- bmcite %>%
  NormalizeData() %>%
  FindVariableFeatures() %>%
  ScaleData() %>%
  RunPCA()
dims <- Determine_nDims(bmcite, elbow_plot = FALSE)
bmcite <- RunUMAP(bmcite, dims = 1:dims$n_dims)
bmcite <- FindNeighbors(bmcite, dims = 1:dims$n_dims)
bmcite <- FindClusters(bmcite, resolution = 0.5)
bmcite <- PrepObject(bmcite, variables = "seurat_clusters", force = TRUE)
```

## 1. Basic UMAP by cell type

```{r basic}
PlotDimPlots(bmcite, group.by = "celltype.l1",split.by = NULL)
```

---

## 2. Split across donors

`split.by = "donor"` creates one panel per donor. Each panel shows only that
donor's cells, colored by cell type. The legend is shared.

```{r split-by-donor, fig.height=4}
PlotDimPlots(bmcite,
  group.by = "celltype.l1",
  split.by = "donor")
```

---

## 3. Adding a composition trend panel

`show_trend = TRUE` appends a stacked bar chart to the right showing the
proportional cell-type composition per split group — reads as a combined
spatial + compositional figure.

```{r trend, fig.height=4}
PlotDimPlots(bmcite,
  group.by   = "celltype.l1",
  split.by   = "donor",
  show_trend = TRUE)
```

---

## 4. Cluster labels at centroids

`show_labels = TRUE` overlays the level name at the centroid of each group,
avoiding legend look-up.

```{r labels}
PlotDimPlots(bmcite,
  group.by    = "celltype.l1",
  show_labels = TRUE)
```

---

## 5. Multiple `group.by` — one figure per variable

Pass a character vector to produce one complete figure per grouping variable.
This is the primary way to compare how different annotations map onto the same
embedding.

```{r multi-group, fig.height=4}
p=PlotDimPlots(bmcite,
  group.by = c("celltype.l1", "celltype.l2"))
patchwork::wrap_plots(p,ncol = 1)
```

---

## 6. Row grouping with `row.by`

`row.by` creates a 2-D grid: rows = levels of `row.by`, columns = levels of
`split.by`. Use this to compare two categorical cuts simultaneously.

```{r row-by, eval=FALSE}
PlotDimPlots(bmcite,
  group.by = "celltype.l1",
  row.by = "donor",
  split.by   = "lane")
```

---

## 7. Custom colors

Override `PrepObject` defaults by passing a named vector to `colors`.

```{r custom-colors, eval=FALSE}
my_cols <- c(
  "CD4 T"  = "#003f5c",
  "CD8 T"  = "#227ed4",
  "NK"     = "#a05195",
  "B"      = "#d45087",
  "Mono"   = "#ffa600",
  "DC"     = "#42a665"
)

PlotDimPlots(bmcite,
  group.by = "celltype.l1",
  colors   = my_cols)
```

---

## 8. Controlling column order

`split.col.levels` enforces a specific column order. Useful when donors or
samples have a meaningful biological order (e.g. time points).

```{r col-levels, eval=FALSE}
PlotDimPlots(bmcite,
  group.by         = "celltype.l1",
  split.by         = "donor",
  split.col.levels = c("batch2", "batch1"))
```

---

## 9. Saving to disk

Supply `output_dir` and `object_name` to auto-save every figure as a PDF. The
filename is constructed from `object_name`, `group.by`, and `split.by`.

```{r save, eval=FALSE}
PlotDimPlots(bmcite,
  group.by    = c("celltype.l1", "celltype.l2", "seurat_clusters"),
  split.by    = "donor",
  show_trend  = TRUE,
  output_dir  = "./bmcite_Figures",
  object_name = "bmcite",
  subset_name = "AllCells")
```

---

## Parameter reference

| Parameter | Type | Default | Description |
|-----------|------|---------|-------------|
| `group.by` | character (vector) | from `PrepObject` | Variable(s) used to color cells |
| `split.by` | character (vector) | from `PrepObject` | Variable(s) that create columns |
| `row.by` | character | `NULL` | Variable that creates rows |
| `show_trend` | logical | `FALSE` | Append stacked-bar trend panel |
| `show_labels` | logical | `FALSE` | Overlay group labels at centroids |
| `colors` | named character | `NULL` | Override palette |
| `reduction` | character | `"umap"` | Reduction to plot |
| `pt.size` | numeric | `0.1` | Point size |
| `trend_width` | numeric | `1` | Relative width of the trend panel |
| `split.col.levels` | character vector | `NULL` | Explicit column ordering |
| `output_dir` | character | `NULL` | Directory for saving figures |
| `object_name` | character | `""` | Label used in saved filenames |
| `subset_name` | character | `""` | Sub-label used in saved filenames |

---

## See also

- Vignette 01: `PrepObject` — set default `group.by`, `split.by`, and colors
- Vignette 02: `Determine_nDims` — choose `dims` before running `RunUMAP()`
- Vignette 04: `PlotFeaturePlots` — gene expression overlaid on the same layout
- Vignette 10: `PlotTrendLabeled` — standalone composition figure

---

## Session info

```{r session-info}
sessionInfo()
```
