---
title: "scSidekick - Feature Expression: PlotFeaturePlots & PlotMultiFeature"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Feature Expression: PlotFeaturePlots and PlotMultiFeature}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 10,
  fig.height = 5
)
options(cli.width = 80)
```

## Overview

Two functions for visualizing gene expression on UMAP (or any reduction):

- **`PlotFeaturePlots()`** - one panel per gene × split group. Saves to disk
  by convention; alias `GenerateFeatureMaps()`.
- **`PlotMultiFeature()`** - co-expression of several genes on a single panel
  (gene mode, intersection mode, or continuous scale mode).

---

```{r pkgs, include=FALSE}
library(scSidekick)
library(Seurat)
library(SeuratData)
library(dplyr)
```

```{r build-bmcite, include=FALSE}
data("bmcite")
bmcite <- UpdateSeuratObject(bmcite)
bmcite <- PrepObject(bmcite,
  variables   = c("lane", "donor", "celltype.l1", "celltype.l2"),
  group.by    = "celltype.l1",
  split.by    = "donor",
  object_name = "bmcite")
bmcite <- bmcite %>%
  NormalizeData() %>%
  FindVariableFeatures() %>%
  ScaleData() %>%
  RunPCA()
dims <- Determine_nDims(bmcite, elbow_plot = FALSE)
bmcite <- RunUMAP(bmcite, dims = 1:dims$n_dims)
bmcite <- FindNeighbors(bmcite, dims = 1:dims$n_dims)
bmcite <- FindClusters(bmcite, resolution = 0.5)
bmcite <- PrepObject(bmcite, variables = "seurat_clusters", force = TRUE)
```

## `PlotFeaturePlots`

### 1. Single gene

```{r single-gene}
PlotFeaturePlots(bmcite,
  features = "CD3E")
```

### 2. Multiple genes - one row per gene

```{r multi-gene-rows, fig.height=8}
PlotFeaturePlots(bmcite,
  features = c("CD3E", "CD79A", "LYZ"))
```

### 3. Split by donor - one column per donor

`split.by` creates one column per level. Combined with multiple features, this
produces a gene × donor grid.

> **Pain point fixed.** Seurat's `FeaturePlot(..., split.by =)` renders the
> panels with **no colour legend at all** - you can't tell what the colours
> mean, and the common `& RestoreLegend()` workaround attaches a guide that
> doesn't match the data. `PlotFeaturePlots()` keeps **one shared, accurate
> colour scale** across every panel, so expression is directly comparable
> across conditions.

```{r split-by-donor, eval=FALSE}
PlotFeaturePlots(bmcite,
  features = c("CD3E", "NKG7", "CD79A"),
  split.by = "donor")
```

### 4. Cluster labels overlay

`label = TRUE` prints cell-type labels at cluster centroids, so you can
immediately connect expression hotspots to their identity.

```{r with-labels, fig.height=7}
PlotFeaturePlots(bmcite,
  features = c("CD3E", "CD79A"),
  label    = TRUE,
  label.by = "celltype.l1")
```

### 5. Multiple `split.by` - iterate automatically

Passing a vector to `split.by` generates one complete figure per split variable.

```{r multi-split, eval=FALSE}
PlotFeaturePlots(bmcite,
  features = c("CD3E", "NKG7", "CD79A"),
  split.by = c("lane"),row.by = "donor",layout_method = "metadata")
```

### 6. Saving to disk

```{r save, eval=FALSE, include=FALSE}
PlotFeaturePlots(bmcite,
  features    = c("CD3E", "CD4", "CD8A", "NKG7", "CD79A", "LYZ"),
  split.by    = "donor",
  output_dir  = tempdir(),
  object_name = "bmcite",
  subset_name = "AllCells")
```

---

## `PlotMultiFeature`

### Mode 1 - Gene mode (default)

Each gene gets a distinct color. Cells expressing a gene above `threshold` are
colored accordingly; cells expressing more than one gene get a blended color.
A single UMAP panel shows all lineages at once.

```{r multi-gene-mode}
PlotMultiFeature(bmcite,
  features = c("CD3E", "CD79A", "LYZ", "NKG7") )
```

### Mode 2 - Intersection mode

Highlights only cells that co-express **all** listed genes above the cutoff.
Useful for identifying rare dual-positive populations.

```{r intersection-mode, fig.width=6}
PlotMultiFeature(bmcite,
  features           = c("CD3E", "CD8A"),
  plot_mode          = "intersection",
  pct_cutoff         = 0.5,
  intersection_label = "CD3+CD8+ T cells")
```

### Mode 3 - Scale mode

Plots a single gene (or averaged signature) as a continuous gradient. Equivalent
to `FeaturePlot()` but integrated with the scSidekick saving pipeline.

```{r scale-mode, fig.width=6}
PlotMultiFeature(bmcite,
  features  = "CD3E",
  plot_mode = "scale")
```

---

## Parameter reference - `PlotFeaturePlots`

| Parameter | Type | Default | Description |
|-----------|------|---------|-------------|
| `features` | character vector | required | Genes to plot |
| `split.by` | character (vector) | `NULL` | Variable(s) for column splitting |
| `label` | logical | `FALSE` | Add group labels at centroids |
| `label_by` | character | `NULL` | Metadata column for centroid labels |
| `output_dir` | character | `NULL` | Directory for saving |
| `object_name` | character | `""` | Label used in filenames |
| `subset_name` | character | `""` | Sub-label used in filenames |

## Parameter reference - `PlotMultiFeature`

| Parameter | Type | Default | Description |
|-----------|------|---------|-------------|
| `features` | character vector | required | Genes to co-visualize |
| `plot_mode` | character | `"gene"` | `"gene"`, `"intersection"`, or `"scale"` |
| `colors` | character vector | `NULL` | Colors per gene (gene mode) |
| `threshold` | numeric | `0.0125` | Min expression to count a cell as "expressing" |
| `pct_cutoff` | numeric | `0.3` | Co-expression cutoff (intersection mode) |
| `intersection_label` | character | `NULL` | Label for co-expressing population |
| `output_dir` | character | `NULL` | Directory for saving |

---

## See also

- Vignette 03: `PlotDimPlots` - cell-type UMAP without expression overlay
- Vignette 05: `GroupHeatmap`, `FastDotPlot` - expression across clusters
- Vignette 07: `PlotSpatialFeaturePlots` - same concept for spatial data

---

## Session info

```{r session-info}
sessionInfo()
```
