---
title: "scSidekick - Heatmaps & Dot Plots: GroupHeatmap, SplitDotPlot, FastDotPlot"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Heatmaps and Dot Plots: GroupHeatmap, SplitDotPlot, FastDotPlot}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 9,
  fig.height = 6
)
options(cli.width = 80)
```

## Overview

Three functions for comparing gene expression across groups:

- **`GroupHeatmap()`** - ComplexHeatmap with average expression per group,
  optional dot-size overlay, and gene-category annotations.
- **`SplitDotPlot()`** - dot plot organized by cell-type gene blocks; accepts
  a `markers_df` data frame.
- **`FastDotPlot()`** - quick dot plot from an explicit gene list or regex
  pattern, no `markers_df` required.

---

```{r pkgs, include=FALSE}
library(scSidekick)
library(Seurat)
library(SeuratData)
library(dplyr)
```

```{r build-bmcite, include=FALSE}
data("bmcite")
bmcite <- UpdateSeuratObject(bmcite)
bmcite <- PrepObject(bmcite,
  variables   = c("lane", "donor", "celltype.l1", "celltype.l2"),
  group.by    = "celltype.l1",
  split.by    = "donor",
  object_name = "bmcite")
bmcite <- bmcite %>%
  NormalizeData() %>%
  FindVariableFeatures() %>%
  ScaleData() %>%
  RunPCA()
dims <- Determine_nDims(bmcite, elbow_plot = FALSE)
bmcite <- RunUMAP(bmcite, dims = 1:dims$n_dims)
bmcite <- FindNeighbors(bmcite, dims = 1:dims$n_dims)
bmcite <- FindClusters(bmcite, resolution = 0.5)
bmcite <- PrepObject(bmcite, variables = "seurat_clusters", force = TRUE)
```

```{r marker-genes, include=FALSE}
# Bone marrow CITE-seq markers
bm_markers <- c(
  "CD3E","CD4","CD8A",       # T cells
  "NKG7","GNLY","NCAM1",     # NK
  "CD79A","MS4A1",           # B cells
  "LYZ","S100A8","CD14",     # Monocytes
  "CLEC9A","FCER1A"          # DCs
)

markers_df <- data.frame(
  Genes    = bm_markers,
  CellType = c(
    rep("T cell",    3),
    rep("NK",        3),
    rep("B cell",    2),
    rep("Monocyte",  3),
    rep("DC",        2)
  )
)
```

## `GroupHeatmap`

### 1. Basic heatmap

```{r heatmap-basic}
GroupHeatmap(bmcite,
  group.by = "celltype.l1",
  features = bm_markers)
```

Each column is one cell type; each row is one gene. Color = scaled mean
expression (z-score across groups). Column colors at the top come from
`PrepObject`.

### 2. Gene-category row annotation with `feature_split`

`feature_split` is a named vector (names = genes, values = category labels).
Genes are grouped into labeled row blocks.

```{r heatmap-feature-split, fig.height=7}
feature_split <- setNames(markers_df$CellType, markers_df$Genes)

GroupHeatmap(bmcite,
  group.by      = "celltype.l1",
  features      = names(feature_split),
  feature_split = feature_split)
```

### 3. Dot overlay

`add_dot = TRUE` overlays circles sized by percent-expressed on top of the
color-encoded expression - a dot-heatmap hybrid.

```{r heatmap-dot, fig.height=7}
GroupHeatmap(bmcite,
  group.by = "celltype.l1",
  features = bm_markers,
  add_dot  = TRUE,
  add_bg   = TRUE)
```

### 4. Multiple `group.by`

```{r heatmap-multi-group, eval=FALSE}
GroupHeatmap(bmcite,
  group.by = c("celltype.l1", "seurat_clusters"),
  features = bm_markers)
```

### 5. Saving to disk

```{r heatmap-save, eval=FALSE}
GroupHeatmap(bmcite,
  group.by      = "celltype.l1",
  features      = names(feature_split),
  feature_split = feature_split,
  add_dot       = TRUE,
  output_dir    = "./bmcite_Figures",
  object_name   = "bmcite")
```

---

## `SplitDotPlot`

`SplitDotPlot` accepts a `markers_df` data frame (gene + cell-type column) and
organizes genes into annotated blocks along the x-axis.

### 1. Basic dot plot

```{r splitdot-basic, fig.height=5}
SplitDotPlot(bmcite,
  markers_df = markers_df,
  group.by   = "celltype.l1")
```

Clusters on the x-axis, grouped by their `CellType` block; genes on the y-axis.
Dot color = scaled mean expression; dot size = percent-expressed.

### 2. With `split.by`

Adding `split.by` stacks one dot plot per level, enabling direct expression
comparison across conditions or batches.

```{r splitdot-split, echo=TRUE}
SplitDotPlot(bmcite,
  markers_df = markers_df,
  group.by   = "celltype.l1",
  split.by   = "donor")
```

### 3. `SplitDotPlot2` - with gene clustering

`SplitDotPlot2` adds hierarchical clustering within gene blocks.

```{r splitdot2, echo=TRUE}
SplitDotPlot2(bmcite,
  markers_df    = markers_df,
  group.by      = "celltype.l1",
  ClusterGenes  = TRUE,
  min.pct.exp   = 10)
```

---

## `FastDotPlot`

`FastDotPlot` is a quick path when you don't need cell-type block annotations.

### 1. Explicit gene list

```{r fastdot-list, fig.height=4}
FastDotPlot(bmcite,
  features = c("CD3E", "CD8A", "NKG7", "CD79A", "LYZ", "CD14"),
  group.by = "celltype.l1")
```

### 2. Regex pattern - stop looking up exact gene names

> **Pain point fixed.** You rarely remember every gene symbol in a family.
> Instead of hand-typing `c("CD3D","CD3E","CD3G","CD4","CD8A", ...)`, give
> `FastDotPlot()` a **regex `pattern`** and it pulls every matching gene
> straight from `rownames()`. Same idea powers the pathway *keyword* search in
> Vignette 08 - describe what you want, not its exact name.

`pattern` selects all matching genes; `k_genes` keeps only the top *k* most
variable per cluster.

```{r fastdot-pattern, echo=TRUE}
FastDotPlot(bmcite,
  pattern  = "^CD",          # every gene whose symbol starts with "CD"
  group.by = "celltype.l1",
  k_genes  = 3)
```

A few handy patterns: `"^HLA-"` (MHC genes), `"^IFI|^ISG|^OAS"` (interferon
response), `"^MT-"` (mitochondrial). `FastDotPlot2()` goes further - it
clusters the matched genes into *k* labelled **"Pattern"** groups and can run
enrichR on each, so a loose regex becomes an annotated, organised dot plot.

---

## Parameter reference - `GroupHeatmap`

| Parameter | Type | Default | Description |
|-----------|------|---------|-------------|
| `group.by` | character (vector) | required | Variable(s) defining columns |
| `features` | character vector | required | Genes to display |
| `feature_split` | named character | `NULL` | Gene-to-category mapping |
| `column_split_by` | character | `NULL` | Variable for column grouping |
| `add_dot` | logical | `TRUE` | Overlay dots by percent-expressed |
| `add_bg` | logical | `TRUE` | Background rectangles |
| `output_dir` | character | `NULL` | Directory for saving |

## Parameter reference - `SplitDotPlot`

| Parameter | Type | Default | Description |
|-----------|------|---------|-------------|
| `markers_df` | data.frame | required | Gene-to-cell-type mapping |
| `gene_column` | character | `"Genes"` | Column name for genes |
| `gene_group_column` | character | `"CellType"` | Column name for group labels |
| `group.by` | character | `"seurat_clusters"` | Metadata column for x-axis groups |
| `split.by` | character | `NULL` | Variable for stacked panels |

## Parameter reference - `FastDotPlot`

| Parameter | Type | Default | Description |
|-----------|------|---------|-------------|
| `features` | character vector | `NULL` | Gene list (or use `pattern`) |
| `pattern` | character | `NULL` | Regex to select genes |
| `group.by` | character | `"seurat_clusters"` | X-axis groups |
| `k_genes` | integer | `1` | Top k genes per cluster (pattern mode) |
| `min.pct.exp` | numeric | `15` | Min % expression filter |

---

## See also

- Vignette 04: `PlotFeaturePlots` - per-gene UMAP expression maps
- Vignette 06: `CellTypeAssignmentHelper` - uses markers for annotation
- Vignette 08: `RunGSEA` - pathway-level analysis

---

## Session info

```{r session-info}
sessionInfo()
```
