---
title: "scSidekick - Cell Annotation: CellTypeAssignmentHelper"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Cell Annotation: CellTypeAssignmentHelper}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, message = FALSE, warning = FALSE)
```

## Overview

`CellTypeAssignmentHelper()` runs three independent annotation methods in a
single function call and saves a summary PDF for each:

| Method | Package | Approach |
|--------|---------|----------|
| **scType** | scType | Marker-based scoring against a tissue-specific database |
| **rcellmarker** | rcellmarker | Enrichment against the CellMarker database |
| **SingleR** | SingleR | Reference-based transcriptomic label transfer |

Running all three simultaneously gives you a majority-vote-style sanity check:
when two or three methods agree, the annotation is high-confidence; when they
disagree, the cluster warrants manual inspection.

---

## Standard setup

```{r standard-setup, eval=FALSE}
library(scSidekick)
library(Seurat)
library(SeuratData)
library(dplyr)

data("bmcite")
bmcite <- UpdateSeuratObject(bmcite)
bmcite <- bmcite %>%
  NormalizeData() %>% FindVariableFeatures() %>% ScaleData() %>% RunPCA()
dims <- Determine_nDims(bmcite, elbow_plot = FALSE)
bmcite <- RunUMAP(bmcite, dims = 1:dims$n_dims)
bmcite <- FindNeighbors(bmcite, dims = 1:dims$n_dims)
bmcite <- FindClusters(bmcite, resolution = 0.5)
bmcite <- PrepObject(bmcite,
  variables   = c("lane", "donor", "celltype.l1", "celltype.l2", "seurat_clusters"),
  group.by    = "celltype.l1",
  split.by    = "donor",
  output_dir  = tempdir(),
  object_name = "bmcite")
```

---

## 1. Preparing differential expression markers

`CellTypeAssignmentHelper` requires a pre-computed markers table. The fastest
approach is `presto::wilcoxauc()`, which runs a Wilcoxon rank-sum test for all
clusters simultaneously.

```{r markers-presto, eval=FALSE}
# Fast (recommended): presto
library(presto)
markers <- presto::wilcoxauc(bmcite, group_by = "seurat_clusters")

# Check the structure
head(markers)
# feature | group | logFC | pct_in | pct_out | padj | ...
```

Alternatively, use Seurat's `FindAllMarkers()`:

```{r markers-seurat, eval=FALSE}
markers_seurat <- Seurat::FindAllMarkers(bmcite,
  only.pos      = TRUE,
  min.pct       = 0.25,
  logfc.threshold = 0.25)
```

`CellTypeAssignmentHelper` accepts either format; it auto-detects column names.
The `presto` output is strongly recommended for large datasets because it runs
in seconds rather than minutes.

---

## 2. Full `CellTypeAssignmentHelper` call

```{r annotation-full, eval=FALSE}
bmcite <- CellTypeAssignmentHelper(bmcite,
  markers          = markers,
  cluster_column   = "seurat_clusters",
  output_dir       = tempdir(),
  sctype_tissues   = c("Immune system"),
  run_sctype       = TRUE,
  run_rcellmarker  = TRUE,
  run_singler      = TRUE,
  singler_ref      = "HumanPrimaryCellAtlasData")
```

The function runs all three methods, stores results in the Seurat object's
`meta.data`, and saves one PDF summary figure per method to `output_dir`.
Progress messages are printed to the console for each method.

---

## 3. `sctype_tissues` options

The `sctype_tissues` argument filters the scType database to the tissues most
relevant to your sample. Providing the correct tissue dramatically improves
annotation accuracy.

| Value | Appropriate for |
|-------|----------------|
| `"Immune system"` | PBMCs, spleen, lymph node, bone marrow |
| `"Brain"` | Cortex, hippocampus, cerebellum, CNS |
| `"Liver"` | Hepatocytes, Kupffer cells, stellate cells |
| `"Pancreas"` | Endocrine / exocrine pancreas |
| `"Kidney"` | Proximal tubule, glomerulus, collecting duct |
| `"Lung"` | Airway epithelium, alveoli, endothelium |
| `"Eye"` | Retina, cornea |
| `"Muscle"` | Skeletal and cardiac muscle |

Pass a character vector to query multiple tissues simultaneously:

```{r sctype-tissues, eval=FALSE}
# For a brain tumor dataset: query both brain and immune compartments
bmcite <- CellTypeAssignmentHelper(bmcite,
  markers        = markers,
  cluster_column = "seurat_clusters",
  output_dir     = tempdir(),
  sctype_tissues = c("Brain", "Immune system"))
```

---

## 4. `singler_ref` options

SingleR requires a reference expression dataset. scSidekick exposes the most
commonly used references from the `celldex` package:

| Value | Reference | Best for |
|-------|-----------|----------|
| `"HumanPrimaryCellAtlasData"` | Human Primary Cell Atlas | General human tissue |
| `"BlueprintEncodeData"` | Blueprint + ENCODE | Haematopoietic and stromal |
| `"MonacoImmuneData"` | Monaco Immune | Detailed immune cell types |
| `"ImmGenData"` | ImmGen | Mouse immune cells |
| `"MouseRNAseqData"` | Mouse bulk RNA | General mouse tissue |
| `"NovershternHematopoieticData"` | Novershtern | Detailed human haematopoietic |

```{r singler-ref, eval=FALSE}
# For a detailed immune classification
bmcite <- CellTypeAssignmentHelper(bmcite,
  markers     = markers,
  output_dir  = tempdir(),
  run_sctype  = FALSE,
  run_singler = TRUE,
  singler_ref = "MonacoImmuneData")
```

---

## 5. Running a subset of methods

Set any `run_*` argument to `FALSE` to skip that method. This is useful when
one method is slow or when you only need a quick single-reference annotation.

```{r run-subset, eval=FALSE}
# scType only (fastest, no internet required)
bmcite <- CellTypeAssignmentHelper(bmcite,
  markers         = markers,
  cluster_column  = "seurat_clusters",
  output_dir      = tempdir(),
  sctype_tissues  = "Immune system",
  run_sctype      = TRUE,
  run_rcellmarker = FALSE,
  run_singler     = FALSE)
```

---

## 6. Reading the output PDFs

Each method produces a PDF in `output_dir` named, for example:
`bmcite_scType_annotation.pdf`, `bmcite_rcellmarker_annotation.pdf`,
`bmcite_SingleR_annotation.pdf`.

Each PDF contains:
- A UMAP colored by the predicted cell type label for that method
- A bar chart showing the annotation confidence (scType score, enrichment score,
  or SingleR delta) per cluster
- A table summarizing the top predicted label and the runner-up for each cluster

Use the confidence charts to identify ambiguous clusters: low scores or a
near-tie between the top two candidates warrant manual review.

---

## 7. Accessing results in the Seurat object

After the function returns, three new columns are added to `bmcite@meta.data`:

```{r access-results, eval=FALSE}
# scType predictions
head(bmcite$scType_CellType)

# rcellmarker predictions
head(bmcite$rcellmarker_CellType)

# SingleR predictions
head(bmcite$SingleRAssignment)

# Quick agreement check
table(bmcite$scType_CellType, bmcite$SingleRAssignment)
```

The agreement table is a fast way to spot discordant clusters before
committing to a final label.

---

## 8. Finalizing annotations

After reviewing the three method outputs, assign a final label to a new column
and re-run `PrepObject()` to register it.

```{r final-annotation, eval=FALSE}
# Manually assign based on consensus + prior knowledge
cluster_labels <- c(
  "0" = "CD4 T cell",
  "1" = "CD14 Monocyte",
  "2" = "CD4 T cell",
  "3" = "B cell",
  "4" = "CD8 T cell",
  "5" = "NK cell",
  "6" = "CD16 Monocyte",
  "7" = "Dendritic cell",
  "8" = "Platelet"
)

pbmc$Assignment <- cluster_labels[as.character(bmcite$seurat_clusters)]

# Register the new column - use force = TRUE to update nk_settings
bmcite <- PrepObject(bmcite,
  variables = c("seurat_clusters", "Sample", "Condition", "Assignment"),
  force     = TRUE)

# Confirm
ShowColors(bmcite)
```

`PrepObject()` with `force = TRUE` assigns a distinct color to every unique
cell type in `Assignment` and stores it alongside the existing cluster colors.
All subsequent plots that use `group.by = "Assignment"` will use this palette
automatically.

---

## Parameter reference

| Parameter | Type | Default | Description |
|-----------|------|---------|-------------|
| `seurat_object` | Seurat | required | Clustered Seurat object |
| `markers` | data.frame | required | DE markers (presto or FindAllMarkers output) |
| `cluster_column` | character | `"Cluster"` | Column in `meta.data` holding cluster labels |
| `output_dir` | character | `NULL` | Directory for saving PDF summaries |
| `sctype_tissues` | character vector | `c("Immune system","Brain")` | Tissues to query in scType database |
| `run_sctype` | logical | `TRUE` | Whether to run scType |
| `run_rcellmarker` | logical | `TRUE` | Whether to run rcellmarker |
| `run_singler` | logical | `TRUE` | Whether to run SingleR |
| `singler_ref` | character | `NULL` | celldex reference name for SingleR |

---

## See also

- Vignette 01: `PrepObject` - register the `Assignment` column after annotation
- Vignette 03: `PlotDimPlots` - visualize the final `Assignment` coloring on UMAP
- Vignette 05: `GroupHeatmap` - confirm marker expression aligns with assignment

---

## Session Info

```{r session-info, eval=FALSE}
sessionInfo()
```
