---
title: "scSidekick — Spatial Transcriptomics: PlotSpatialDimPlots, PlotSpatialFeaturePlots, PlotMasterMaps"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Spatial Transcriptomics: PlotSpatialDimPlots, PlotSpatialFeaturePlots, PlotMasterMaps}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

## Overview

scSidekick's three spatial plotting functions mirror their scRNA-seq counterparts
but render expression and metadata directly on the tissue image:

- **`PlotSpatialDimPlots()`** — color spots by a metadata variable (e.g.
  cluster, condition, spatial domain). Long-form alias:
  `GenerateSpatialDimMaps()`.
- **`PlotSpatialFeaturePlots()`** — color spots by gene expression. Long-form
  alias: `GenerateSpatialFeatureMaps()`.
- **`PlotMasterMaps()`** — combined figure: tissue image + gene expression
  panels + cluster-colored UMAP side by side. Long-form alias:
  `GenerateMasterGeneMaps()`.

All three functions accept `group.by`, `split.by`, `layout_method`, and
`output_dir` with the same conventions as the scRNA-seq equivalents.

---

## Demo dataset — mouse brain (Visium)

```{r spatial-setup, eval=FALSE}
library(scSidekick)
library(Seurat)
library(SeuratData)

InstallData("stxBrain")
brain <- LoadData("stxBrain", type = "anterior1")

# Standard processing
brain <- NormalizeData(brain)
brain <- FindVariableFeatures(brain)
brain <- ScaleData(brain)
brain <- RunPCA(brain)
brain <- RunUMAP(brain, dims = 1:30)
brain <- FindNeighbors(brain, dims = 1:30)
brain <- FindClusters(brain, resolution = 0.5)

# Fake condition column for split demos
set.seed(42)
brain$Condition <- sample(c("WT","KO"), ncol(brain), replace = TRUE)

# Register colors
brain <- PrepObject(brain,
  variables   = c("seurat_clusters", "Condition"),
  group.by    = "seurat_clusters",
  output_dir  = tempdir(),
  object_name = "MouseBrain")
```

After `PrepObject()`, the cluster palette is stored and will appear
consistently across all spatial figures.

---

## `PlotSpatialDimPlots`

### 1. Basic spatial cluster map

```{r spatial-dim-basic, eval=FALSE}
PlotSpatialDimPlots(brain,
  group.by = "seurat_clusters")
```

Each spot on the tissue section is colored by its cluster assignment. The
tissue H&E image is shown as a grey-scale background. A legend maps colors
to cluster numbers. The palette comes from `PrepObject`.

### 2. Multiple variables — iterate

Passing a vector to `group.by` produces one spatial map per variable.

```{r spatial-dim-multi, eval=FALSE}
PlotSpatialDimPlots(brain,
  group.by = c("seurat_clusters", "Condition"))
```

Two figures are generated: one showing spatial cluster distribution, one
showing which spots are annotated WT vs KO.

### 3. Uniform spot size

By default Visium spot sizes are scaled by local spot density. Set
`uniform_size = TRUE` to use a fixed, equal size for all spots — useful for
comparing two sections where spot density differs.

```{r spatial-dim-uniform, eval=FALSE}
PlotSpatialDimPlots(brain,
  group.by     = "seurat_clusters",
  uniform_size = TRUE)
```

### 4. Metadata-driven layout with `layout_method`

`layout_method = "metadata"` uses `row_var` to stack multiple spatial maps
vertically, one row per level of `row_var`. Useful when you have multiple
tissue sections or ROIs stored in the same object.

```{r spatial-dim-metadata, eval=FALSE}
PlotSpatialDimPlots(brain,
  group.by      = "seurat_clusters",
  layout_method = "metadata",
  row_var       = "Condition")
```

Each row of the output corresponds to one condition level. Columns would
correspond to levels of `split.by` if provided.

### 5. Colors from `PrepObject` vs custom

```{r spatial-dim-colors, eval=FALSE}
# Colors from PrepObject (default)
PlotSpatialDimPlots(brain,
  group.by = "seurat_clusters")

# Override with a custom palette
my_cols <- GetColors(brain, "seurat_clusters")
my_cols["0"] <- "#FF0000"  # re-color cluster 0

PlotSpatialDimPlots(brain,
  group.by = "seurat_clusters",
  colors   = my_cols)
```

### 6. Saving to disk

```{r spatial-dim-save, eval=FALSE}
PlotSpatialDimPlots(brain,
  group.by    = "seurat_clusters",
  output_dir  = tempdir(),
  object_name = "MouseBrain")
```

---

## `PlotSpatialFeaturePlots`

### 1. Basic spatial gene expression

```{r spatial-feat-basic, eval=FALSE}
PlotSpatialFeaturePlots(brain,
  features = c("Hpca", "Ttr"))
```

Each gene generates one spatial map. Spot color encodes normalized expression
from grey (zero) to the high-expression color (default: dark red). `Hpca`
marks hippocampal neurons; `Ttr` marks the choroid plexus — both should appear
in their expected anatomical locations.

### 2. Metadata-driven layout

```{r spatial-feat-metadata, eval=FALSE}
PlotSpatialFeaturePlots(brain,
  features      = c("Hpca", "Ttr"),
  layout_method = "metadata",
  row_var       = "Condition")
```

Rows correspond to the levels of `Condition`. Each gene's spatial pattern is
shown once per condition. This layout is most useful for comparing expression
between biological groups when each group occupies a distinct tissue section
in the same object.

### 3. Saving to disk

```{r spatial-feat-save, eval=FALSE}
PlotSpatialFeaturePlots(brain,
  features    = c("Hpca", "Ttr"),
  output_dir  = tempdir(),
  object_name = "MouseBrain",
  subset_name = "Anterior1")
```

Files are named by gene: e.g. `MouseBrain_Anterior1_spatial_Hpca.pdf`.

---

## `PlotMasterMaps`

`PlotMasterMaps` is a composite figure function: for each gene it assembles a
multi-panel page containing the tissue spatial map, a UMAP colored by the
specified `group.by`, and a UMAP colored by gene expression — all on one
page, ready for a supplementary figure or lab presentation.

### 1. Basic master map

```{r master-basic, eval=FALSE}
PlotMasterMaps(brain,
  features   = c("Hpca", "Ttr"),
  group.by   = "seurat_clusters",
  output_dir = tempdir())
```

Each gene produces a PDF page with three panels arranged horizontally:
1. Spatial expression map (spot colors = expression level)
2. UMAP colored by `group.by` (cluster) for orientation
3. UMAP colored by gene expression

This lets you simultaneously see where in the tissue the gene is expressed and
which clusters drive that expression.

### 2. Multiple `group.by`

```{r master-multi-group, eval=FALSE}
PlotMasterMaps(brain,
  features   = c("Hpca", "Ttr"),
  group.by   = c("seurat_clusters", "Condition"),
  output_dir = tempdir())
```

Each combination of gene × `group.by` variable produces its own panel page.
With 2 genes and 2 `group.by` variables, you get 4 pages.

### 3. Custom gene expression color scale

```{r master-gene-colors, eval=FALSE}
PlotMasterMaps(brain,
  features    = "Hpca",
  group.by    = "seurat_clusters",
  gene_colors = c("lightgrey", "#006837"),
  output_dir  = tempdir(),
  object_name = "MouseBrain")
```

`gene_colors` is a two-element vector: `[1]` is the zero-expression color,
`[2]` is the high-expression color. The default is grey → dark red.

---

## Parameter reference — `PlotSpatialDimPlots`

| Parameter | Type | Default | Description |
|-----------|------|---------|-------------|
| `group.by` | character (vector) | from `PrepObject` | Variable(s) to color spots by |
| `layout_method` | character | `"auto"` | `"auto"` or `"metadata"` |
| `row_var` | character | `NULL` | Row-grouping variable (metadata layout) |
| `colors` | named character | `NULL` | Override palette |
| `uniform_size` | logical | `FALSE` | Use equal spot size |
| `output_dir` | character | `NULL` | Directory for saving |
| `object_name` | character | `""` | Label used in filenames |

## Parameter reference — `PlotSpatialFeaturePlots`

| Parameter | Type | Default | Description |
|-----------|------|---------|-------------|
| `features` | character vector | required | Genes to plot |
| `layout_method` | character | `"auto"` | `"auto"` or `"metadata"` |
| `row_var` | character | `NULL` | Row-grouping variable |
| `output_dir` | character | `NULL` | Directory for saving |
| `object_name` | character | `""` | Label used in filenames |
| `subset_name` | character | `""` | Sub-label used in filenames |

## Parameter reference — `PlotMasterMaps`

| Parameter | Type | Default | Description |
|-----------|------|---------|-------------|
| `features` | character vector | required | Genes to visualize |
| `group.by` | character (vector) | `"seurat_clusters"` | Variable(s) for the UMAP orientation panel |
| `colors` | named character | `NULL` | Override UMAP cluster colors |
| `gene_colors` | character (length 2) | `c("grey","darkred")` | Zero and max expression colors |
| `output_dir` | character | required | Directory for saving |
| `object_name` | character | `""` | Label used in filenames |

---

## See also

- Vignette 01: `PrepObject` — register spatial cluster colors
- Vignette 03: `PlotDimPlots` — UMAP-only equivalent for scRNA-seq
- Vignette 04: `PlotFeaturePlots` — scRNA-seq gene expression on UMAP

---

## Session Info

```{r session-info, eval=FALSE}
sessionInfo()
```
