---
title: "scSidekick - Cell-Cell Communication: RunCellChat & CompareCellChat"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Cell-Cell Communication: RunCellChat and CompareCellChat}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, message = FALSE, warning = FALSE)
```

## Overview

scSidekick wraps CellChat to provide a streamlined, reproducible cell-cell
communication workflow. Two functions cover the full analysis:

- **`RunCellChat()`** - runs the complete CellChat pipeline for each level of
  a grouping variable (e.g. one CellChat object per condition) and saves
  summary figures.
- **`CompareCellChat()`** - takes a list of pre-computed CellChat objects and
  generates comparative plots (interaction count differences, chord diagrams,
  pathway-level comparisons).

Both functions are designed for multi-condition studies where you want to ask
"which interactions are gained or lost between Control and Treated?"

---

## Standard setup

```{r standard-setup, eval=FALSE}
library(scSidekick)
library(Seurat)
library(SeuratData)
library(CellChat)

library(dplyr)
data("bmcite")
bmcite <- UpdateSeuratObject(bmcite)
bmcite <- bmcite %>%
  NormalizeData() %>% FindVariableFeatures() %>% ScaleData() %>% RunPCA()
dims <- Determine_nDims(bmcite, elbow_plot = FALSE)
bmcite <- RunUMAP(bmcite, dims = 1:dims$n_dims)
bmcite <- FindNeighbors(bmcite, dims = 1:dims$n_dims)
bmcite <- FindClusters(bmcite, resolution = 0.5)
bmcite <- PrepObject(bmcite,
  variables   = c("lane", "donor", "celltype.l1", "celltype.l2", "seurat_clusters"),
  group.by    = "celltype.l1",
  split.by    = "donor",
  output_dir  = tempdir(),
  object_name = "bmcite")
```

Note: for real analyses, each "group" in `group.by` should have at least a few
hundred cells. With the fake `Condition` column here, results are illustrative
only.

---

## `RunCellChat`

### 1. Basic call

`RunCellChat()` splits the Seurat object by the levels of `group.by`, builds
one CellChat object per level, runs the full inference pipeline on each, and
saves summary figures.

```{r cellchat-basic, eval=FALSE}
cc_results <- RunCellChat(bmcite,
  cell.by    = "seurat_clusters",   # cell identity column
  group.by   = "celltype.l1",       # one CellChat run per cell type
  output_dir = tempdir())
```

`cell.by` defines the cell-type identity used for communication inference.
`group.by` defines the groups (e.g. conditions) that are run separately.
The function returns a named list of CellChat objects, one per level of
`group.by`.

### 2. Specifying groups explicitly

Use `groups` to run only a subset of levels, or to fix the order:

```{r cellchat-groups, eval=FALSE}
cc_results <- RunCellChat(bmcite,
  cell.by    = "seurat_clusters",
  group.by   = "Condition",
  groups     = c("Control", "Treated"),
  output_dir = tempdir())
```

### 3. Mouse data

Set `species = "mouse"` to use the CellChatDB.mouse ligand-receptor database:

```{r cellchat-mouse, eval=FALSE}
# brain is the stxBrain object from vignette 07
cc_brain <- RunCellChat(brain,
  cell.by    = "seurat_clusters",
  group.by   = "Condition",
  species    = "mouse",
  output_dir = tempdir())
```

### 4. Custom colors

Pass a named character vector to `colors` to override the cluster palette
used in CellChat chord diagrams and network plots. Colors should be keyed by
the levels of `cell.by`.

```{r cellchat-colors, eval=FALSE}
cluster_cols <- GetColors(bmcite, "seurat_clusters")

cc_results <- RunCellChat(bmcite,
  cell.by    = "seurat_clusters",
  group.by   = "Condition",
  colors     = cluster_cols,
  output_dir = tempdir())
```

### 5. Resuming an interrupted run

`resume = TRUE` checks `output_dir` for already-completed CellChat `.rds`
files and skips those groups, resuming from where the run stopped. Essential
for large datasets.

```{r cellchat-resume, eval=FALSE}
cc_results <- RunCellChat(bmcite,
  cell.by    = "seurat_clusters",
  group.by   = "Condition",
  output_dir = tempdir(),
  resume     = TRUE)
```

### 6. Communication pattern analysis

`run.patterns = TRUE` (default) additionally runs `identifyCommunicationPatterns`
to extract global signaling patterns (incoming and outgoing) using NMF. Set to
`FALSE` to skip this step and save time.

```{r cellchat-patterns, eval=FALSE}
cc_results <- RunCellChat(bmcite,
  cell.by      = "seurat_clusters",
  group.by     = "Condition",
  output_dir   = tempdir(),
  run.patterns = FALSE)   # skip pattern analysis
```

### 7. Outputs produced per group

For each level of `group.by`, `RunCellChat()` saves the following files to
`output_dir`:

| File | Content |
|------|---------|
| `{group}_CellChat.rds` | Serialized CellChat object for downstream use |
| `{group}_interaction_heatmap.pdf` | Heatmap of interaction counts/weights between cell types |
| `{group}_chord_diagram.pdf` | Chord diagram of significant ligand-receptor pairs |
| `{group}_bubble_plot.pdf` | Bubble plot: dot size = p-value, color = communication probability |
| `{group}_pathway_summary.pdf` | Bar chart of top signaling pathways by total interaction strength |
| `{group}_outgoing_patterns.pdf` | (if `run.patterns = TRUE`) Outgoing communication patterns |
| `{group}_incoming_patterns.pdf` | (if `run.patterns = TRUE`) Incoming communication patterns |

---

## `CompareCellChat`

`CompareCellChat()` takes a named list of CellChat objects (one per condition)
and generates comparative visualizations side by side.

> **Pain point fixed.** Two things go wrong natively. (1) When a signalling
> pathway is detected in one cohort but absent in another, the per-pathway plot
> **errors on the cohorts that lack it**, so those panels are silently dropped
> and the *absence* (often the whole point) becomes invisible. (2) Each object
> draws its cell groups in its **own colours and order**, so panels aren't
> comparable position-for-position. `CompareCellChat()` puts **every** cohort on
> one shared node layout with the **same colours and the same group order in
> every panel**, for every pathway present in *any* object - so you compare
> apples to apples. Use `thresh = 1` to also compare pathways that are active at
> the ligand-receptor level but didn't reach pathway-level significance in some
> cohorts.

### 1. Basic comparison

```{r compare-basic, eval=FALSE}
# cc_results is the list returned by RunCellChat()
CompareCellChat(
  cellchat_list = cc_results,
  output_dir = tempdir()
)
```

Pass the list returned directly by `RunCellChat()`. The names of the list
elements (e.g. `"Control"` and `"Treated"`) are used as labels in all plots.

### 2. Loading saved CellChat objects

If you previously ran `RunCellChat()` in a different session, reload the
`.rds` files and build the list manually:

```{r compare-reload, eval=FALSE}
cc_ctrl    <- readRDS(file.path(tempdir(), "Control_CellChat.rds"))
cc_treated <- readRDS(file.path(tempdir(), "Treated_CellChat.rds"))

CompareCellChat(
  cellchat_list = list(Control = cc_ctrl, Treated = cc_treated),
  output_dir = tempdir()
)
```

### 3. What the comparison plots show

`CompareCellChat()` saves the following files to `output_dir`:

| File | What it shows |
|------|---------------|
| `Comparison_interaction_count_bar.pdf` | Total interaction counts per cell type per condition; bars grouped by cell type |
| `Comparison_interaction_strength_bar.pdf` | Same as above but for total interaction strength (probability) |
| `Comparison_differential_heatmap.pdf` | Differential interaction strength (Treated minus Control); red = gained, blue = lost |
| `Comparison_chord_gained.pdf` | Chord diagram showing interactions present or strengthened in the second condition |
| `Comparison_chord_lost.pdf` | Chord diagram showing interactions diminished or lost in the second condition |
| `Comparison_pathway_heatmap.pdf` | Pathway-level NES comparison across all conditions; rows = pathways, columns = conditions |
| `Comparison_scatter_pathway.pdf` | Scatter plot: x = information flow (condition 1), y = information flow (condition 2); deviations from diagonal = changed pathways |

The differential heatmap and scatter plot are the two most commonly cited
figures for condition comparisons in publications.

---

## Typical two-condition workflow

```{r full-workflow, eval=FALSE}
# Step 1: run CellChat for each condition
cc_results <- RunCellChat(bmcite,
  cell.by    = "seurat_clusters",
  group.by   = "Condition",
  groups     = c("Control", "Treated"),
  colors     = GetColors(bmcite, "seurat_clusters"),
  output_dir = file.path(tempdir(), "CellChat"),
  resume     = TRUE)

# Step 2: compare
CompareCellChat(
  cellchat_list = cc_results,
  output_dir = file.path(tempdir(), "CellChat", "Comparison"))
```

---

## Parameter reference - `RunCellChat`

| Parameter | Type | Default | Description |
|-----------|------|---------|-------------|
| `seurat_object` | Seurat | required | Processed, annotated Seurat object |
| `cell.by` | character | required | Metadata column defining cell identity |
| `group.by` | character | required | Metadata column defining groups (one run per level) |
| `groups` | character vector | `NULL` | Subset or reorder levels of `group.by` |
| `species` | character | `"human"` | `"human"` or `"mouse"` |
| `colors` | named character | `NULL` | Override cell-type palette |
| `output_dir` | character | required | Directory for saving outputs |
| `resume` | logical | `FALSE` | Skip groups with existing `.rds` files |
| `run.patterns` | logical | `TRUE` | Run communication pattern NMF analysis |

## Parameter reference - `CompareCellChat`

| Parameter | Type | Default | Description |
|-----------|------|---------|-------------|
| `cellchat_list` | named list | required | Named list of CellChat objects (or paths to `.rds` files) |
| `output_dir` | character | required | Directory for saving comparison plots |
| `groups.show` | character vector | `NULL` | Subset of groups to include in comparison plots |
| `group.merged` | character | `NULL` | Merge these groups before comparison |
| `colors` | named character | `NULL` | Override cell-type palette |
| `species` | character | `"human"` | `"human"` or `"mouse"` |
| `thresh` | numeric | `0.05` | P-value threshold for significant interactions |

---

## See also

- Vignette 01: `PrepObject` - register cell-type colors used in chord diagrams
- Vignette 06: `CellTypeAssignmentHelper` - assign cell types before CellChat
- Vignette 08: `RunGSEA` - pathway-level analysis complementary to CellChat

---

## Session Info

```{r session-info, eval=FALSE}
sessionInfo()
```
