---
title: "scSidekick - Pathway Analysis: RunGSEA_pseudobulk"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Pathway Analysis: Pseudobulk GSEA}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, message = FALSE, warning = FALSE)
```

## Overview

`RunGSEA_pseudobulk()` performs pseudobulk gene set enrichment analysis using
sample-level limma-voom contrasts - a statistically rigorous alternative to the
cell-level Wilcoxon approach in `RunGSEA()`.

The key idea: cells from the same biological sample are aggregated into a
pseudobulk profile before DE testing, so each biological replicate contributes
one observation rather than thousands. This avoids pseudoreplication and produces
calibrated p-values.

**When to use each approach:**

| | `RunGSEA` | `RunGSEA_pseudobulk` |
|---|---|---|
| Input | Any Seurat object | Seurat object with sample metadata |
| DE method | Wilcoxon (cell-level) | limma-voom (sample-level) |
| Replicates needed | No | Yes (>=3 per group) |
| Controls confounders | No | Yes (covariates) |
| Best for | Exploratory; no replicates | Multi-sample studies |

In addition to GSEA, `RunGSEA_pseudobulk()` optionally runs per-patient ssGSEA
(set `run.ssgsea = TRUE`, the default) to produce individual-level pathway
heatmaps showing patient-by-patient variability.

---

## Standard setup

```{r setup-seurat}
library(scSidekick)
library(Seurat)
library(dplyr)

# Assumes a multi-sample Seurat object with:
#   celltype  - annotated cell types
#   sample    - unique biological sample ID (one per donor/timepoint/condition)
#   condition - the comparison variable (e.g. "Control" vs "Disease")
```

---

## 1. Minimal call

The three required metadata columns are `group.by` (cell type), `sample.by`
(sample/donor ID), and `contrast.by` (the condition to compare).

```{r pb-minimal}
res <- RunGSEA_pseudobulk(
  seurat_object = seu,
  group.by      = "celltype",
  sample.by     = "sample",
  contrast.by   = "condition",
  output_dir    = "./Figures/Pathways_PB"
)
```

This runs GSEA for each cell type in `group.by`, comparing all levels of
`condition` against each other with default limma-voom contrasts.

---

## 2. Specifying contrasts

By default, `RunGSEA_pseudobulk` tests all pairwise comparisons within
`contrast.by`. Supply `contrasts` explicitly to control which comparisons are
made, in which direction.

```{r pb-contrasts}
res <- RunGSEA_pseudobulk(
  seurat_object = seu,
  group.by      = "celltype",
  sample.by     = "sample",
  contrast.by   = "condition",
  contrasts     = list(
    Disease_vs_Control = "Disease - Control",
    Treatment_vs_Control = "Treatment - Control"
  ),
  output_dir    = "./Figures/Pathways_PB"
)
```

Each name becomes the label for a set of output figures; each value is a
standard limma contrast expression.

---

## 3. Adding covariates

Correct for batch effects or other sample-level metadata by passing them to
`covariates`. These are included as additional terms in the limma-voom model.

```{r pb-covariates}
res <- RunGSEA_pseudobulk(
  seurat_object = seu,
  group.by      = "celltype",
  sample.by     = "sample",
  contrast.by   = "condition",
  covariates    = c("batch", "sex"),
  output_dir    = "./Figures/Pathways_PB"
)
```

---

## 4. Splitting by broad cell type

Use `split.by` to run the full pipeline independently within each major lineage.
This is the pseudobulk equivalent of the `split.by` parameter in `RunGSEA`.

```{r pb-split}
res <- RunGSEA_pseudobulk(
  seurat_object = seu,
  group.by      = "celltype_fine",
  sample.by     = "sample",
  contrast.by   = "condition",
  split.by      = "celltype_broad",
  output_dir    = "./Figures/Pathways_PB"
)
```

---

## 5. Custom pathway sets

Same syntax as `RunGSEA`. Default: Hallmark, KEGG, Reactome, WikiPathways.

```{r pb-custom-sets}
res <- RunGSEA_pseudobulk(
  seurat_object = seu,
  group.by      = "celltype",
  sample.by     = "sample",
  contrast.by   = "condition",
  pathway_sets  = list(
    Hallmark    = list(category = "H"),
    ImmuneSigDB = list(category = "C7", subcategory = "IMMUNESIGDB")
  ),
  output_dir    = "./Figures/Pathways_PB"
)
```

---

## 6. Using a precomputed pseudobulk

If you have already aggregated pseudobulk counts (e.g. from `ComputePseudobulk`),
pass them directly via `pseudobulk` to skip re-aggregation.

```{r pb-precomputed}
pb <- ComputePseudobulk(seu,
  group.by  = "celltype",
  sample.by = "sample")

res <- RunGSEA_pseudobulk(
  pseudobulk = pb,
  group.by   = "celltype",
  sample.by  = "sample",
  contrast.by = "condition",
  output_dir = "./Figures/Pathways_PB"
)
```

---

## 7. Per-patient ssGSEA heatmaps

`run.ssgsea = TRUE` (default) computes per-patient pathway activity via ssGSEA
after the main GSEA, producing heatmaps where columns are individual samples.
This makes patient-level variability visible alongside the group-level result.

Use `ssgsea.use.padj = TRUE` (default) to filter ssGSEA rows by BH-adjusted
p-value, or `ssgsea.use.padj = FALSE` to use raw p-values.

```{r pb-ssgsea}
res <- RunGSEA_pseudobulk(
  seurat_object  = seu,
  group.by       = "celltype",
  sample.by      = "sample",
  contrast.by    = "condition",
  run.ssgsea     = TRUE,
  ssgsea.use.padj = TRUE,
  top_n_ssgsea   = 20L,
  output_dir     = "./Figures/Pathways_PB"
)
```

---

## 8. Customizing heatmap colors

The same `heatmap_colors` parameter works here as in `RunGSEA`. Pass any
`circlize::colorRamp2` object to override the default 5-stop blue-white-orange-red
palette.

```{r pb-colors}
col_custom <- circlize::colorRamp2(
  c(-3, -1.5, 0, 1.5, 3),
  c("#762a83", "#c2a5cf", "white", "#7fbf7b", "#1b7837")
)

res <- RunGSEA_pseudobulk(
  seurat_object  = seu,
  group.by       = "celltype",
  sample.by      = "sample",
  contrast.by    = "condition",
  heatmap_colors = col_custom,
  output_dir     = "./Figures/Pathways_PB"
)
```

---

## 9. Resuming an interrupted run

`save.rds = TRUE` (default) saves intermediate RDS files after each cell type.
If the run is interrupted, use `redo_plots = TRUE` to reload the saved RDS
files and regenerate all PDFs without re-running the DE analysis.

```{r pb-resume}
# Regenerate all PDFs from saved RDS (no re-computation)
res <- RunGSEA_pseudobulk(
  seurat_object = seu,
  group.by      = "celltype",
  sample.by     = "sample",
  contrast.by   = "condition",
  redo_plots    = TRUE,
  output_dir    = "./Figures/Pathways_PB"
)
```

---

## 10. Return value

`RunGSEA_pseudobulk` returns a nested list, one element per cell type. Each
element contains:

```{r pb-return}
# GSEA results table for one cell type
res$`CD4 T`$gsea_results

# NES heatmap ggplot object
res$`CD4 T`$nes_heatmap

# ssGSEA score matrix (pathways x samples)
res$`CD4 T`$ssgsea_scores
```

---

## Parameter reference

| Parameter | Type | Default | Description |
|-----------|------|---------|-------------|
| `seurat_object` | Seurat or `NULL` | `NULL` | Seurat object (use with `pseudobulk` to skip re-aggregation) |
| `group.by` | character | `NULL` | Cell-type column; one GSEA run per level |
| `sample.by` | character | `NULL` | Sample/donor ID column for pseudobulk aggregation |
| `contrast.by` | character | `NULL` | Condition column for limma contrasts |
| `split.by` | character or `NULL` | `NULL` | Run pipeline independently per level |
| `subset.by` | character or `NULL` | `NULL` | Run pipeline on each subset separately |
| `covariates` | character vector | `NULL` | Sample-level covariates added to limma model |
| `contrasts` | named list | `NULL` | Named limma contrast expressions; `NULL` = all pairwise |
| `pathway_sets` | named list | Hallmark + KEGG + Reactome + WP | MSigDB collections to test |
| `species` | character | `"human"` | Species for `msigdbr` |
| `assay` | character or `NULL` | `NULL` | Seurat assay containing raw counts |
| `min.cells` | integer | `10` | Minimum cells a sample must contribute per cell type |
| `min.samples.per.group` | integer | `3` | Minimum samples per condition level |
| `run.ssgsea` | logical | `TRUE` | Also run per-patient ssGSEA |
| `ssgsea.use.padj` | logical | `TRUE` | Use BH-adjusted p-value for ssGSEA row selection |
| `nes.cutoff` | numeric | `1.0` | Minimum absolute NES for heatmap inclusion |
| `top_n_heatmap` | integer | `30` | Max pathways shown in NES heatmaps |
| `top_n_ssgsea` | integer | `30` | Max pathways shown in ssGSEA heatmaps |
| `heatmap_colors` | `colorRamp2` or `NULL` | `NULL` | Custom heatmap color function |
| `heatmap_params` | named list | `list(...)` | Extra args forwarded to `ComplexHeatmap::Heatmap()` |
| `save.rds` | logical | `TRUE` | Save intermediate RDS after each cell type |
| `redo_plots` | logical | `FALSE` | Reload saved RDS and regenerate PDFs only |
| `caffeinate` | logical | `FALSE` | Prevent system sleep during the run |

---

## See also

- Vignette 08: `RunGSEA` - cell-level Wilcoxon-based GSEA (no replicates needed)
- Vignette 10: `RunSCssGSEA` - per-cell pathway scoring; `PlotPathwayButterfly` - 4-pathway state hierarchy
- Vignette 11: `RunCellChat` - pathway-level cell-cell communication

---

## Session info

```{r session-info}
sessionInfo()
```
