---
title: "scSidekick - Pathway Analysis: Single-Cell Scoring & Enrichment"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Pathway Analysis: Single-Cell Scoring and Enrichment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, message = FALSE, warning = FALSE)
```

## Overview

This vignette covers three complementary approaches to pathway analysis at the
single-cell level:

| Function | What it does |
|---|---|
| `RunSCssGSEA()` | Score every cell against gene sets; test for group differences |
| `PlotPathwayButterfly()` | Map cells onto a 2-D axis defined by exactly 4 pathways |
| `RunEnrichment()` | Over-representation analysis (ORA) on a gene list or DEG table |

These tools answer different questions:

- **RunSCssGSEA**: "Which pathways are differentially active between my cell
  types, at the single-cell level?"
- **PlotPathwayButterfly**: "Where does each cell sit in a 4-pathway state space?
  Do my cell types polarize toward specific states?"
- **RunEnrichment**: "Given my list of significant DEGs, which pathways are
  over-represented?"

---

# Part 1: RunSCssGSEA

## 1.1 Overview

`RunSCssGSEA()` computes a pathway activity score for every cell using either
GSVA's ssGSEA algorithm or UCell, then tests for significant differences
between groups using ANOVA or Kruskal-Wallis. Results are visualized as a
mean z-score heatmap (pathways x groups) and optional per-pathway boxplots.

**Gene-set input modes (pick one):**

| Mode | Parameter | Example |
|---|---|---|
| 1. Custom gene list | `gene_sets` | `list(MySet = c("GENE1","GENE2"))` |
| 2. DEG data frame | `deg_df` | RunWilcoxAUC or FindMarkers output |
| 3. MSigDB collection | `gene_set_library` | `"H"` (Hallmark) |
| 4. MSigDB search | `search_terms` | `c("interferon","apoptosis")` |

Modes 3 and 4 can be combined.

---

## 1.2 Setup

```{r setup-seurat}
library(scSidekick)
library(Seurat)
library(SeuratData)

data("bmcite")
bmcite <- UpdateSeuratObject(bmcite)
bmcite <- PrepObject(bmcite,
  variables   = c("celltype.l1", "celltype.l2", "donor"),
  group.by    = "celltype.l1",
  output_dir  = tempdir(),
  object_name = "bmcite")
```

---

## 1.3 Mode 1: Custom gene sets

Pass a named list of gene vectors. Each element becomes one row in the heatmap.

```{r sc-custom}
RunSCssGSEA(bmcite,
  group.by  = "celltype.l1",
  gene_sets = list(
    T_effector  = c("GZMB", "PRF1", "IFNG", "TNF"),
    T_exhausted = c("PDCD1", "HAVCR2", "LAG3", "TIGIT"),
    NK_cytotoxic = c("NCR1", "KLRB1", "GNLY", "NKG7")
  ),
  output_dir  = tempdir(),
  object_name = "bmcite")
```

---

## 1.4 Mode 2: DEG data frame

Build gene sets automatically from a DEG table. Each group's top DEGs become
one gene set. The `deg_group_column` can be a single column (groups are
heatmap rows) or two columns (the first column is used to split rows on the
heatmap).

```{r sc-deg}
# Get DEGs first
markers <- RunWilcoxAUC(bmcite, group.by = "celltype.l1")

RunSCssGSEA(bmcite,
  group.by        = "celltype.l1",
  deg_df          = markers,
  deg_gene_column = "feature",
  deg_group_column = "group",
  deg_fc_column   = "logFC",
  deg_padj_column = "padj",
  deg_padj_cutoff = 0.05,
  deg_top_n       = 30L,
  output_dir      = tempdir(),
  object_name     = "bmcite")
```

---

## 1.5 Mode 3: MSigDB collection

Fetch an entire MSigDB collection and score all cells against it.
Large collections (e.g. GO:BP with thousands of sets) should be filtered with
`search_terms` or `min.size` to keep the run tractable.

```{r sc-msigdb}
RunSCssGSEA(bmcite,
  group.by         = "celltype.l1",
  gene_set_library = "H",   # Hallmark
  min.size         = 10L,
  output_dir       = tempdir(),
  object_name      = "bmcite")
```

---

## 1.6 Mode 4: MSigDB keyword search

Search gene-set names with case-insensitive regex. Pass a **character vector
for OR** logic, or a **list of vectors for AND-groups**.

```{r sc-search}
# OR: any set mentioning "interferon" OR "apoptosis"
RunSCssGSEA(bmcite,
  group.by      = "celltype.l1",
  search_terms  = c("interferon", "apoptosis"),
  pathway_sets  = list(Hallmark = list(category = "H")),
  output_dir    = tempdir(),
  object_name   = "bmcite")

# AND-group: sets mentioning BOTH "T_CELL" AND "ACTIVATION"
RunSCssGSEA(bmcite,
  group.by     = "celltype.l1",
  search_terms = list(c("T_CELL", "ACTIVATION")),
  pathway_sets = list(
    C7 = list(category = "C7", subcategory = "IMMUNESIGDB")),
  output_dir   = tempdir(),
  object_name  = "bmcite")
```

---

## 1.7 Switching to UCell backend

The default scoring backend is GSVA's ssGSEA. Use `method = "ucell"` for
the UCell rank-based algorithm, which scales better to large datasets and
avoids the GSVA memory overhead.

```{r sc-ucell}
RunSCssGSEA(bmcite,
  group.by         = "celltype.l1",
  gene_set_library = "H",
  method           = "ucell",
  ucell_max_rank   = 1500L,
  output_dir       = tempdir(),
  object_name      = "bmcite")
```

---

## 1.8 Downsampling large datasets

For objects with >50,000 cells, pass `downsample` to cap the number of cells
per group before scoring.

```{r sc-downsample}
RunSCssGSEA(large_obj,
  group.by         = "celltype",
  gene_set_library = "H",
  downsample       = 5000L,
  output_dir       = "./Figures/Pathways")
```

---

## 1.9 Customizing heatmap colors

The default is a 5-stop diverging palette centered at 0 (z = -2 to +2).
Override with any `circlize::colorRamp2` object.

```{r sc-colors}
# Classic blue-white-red
col_bwr <- circlize::colorRamp2(c(-2, -1, 0, 1, 2),
                                 c("#2166ac", "#92c5de", "white",
                                   "#f4a582", "#d6604d"))

RunSCssGSEA(bmcite,
  group.by         = "celltype.l1",
  gene_set_library = "H",
  heatmap_colors   = col_bwr,
  output_dir       = tempdir(),
  object_name      = "bmcite")
```

---

## 1.10 Resuming a run

`resume = TRUE` reloads a previously saved score CSV checkpoint, skipping
re-scoring for the cells already processed.

```{r sc-resume}
RunSCssGSEA(bmcite,
  group.by         = "celltype.l1",
  gene_set_library = "H",
  resume           = TRUE,
  output_dir       = tempdir(),
  object_name      = "bmcite")
```

---

## RunSCssGSEA parameter reference

| Parameter | Type | Default | Description |
|-----------|------|---------|-------------|
| `seurat_object` | Seurat | required | Annotated Seurat object |
| `group.by` | character | required | Metadata column for grouping |
| `split.by` | character or `NULL` | `NULL` | Run independently per level |
| `gene_sets` | named list or `NULL` | `NULL` | Mode 1: custom gene lists |
| `deg_df` | data frame or `NULL` | `NULL` | Mode 2: DEG table |
| `gene_set_library` | character or `NULL` | `NULL` | Mode 3: MSigDB collection code |
| `search_terms` | character/list or `NULL` | `NULL` | Mode 4: keyword filter |
| `pathway_sets` | named list | Hallmark | MSigDB collections for Modes 3/4 |
| `method` | character | `"gsva"` | Scoring backend: `"gsva"` or `"ucell"` |
| `downsample` | integer or `NULL` | `NULL` | Max cells per group |
| `min.size` | integer | `10` | Min gene-set size after intersection |
| `ssgsea.norm` | logical | `FALSE` | Normalize ssGSEA scores to range |
| `ucell_max_rank` | integer | `1500` | UCell only: top genes considered |
| `p_cutoff` | numeric | `0.05` | Adjusted p-value threshold for heatmap rows |
| `show_only_significant` | logical | `TRUE` | Show only significant pathways |
| `heatmap_colors` | `colorRamp2` or `NULL` | `NULL` | Custom color function |
| `heatmap_params` | named list | `list(...)` | Extra args to `ComplexHeatmap::Heatmap()` |
| `add_boxplots` | logical | `TRUE` | Per-pathway boxplots |
| `add_pvalues` | logical | `TRUE` | Add p-values to boxplots |
| `resume` | logical | `FALSE` | Reload checkpoint CSV |
| `caffeinate` | logical | `FALSE` | Prevent system sleep |

---

# Part 2: PlotPathwayButterfly

## 2.1 Overview

`PlotPathwayButterfly()` maps every cell onto a 2-D coordinate system defined
by exactly **four** pathways arranged in quadrants. This makes cell state
polarization visually immediate: cells that are high in the top-left pathway
cluster in the upper-left of the plot; cells sitting at the center are in
transition.

**Quadrant layout:**

```
              top-left | top-right
              ---------|----------
           bottom-left | bottom-right
```

The Y-axis separates the dominant axis (top vs bottom pathways); the X-axis
gives the left-right balance within whichever axis dominates for each cell.

**Output layout:** one large overview panel (all cells colored by `group.by`)
flanked by one small panel per group level (that group highlighted in color;
remaining cells in gray).

This visualization is adapted from the scrabble R package (Suva Lab).

---

## 2.2 Basic call with MSigDB Hallmark pathways

Supply the four pathway names as they appear in the MSigDB collection. The
`pathway_sets` argument specifies which collection to search.

```{r butterfly-basic}
PlotPathwayButterfly(
  seurat_object = bmcite,
  group.by      = "celltype.l1",
  quadrant_order = c(
    "HALLMARK_INTERFERON_GAMMA_RESPONSE",  # bottom-left
    "HALLMARK_INFLAMMATORY_RESPONSE",      # bottom-right
    "HALLMARK_G2M_CHECKPOINT",             # top-left
    "HALLMARK_MYC_TARGETS_V1"             # top-right
  ),
  pathway_sets  = list(Hallmark = list(category = "H")),
  output_dir    = tempdir()
)
```

---

## 2.3 Custom gene sets

When you have your own gene sets (not from MSigDB), pass them directly via
`gene_sets`. The names must match `quadrant_order`.

```{r butterfly-custom}
PlotPathwayButterfly(
  seurat_object  = bmcite,
  group.by       = "celltype.l1",
  gene_sets      = list(
    Cytotoxic    = c("GZMB", "PRF1", "IFNG"),
    NaiveLike    = c("CCR7", "SELL", "TCF7"),
    Exhausted    = c("PDCD1", "LAG3", "HAVCR2"),
    Regulatory   = c("FOXP3", "IL2RA", "CTLA4")
  ),
  quadrant_order = c("NaiveLike", "Cytotoxic", "Regulatory", "Exhausted"),
  output_dir     = tempdir()
)
```

The names in `quadrant_order` map to the four quadrant positions:
**bottom-left, bottom-right, top-left, top-right**.

---

## 2.4 Custom corner labels

By default, corner labels are auto-shortened from the pathway name (prefix and
underscores removed). Override with `quadrant_labels` for cleaner display.

```{r butterfly-labels}
PlotPathwayButterfly(
  seurat_object  = bmcite,
  group.by       = "celltype.l1",
  gene_sets      = list(
    Cytotoxic    = c("GZMB", "PRF1", "IFNG"),
    NaiveLike    = c("CCR7", "SELL", "TCF7"),
    Exhausted    = c("PDCD1", "LAG3", "HAVCR2"),
    Regulatory   = c("FOXP3", "IL2RA", "CTLA4")
  ),
  quadrant_order  = c("NaiveLike", "Cytotoxic", "Regulatory", "Exhausted"),
  quadrant_labels = c("Naive",  "Cytotoxic",    # bl, br
                      "Treg",   "Exhausted"),    # tl, tr
  output_dir      = tempdir()
)
```

---

## 2.5 Split by donor or condition

`split.by` produces a separate figure per split level, each with its own axis
limits. Use this to compare butterfly distributions across donors or conditions.

```{r butterfly-split}
PlotPathwayButterfly(
  seurat_object  = bmcite,
  group.by       = "celltype.l1",
  split.by       = "donor",
  gene_sets      = list(
    Cytotoxic    = c("GZMB", "PRF1", "IFNG"),
    NaiveLike    = c("CCR7", "SELL", "TCF7"),
    Exhausted    = c("PDCD1", "LAG3", "HAVCR2"),
    Regulatory   = c("FOXP3", "IL2RA", "CTLA4")
  ),
  quadrant_order = c("NaiveLike", "Cytotoxic", "Regulatory", "Exhausted"),
  output_dir     = tempdir()
)
```

---

## 2.6 Handling large datasets

For objects with >50,000 cells, enable subsampling to keep scoring fast:

```{r butterfly-subsample}
PlotPathwayButterfly(
  seurat_object     = large_obj,
  group.by          = "celltype",
  quadrant_order    = c("HALLMARK_MYC_TARGETS_V1",
                        "HALLMARK_E2F_TARGETS",
                        "HALLMARK_INTERFERON_GAMMA_RESPONSE",
                        "HALLMARK_INFLAMMATORY_RESPONSE"),
  subsample         = TRUE,
  n_cells           = 30000L,
  output_dir        = "./Figures/Butterfly"
)
```

Without subsampling, the function emits a warning at `subsample_warn_at = 50000`
cells (adjustable).

---

## 2.7 Return value

`PlotPathwayButterfly` invisibly returns a named list (one element per split
level). Each element contains the raw coordinates and scores.

```{r butterfly-return}
res <- PlotPathwayButterfly(
  seurat_object  = bmcite,
  group.by       = "celltype.l1",
  gene_sets      = list(
    Cytotoxic    = c("GZMB", "PRF1", "IFNG"),
    NaiveLike    = c("CCR7", "SELL", "TCF7"),
    Exhausted    = c("PDCD1", "LAG3", "HAVCR2"),
    Regulatory   = c("FOXP3", "IL2RA", "CTLA4")
  ),
  quadrant_order = c("NaiveLike", "Cytotoxic", "Regulatory", "Exhausted")
)

# Hierarchy coordinates (X, Y) for each cell
head(res$All$hierarchy)

# Raw pathway scores (cells x pathways)
head(res$All$scores)
```

---

## PlotPathwayButterfly parameter reference

| Parameter | Type | Default | Description |
|-----------|------|---------|-------------|
| `seurat_object` | Seurat | required | Annotated Seurat object |
| `group.by` | character | required | Metadata column for coloring |
| `quadrant_order` | character(4) | required | Four pathway names: bottom-left, bottom-right, top-left, top-right |
| `split.by` | character or `NULL` | `NULL` | One figure per level |
| `gene_sets` | named list or `NULL` | `NULL` | Custom gene sets; `NULL` = fetch from MSigDB |
| `pathway_sets` | named list | Hallmark | MSigDB collections when `gene_sets = NULL` |
| `species` | character | `"Homo sapiens"` | Species for `msigdbr` |
| `quadrant_labels` | character(4) or `NULL` | `NULL` | Corner display labels; `NULL` = auto-shortened pathway name |
| `group_colors` | named character or `NULL` | `NULL` | Color map for `group.by` levels |
| `assay` | character | `"RNA"` | Assay to extract counts from |
| `layer` | character | `"counts"` | Layer/slot name |
| `subsample` | logical | `FALSE` | Randomly sample cells before scoring |
| `n_cells` | integer | `20000` | Cells to sample when `subsample = TRUE` |
| `center` | logical | `TRUE` | Column-center scores before hierarchy transform |
| `log_scale` | logical | `TRUE` | Apply log2 scaling to hierarchy coordinates |
| `overview_size` | numeric | `2` | Width of overview panel relative to one small panel column |
| `point_size` | numeric | `0.3` | Point size |
| `point_alpha` | numeric | `0.6` | Point alpha |
| `width` | numeric or `NULL` | `NULL` | PDF width (auto-sized when `NULL`) |
| `height` | numeric or `NULL` | `NULL` | PDF height (auto-sized when `NULL`) |
| `output_dir` | character or `NULL` | `NULL` | Save directory; `NULL` = print only |
| `caffeinate` | logical | `FALSE` | Prevent system sleep |

---

# Part 3: RunEnrichment

## 3.1 Overview

`RunEnrichment()` runs over-representation analysis (ORA) via enrichR on a
gene list or on significant DEGs from a data frame. It automatically calls
`PlotEnrichment()` to save dot plots of enriched terms.

ORA is simpler than GSEA but answers a different question: given this gene list,
which pathways contain more of these genes than expected by chance? No ranking
is used - just presence or absence.

---

## 3.2 From a gene list

Pass a character vector of gene symbols directly.

```{r ora-genes}
RunEnrichment(
  features   = c("GZMB", "PRF1", "IFNG", "TNF", "FASLG", "NKG7"),
  databases  = c("MSigDB_Hallmark_2020",
                 "GO_Biological_Process_2023",
                 "KEGG_2021_Human"),
  output_dir = tempdir()
)
```

Run `enrichR::listEnrichrDbs()` for all available database names.

---

## 3.3 From a DEG table

Pass a DEG data frame; `RunEnrichment` filters for significant DEGs before
testing. Columns are auto-detected from common RunWilcoxAUC/FindMarkers naming.

```{r ora-deg}
markers <- RunWilcoxAUC(bmcite, group.by = "celltype.l1")

RunEnrichment(
  deg_df     = markers,
  direction  = "up",         # "up", "down", or "both"
  fc_cutoff  = 0.25,
  padj_cutoff = 0.05,
  databases  = c("MSigDB_Hallmark_2020", "GO_Biological_Process_2023"),
  output_dir = tempdir(),
  object_name = "bmcite"
)
```

---

## 3.4 Per-group enrichment

When the DEG table has a `group_column`, `RunEnrichment` tests each group's
DEGs separately and saves one plot per group.

```{r ora-per-group}
RunEnrichment(
  deg_df       = markers,
  group_column = "group",
  direction    = "up",
  databases    = c("MSigDB_Hallmark_2020"),
  top_n_plot   = 15L,
  output_dir   = tempdir(),
  object_name  = "bmcite"
)
```

---

## 3.5 Choosing databases

The default databases are `MSigDB_Hallmark_2020`, `GO_Biological_Process_2023`,
and `KEGG_2021_Human`. You can test any databases available in enrichR.

```{r ora-databases}
# List all available enrichR databases
enrichR::listEnrichrDbs()

RunEnrichment(
  features  = c("GZMB", "PRF1", "IFNG"),
  databases = c("Reactome_2022",
                "WikiPathway_2023_Human",
                "CellMarker_Augmented_2021"),
  output_dir = tempdir()
)
```

---

## 3.6 Return value

`RunEnrichment` returns a nested list: `[[group]][[database]]` containing
enrichR result data frames.

```{r ora-return}
res <- RunEnrichment(
  deg_df       = markers,
  group_column = "group",
  direction    = "up",
  databases    = c("MSigDB_Hallmark_2020"),
  output_dir   = tempdir()
)

# Enrichment results for one group and database
head(res$`CD4 T`$MSigDB_Hallmark_2020)
```

---

## RunEnrichment parameter reference

| Parameter | Type | Default | Description |
|-----------|------|---------|-------------|
| `features` | character or `NULL` | `NULL` | Gene list for ORA; use with OR `deg_df` |
| `deg_df` | data frame or `NULL` | `NULL` | DEG table; genes filtered before testing |
| `gene_column` | character or `NULL` | `NULL` | Gene column in `deg_df`; auto-detected if `NULL` |
| `fc_column` | character or `NULL` | `NULL` | logFC column; auto-detected if `NULL` |
| `padj_column` | character or `NULL` | `NULL` | Adjusted p-value column; auto-detected if `NULL` |
| `group_column` | character or `NULL` | `NULL` | Group column in `deg_df`; `NULL` = test all together |
| `fc_cutoff` | numeric | `0.25` | Minimum absolute logFC |
| `padj_cutoff` | numeric | `0.05` | Maximum adjusted p-value |
| `direction` | character | `"up"` | Which DEGs: `"up"`, `"down"`, or `"both"` |
| `databases` | character vector | Hallmark + GO:BP + KEGG | enrichR databases to query |
| `top_n_plot` | integer | `20` | Maximum terms per database to plot |
| `plot_type` | character | `"dot"` | Plot type passed to `PlotEnrichment` |
| `padj_cutoff_plot` | numeric | `0.05` | Significance cutoff for plot filtering |
| `output_dir` | character or `NULL` | `NULL` | Save directory; `NULL` = print only |
| `object_name` | character | `""` | Prefix for output file names |
| `subset_name` | character | `""` | Optional subset label for file names |

---

## See also

- Vignette 08: `RunGSEA` - rank-based GSEA on the full Seurat object
- Vignette 09: `RunGSEA_pseudobulk` - sample-level pseudobulk GSEA with limma-voom
- Vignette 05: `GroupHeatmap` / `PlotDotPlot` - gene-level expression visualizations
- Vignette 11: `RunCellChat` - pathway-level cell-cell communication

---

## Session info

```{r session-info}
sessionInfo()
```
