---
title: "scSidekick — Reporting & Utilities"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Reporting and Utilities}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 8,
  fig.height = 5
)
options(cli.width = 80)
```

## Overview

Utility and reporting functions in scSidekick:

- **`GetColors()`** — retrieve a stored color vector by variable name
- **`ShowColors()`** — visual swatch grid of all registered palettes
- **`theme_NourMin()`** — minimal ggplot2 theme consistent with scSidekick figures
- **`PlotTrendLabeled()`** — standalone composition bar chart for manuscripts
- **`PlotComposition()`** — composition bar chart without trend direction
- **`ExtractMethods()`** — auto-generate a methods paragraph from the Seurat command log
- **`create_analysis_pptx()`** — assemble a PowerPoint deck from saved figures

---

```{r pkgs, include=FALSE}
library(scSidekick)
library(Seurat)
library(SeuratData)
library(dplyr)
library(ggplot2)
```

```{r build-bmcite, include=FALSE}
data("bmcite")
bmcite <- UpdateSeuratObject(bmcite)
bmcite <- PrepObject(bmcite,
  variables   = c("lane", "donor", "celltype.l1", "celltype.l2"),
  group.by    = "celltype.l1",
  split.by    = "donor",
  object_name = "bmcite")
bmcite <- bmcite %>%
  NormalizeData() %>%
  FindVariableFeatures() %>%
  ScaleData() %>%
  RunPCA()
dims <- Determine_nDims(bmcite, elbow_plot = FALSE)
bmcite <- RunUMAP(bmcite, dims = 1:dims$n_dims)
bmcite <- FindNeighbors(bmcite, dims = 1:dims$n_dims)
bmcite <- FindClusters(bmcite, resolution = 0.5)
bmcite <- PrepObject(bmcite, variables = "seurat_clusters", force = TRUE)
```

## 1. `GetColors()` — retrieve stored colors

`GetColors()` returns the named color vector for a variable, ready to pass into
`scale_color_manual()` or any scSidekick `colors` argument.

```{r get-colors}
# Single variable
ct_cols <- GetColors(bmcite, "celltype.l1")
ct_cols
```

```{r get-all-colors}
# All registered variables at once
all_cols <- GetColors(bmcite)
names(all_cols)
```

---

## 2. `ShowColors()` — visual swatch grid

`ShowColors()` renders a color tile for every registered variable and level.

```{r show-colors, fig.height=7}
ShowColors(bmcite)
```

---

## 3. `theme_NourMin()` — minimal ggplot2 theme

`theme_NourMin()` removes excess chart junk while keeping axis labels, titles,
and a clean panel border. Layer it after any standard `theme_*()` call.

```{r theme-demo, fig.height=4, fig.width=6}
VlnPlot(bmcite, features = "CD3E", group.by = "celltype.l1", pt.size = 0) +
  theme_NourMin() +
  scale_fill_manual(values = GetColors(bmcite, "celltype.l1")) +
  theme(legend.position = "none")
```

---

## 4. `PlotTrendLabeled()` — composition bar chart

`PlotTrendLabeled()` is a standalone composition figure: one stacked bar per
`group.by` level within each `split.by` panel, with percentage labels on each
segment. Use this for publications; use `PlotDimPlots(show_trend=TRUE)` for
exploratory talks.

```{r trend-labeled, fig.height=5, fig.width=10}
PlotTrendLabeled(bmcite,
  stat.by    = "celltype.l1",
  group.by   = "donor",
  split.by   = "lane",
  label.side = "right")
```

---

## 5. `PlotComposition()` — composition without trend direction

`PlotComposition()` shows proportional composition as a straightforward stacked
bar chart without the directional label emphasis. Good for summarizing
multi-sample distributions.

```{r composition, fig.height=4}
PlotComposition(bmcite,
  group.by = "celltype.l1",
  split.by = "donor")
```

---

## 6. `ExtractMethods()` — auto-generate a methods paragraph

`ExtractMethods()` reads the Seurat command log (`bmcite@commands`) and
reconstructs key analysis steps as a prose paragraph and a structured summary.

```{r extract-methods, eval=FALSE}
methods_out <- ExtractMethods(bmcite, cite_seurat = TRUE)

# Paste into a manuscript
cat(methods_out$methods_text)

# Step-by-step summary (data frame)
print(methods_out$summary)
```

---

## 7. `Nour_pal()` — the built-in palette system

scSidekick ships with `Nour18` (18 qualitative colors) and `Nour20` (20-color
full-spectrum). Both plug into `Nour_pal()`.

```{r palette-demo, fig.width=9, fig.height=3}
# Display all 18 colors
scales::show_col(Nour18, cex_label = 0.6)
```

```{r palette-ggplot, fig.width=6, fig.height=3}
# Use in a custom ggplot
df <- data.frame(x = 1:8, y = 1, grp = LETTERS[1:8])
ggplot(df, aes(x, y, color = grp)) +
  geom_point(size = 6) +
  scale_color_manual(values = Nour_pal("main")(8)) +
  theme_NourMin()
```

---

## 8. `create_analysis_pptx()` — auto-assemble a slide deck

`create_analysis_pptx()` scans `output_dir` for saved PDF/PNG figures from
scSidekick functions, groups them into sections by filename keyword, and writes
a `.pptx` file.

```{r pptx, eval=FALSE}
create_analysis_pptx(
  output_dir  = "./bmcite_Figures",
  object_name = "bmcite")
```

Section grouping rules (keyword in filename → section label):

| Keyword | Section |
|---------|---------|
| `UMAP`, `DimPlot` | Dimensionality Reduction |
| `Feature`, `Expression` | Gene Expression |
| `Heatmap`, `Dot` | Heatmaps & Dot Plots |
| `Spatial` | Spatial |
| `GSEA`, `Pathway` | Pathway Analysis |
| `CellChat`, `Communication` | Cell-Cell Communication |
| `Annotation`, `Assignment` | Cell Annotation |
| anything else | Miscellaneous |

Include or exclude sections by name:

```{r pptx-sections, eval=FALSE}
create_analysis_pptx(
  output_dir       = "./bmcite_Figures",
  object_name      = "bmcite",
  include_sections = c("Dimensionality Reduction", "Gene Expression"))
```

---

## When to use `show_trend` vs `PlotTrendLabeled`

| | `PlotDimPlots(show_trend=TRUE)` | `PlotTrendLabeled()` |
|---|---|---|
| Primary figure | UMAP scatter | Bar chart |
| Trend panel | Appended right of UMAPs | Standalone |
| Labels | None by default | Percentage labels |
| Best for | Talks, exploratory | Manuscripts |

---

## See also

- Vignette 01: `PrepObject`, `ShowColors`, `GetColors` — color management
- Vignette 03: `PlotDimPlots` — UMAP with optional trend panel
- All other vignettes save to `output_dir`, which feeds into `create_analysis_pptx`

---

## Session info

```{r session-info}
sessionInfo()
```
