---
title: "scSidekick - Reporting & Utilities"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Reporting and Utilities}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 8,
  fig.height = 5
)
options(cli.width = 80)
```

## Overview

scSidekick is built so that **the write-up writes itself**. As you analyse, the
package captures the provenance you'll need months later at paper time:

- **Auto-logged figure legends** - every figure-saving function drops a
  plain-text `.legend` sidecar next to the figure, describing exactly what was
  plotted.
- **`ExtractMethods()`** - turns the Seurat command log into a ready-to-paste
  methods paragraph.
- **Passive session logging** - `start_session_logger()` records your commands
  and plots; `summarize_r_session()` turns a messy session into a clean report
  and a tidy R script.
- **`create_analysis_pptx()`** - assembles all your saved figures into a
  sectioned PowerPoint deck in one call.

Plus the everyday utilities: `GetColors()`, `ShowColors()`, `theme_NourMin()`,
`Nour_pal()`, `PlotTrendLabeled()`, and `PlotComposition()`.

---

```{r pkgs, include=FALSE}
library(scSidekick)
library(Seurat)
library(SeuratData)
library(dplyr)
library(ggplot2)
```

```{r build-bmcite, include=FALSE}
data("bmcite")
bmcite <- UpdateSeuratObject(bmcite)
bmcite <- PrepObject(bmcite,
  variables   = c("lane", "donor", "celltype.l1", "celltype.l2"),
  group.by    = "celltype.l1",
  split.by    = "donor",
  object_name = "bmcite")
bmcite <- bmcite %>%
  NormalizeData() %>%
  FindVariableFeatures() %>%
  ScaleData() %>%
  RunPCA()
dims <- Determine_nDims(bmcite, elbow_plot = FALSE)
bmcite <- RunUMAP(bmcite, dims = 1:dims$n_dims)
bmcite <- FindNeighbors(bmcite, dims = 1:dims$n_dims)
bmcite <- FindClusters(bmcite, resolution = 0.5)
bmcite <- PrepObject(bmcite, variables = "seurat_clusters", force = TRUE)
```

## 1. `ExtractMethods()` - a methods paragraph for free

`ExtractMethods()` reads the Seurat command log (`bmcite@commands`) and
reconstructs the key analysis steps as a prose paragraph *and* a structured
summary - no more reverse-engineering your own pipeline at submission time.

```{r extract-methods}
methods_out <- ExtractMethods(bmcite, cite_seurat = TRUE)
cat(methods_out$methods_text)
```

The companion `$summary` is a structured list of the detected steps and their
parameters, handy for a supplementary table.

---

## 2. Auto-logged figure legends

Every scSidekick function that *saves* a figure also writes a `.legend` text
file beside it, generated from the exact arguments used. You never have to
remember what a figure showed.

```{r legend-demo}
fig_dir <- file.path(tempdir(), "figs")
dir.create(fig_dir, showWarnings = FALSE)

# Save a feature map; a PDF and a matching .legend are written:
invisible(GenerateFeatureMaps(bmcite,
  features    = "CD3E",
  output_dir  = fig_dir,
  object_name = "bmcite"))

list.files(fig_dir)
```

```{r legend-read}
# Read back the auto-generated legend:
legend_file <- list.files(fig_dir, pattern = "\\.legend$", full.names = TRUE)
cat(readLines(legend_file), sep = "\n")
```

You can also log legends or analysis parameters yourself with
`log_figure_legend()` and `log_analysis_params()` when you build a custom
figure outside the package.

---

## 3. Passive session logging

`start_session_logger()` installs a lightweight hook that records the commands
you run and the plots you draw for the rest of the session. Afterwards,
`summarize_r_session()` distills the log into a Markdown report and a clean,
runnable R script - turning an exploratory session into reproducible output.

```{r logger, eval=FALSE}
# At the top of an analysis session:
start_session_logger()

# ... do your analysis (every command and plot is captured) ...

# At the end, generate a report + a tidy script:
summarize_r_session(
  output_md = "analysis_report.md",
  output_r  = "analysis_clean.R")
```

`logger_log_file()` and `logger_plot_dir()` tell you where the active log and
captured plots live.

---

## 4. `create_analysis_pptx()` - auto-assemble a slide deck

`create_analysis_pptx()` scans `output_dir` for saved PDF/PNG figures, groups
them into sections by filename keyword, and writes a `.pptx` - your whole
analysis as a deck, in one call.

```{r pptx, eval=FALSE}
create_analysis_pptx(
  output_dir  = "./bmcite_Figures",
  object_name = "bmcite")
```

Section grouping rules (keyword in filename -> section label):

| Keyword | Section |
|---------|---------|
| `UMAP`, `DimPlot` | Dimensionality Reduction |
| `Feature`, `Expression` | Gene Expression |
| `Heatmap`, `Dot` | Heatmaps & Dot Plots |
| `Spatial` | Spatial |
| `GSEA`, `Pathway` | Pathway Analysis |
| `CellChat`, `Communication` | Cell-Cell Communication |
| `Annotation`, `Assignment` | Cell Annotation |
| anything else | Miscellaneous |

Include or exclude sections by name:

```{r pptx-sections, eval=FALSE}
create_analysis_pptx(
  output_dir       = "./bmcite_Figures",
  object_name      = "bmcite",
  include_sections = c("Dimensionality Reduction", "Gene Expression"))
```

---

## 5. `GetColors()` - retrieve stored colors

`GetColors()` returns the named color vector for a variable, ready to pass into
`scale_color_manual()` or any scSidekick `colors` argument.

```{r get-colors}
ct_cols <- GetColors(bmcite, "celltype.l1")
ct_cols
```

```{r get-all-colors}
all_cols <- GetColors(bmcite)
names(all_cols)
```

---

## 6. `ShowColors()` - visual swatch grid

```{r show-colors, fig.height=7}
ShowColors(bmcite)
```

---

## 7. `PlotTrendLabeled()` & `PlotComposition()` - composition figures

`PlotTrendLabeled()` is a publication-ready composition figure with percentage
labels; `PlotComposition()` is the same idea without the directional labels.

```{r trend-labeled, fig.height=5, fig.width=10}
PlotTrendLabeled(bmcite,
  stat.by    = "celltype.l1",
  group.by   = "lane",
  split.by   = "donor",
  label.side = "right")
```

```{r composition, fig.height=4}
PlotComposition(bmcite,
  group.by = "celltype.l1",
  split.by = "donor")
```

---

## 8. `Nour_pal()` - the built-in palette system

scSidekick ships two hand-curated palettes and a unified accessor:

- **`Nour18`** - 18 qualitative colors (blues, purples, pinks, reds, oranges,
  greens). Best for datasets with up to 18 cell types.
- **`Nour20`** - 20 colors sweeping the full visible spectrum. Better for
  atlases with many fine-grained subtypes.
- **`Nour_pal()`** - returns a color-function (like `scale_color_brewer` does
  internally), so you can pull any number of colors from either palette by name.

```{r palette-nour18, fig.width=9, fig.height=3}
# Nour18 - 18 qualitative colors
scales::show_col(Nour18, cex_label = 0.6)
```

```{r palette-nour20, fig.width=10, fig.height=3}
# Nour20 - full-spectrum 20-color palette
scales::show_col(Nour20, cex_label = 0.6)
```

`Nour_pal()` pulls a subset of either palette by name and plugs straight into
`scale_color_manual()` / `scale_fill_manual()`:

```{r palette-ggplot, fig.width=6, fig.height=3}
df <- data.frame(x = 1:8, y = 1, grp = LETTERS[1:8])
ggplot(df, aes(x, y, color = grp)) +
  geom_point(size = 6) +
  scale_color_manual(values = Nour_pal("main")(8)) +
  theme_NourMin()
```

`SelectColors()` lets you interactively pick a subset of colors by name from
`Nour18` - useful when you want to hand-pick specific hues for a figure:

```{r select-colors, eval=FALSE}
# Returns the hex codes for just the colors you name
SelectColors("blue1", "teal", "pink2", "orange")
```

---

## 9. `RenameClusters()` - clean cluster labels

Seurat clusters come out as `0, 1, 2, ...`. `RenameClusters()` converts them
to zero-padded, prefixed labels (`C01`, `C02`, ...) so axes sort correctly and
look publication-ready.

```{r rename-clusters}
# Simulate raw Seurat cluster IDs
raw_clusters <- factor(c(0, 1, 2, 0, 3, 1, 2))

RenameClusters(raw_clusters, prefix = "C", start = 1)
```

Common usage right after `FindClusters()`:

```{r rename-clusters-seurat, eval=FALSE}
bmcite$Cluster <- RenameClusters(bmcite$seurat_clusters, prefix = "C", start = 1)
# seurat cluster 0 -> "C01", cluster 1 -> "C02", etc.
# factor levels are in cluster order, so ggplot axes always sort correctly.

bmcite <- PrepObject(bmcite, variables = "Cluster", group.by = "Cluster")
PlotDimPlots(bmcite, group.by = "Cluster")
```

---

## See also

- Vignette 01: `PrepObject`, `ShowColors`, `GetColors` - color management
- Vignette 11: `PlotMetaSummary()` and the composition chart family
- Every analysis vignette saves to `output_dir` - which feeds straight into
  `create_analysis_pptx()` and produces `.legend` sidecars automatically.

---

## Session info

```{r session-info}
sessionInfo()
```
