---
title: "scSidekick - Cell-Cell Communication: CellChat Suite"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Cell-Cell Communication: CellChat Suite}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, message = FALSE, warning = FALSE)
```

## Overview

scSidekick wraps CellChat to provide a streamlined, reproducible cell-cell
communication workflow. Four functions cover the full pipeline:

| Function | What it does |
|---|---|
| `RunCellChat()` | Run the full CellChat pipeline for each condition; save summary figures |
| `CompareCellChat()` | Generate cross-condition comparison plots from pre-computed CellChat objects |
| `RankCellChatPathways()` | Rank pathways by differential activity; flow heatmap + sender/receiver tiles |
| `RenameCellTypeInCC()` | Rename a cell type label throughout a CellChat object post-hoc |

All functions are designed for multi-condition studies where you want to ask
"which interactions and pathways are gained or lost between Control and Treated?"

> **Note:** scSidekick requires CellChat >= 2.1.0 (the `jinworks/CellChat` fork).
> Install with: `devtools::install_github("jinworks/CellChat")`

---

## Standard setup

```{r standard-setup}
library(scSidekick)
library(Seurat)
library(SeuratData)
library(CellChat)
library(dplyr)

data("bmcite")
bmcite <- UpdateSeuratObject(bmcite)
bmcite <- bmcite %>%
  NormalizeData() %>% FindVariableFeatures() %>% ScaleData() %>% RunPCA()
dims <- Determine_nDims(bmcite, elbow_plot = FALSE)
bmcite <- RunUMAP(bmcite, dims = 1:dims$n_dims)
bmcite <- FindNeighbors(bmcite, dims = 1:dims$n_dims)
bmcite <- FindClusters(bmcite, resolution = 0.5)
bmcite <- PrepObject(bmcite,
  variables   = c("lane", "donor", "celltype.l1", "celltype.l2", "seurat_clusters"),
  group.by    = "celltype.l1",
  split.by    = "donor",
  output_dir  = tempdir(),
  object_name = "bmcite")
```

---

# Part 1: RunCellChat

## 1. Basic call

`RunCellChat()` splits the Seurat object by the levels of `group.by`, builds
one CellChat object per level, runs the full inference pipeline on each, and
saves summary figures.

```{r cellchat-basic}
cc_results <- RunCellChat(bmcite,
  cell.by    = "celltype.l1",   # cell identity column for communication inference
  group.by   = "donor",         # one CellChat run per condition/group
  output_dir = file.path(tempdir(), "CellChat"))
```

`cell.by` defines the cell-type identity used for communication inference.
`group.by` defines the groups run separately (typically conditions). The function
returns a named list of CellChat objects, one per level of `group.by`.

CellChat RDS files are saved in a `CellChatObjects/` subdirectory of
`output_dir` by default.

---

## 2. Specifying groups explicitly

Use `groups` to run only a subset of levels or to fix the comparison order:

```{r cellchat-groups}
cc_results <- RunCellChat(bmcite,
  cell.by    = "celltype.l1",
  group.by   = "Condition",
  groups     = c("Control", "Treated"),
  output_dir = file.path(tempdir(), "CellChat"))
```

---

## 3. Mouse data

Set `species = "mouse"` to use the CellChatDB.mouse ligand-receptor database:

```{r cellchat-mouse}
cc_brain <- RunCellChat(brain,
  cell.by    = "seurat_clusters",
  group.by   = "Condition",
  species    = "mouse",
  output_dir = file.path(tempdir(), "CellChat"))
```

---

## 4. Custom cell-type colors

Pass a named character vector to `colors` to override the palette used in chord
diagrams and network plots. Keys must match levels of `cell.by`.

```{r cellchat-colors}
cluster_cols <- GetColors(bmcite, "celltype.l1")

cc_results <- RunCellChat(bmcite,
  cell.by    = "celltype.l1",
  group.by   = "Condition",
  colors     = cluster_cols,
  output_dir = file.path(tempdir(), "CellChat"))
```

---

## 5. Resuming an interrupted run

`resume = TRUE` checks `robj_dir` for already-completed CellChat `.rds` files
and skips those groups, resuming from where the run stopped.

`redo_plots = TRUE` reloads all saved `.rds` files and regenerates PDFs without
re-running any CellChat inference - useful after updating colors or fixing
cell-type names with `RenameCellTypeInCC`.

```{r cellchat-resume}
# Resume an interrupted run
cc_results <- RunCellChat(bmcite,
  cell.by    = "celltype.l1",
  group.by   = "Condition",
  output_dir = file.path(tempdir(), "CellChat"),
  resume     = TRUE)

# Regenerate all PDFs from saved RDS (no re-inference)
cc_results <- RunCellChat(bmcite,
  cell.by    = "celltype.l1",
  group.by   = "Condition",
  output_dir = file.path(tempdir(), "CellChat"),
  redo_plots = TRUE)
```

---

## 6. Communication pattern analysis

`run.patterns = TRUE` (default) additionally runs `identifyCommunicationPatterns`
to extract global signaling patterns (incoming and outgoing) using NMF.
`pattern_k_range` controls which values of k are tested; the best k is selected
automatically.

```{r cellchat-patterns}
# Skip pattern analysis to save time
cc_results <- RunCellChat(bmcite,
  cell.by        = "celltype.l1",
  group.by       = "Condition",
  output_dir     = file.path(tempdir(), "CellChat"),
  run.patterns   = FALSE)

# Custom k range
cc_results <- RunCellChat(bmcite,
  cell.by        = "celltype.l1",
  group.by       = "Condition",
  output_dir     = file.path(tempdir(), "CellChat"),
  run.patterns   = TRUE,
  pattern_k_range = 2:6)
```

---

## 7. Outputs produced per group

For each level of `group.by`, `RunCellChat()` saves the following files:

| File | Content |
|------|---------|
| `CellChatObjects/{group}_CellChat.rds` | Serialized CellChat object |
| `{group}_interaction_heatmap.pdf` | Heatmap of interaction counts/weights |
| `{group}_chord_diagram.pdf` | Chord diagram of significant L-R pairs |
| `{group}_bubble_plot.pdf` | Dot size = p-value, color = probability |
| `{group}_pathway_summary.pdf` | Bar chart of top signaling pathways |
| `{group}_outgoing_patterns.pdf` | Outgoing communication patterns (if `run.patterns`) |
| `{group}_incoming_patterns.pdf` | Incoming communication patterns (if `run.patterns`) |

---

# Part 2: CompareCellChat

`CompareCellChat()` takes a named list of CellChat objects (one per condition)
and generates comparative visualizations side by side.

> **Pain point fixed.** Two things go wrong natively. (1) When a signaling
> pathway is detected in one condition but absent in another, the per-pathway
> plot errors on conditions that lack it, so those panels are silently dropped
> and the *absence* becomes invisible. (2) Each object draws its cell groups in
> its own colors and order, so panels are not comparable position-for-position.
> `CompareCellChat()` puts every condition on a shared node layout with the
> same colors and the same group order in every panel, for every pathway present
> in *any* object. Use `thresh = 1` to also compare pathways that are active at
> the L-R level but did not reach pathway-level significance in some conditions.

## 1. Basic comparison

```{r compare-basic}
# cc_results is the named list returned by RunCellChat()
CompareCellChat(
  cellchat_list = cc_results,
  output_dir    = file.path(tempdir(), "CellChat", "Comparison"))
```

---

## 2. Loading saved CellChat objects

If you previously ran `RunCellChat()` in a different session, reload the
`.rds` files and build the list manually:

```{r compare-reload}
cc_ctrl    <- readRDS(file.path(tempdir(), "CellChat", "CellChatObjects",
                                "Control_CellChat.rds"))
cc_treated <- readRDS(file.path(tempdir(), "CellChat", "CellChatObjects",
                                "Treated_CellChat.rds"))

CompareCellChat(
  cellchat_list = list(Control = cc_ctrl, Treated = cc_treated),
  output_dir    = file.path(tempdir(), "CellChat", "Comparison"))
```

---

## 3. What the comparison plots show

| File | What it shows |
|------|---------------|
| `Comparison_interaction_count_bar.pdf` | Interaction counts per cell type per condition |
| `Comparison_interaction_strength_bar.pdf` | Interaction strength (probability) per cell type |
| `Comparison_differential_heatmap.pdf` | Differential strength (Treated - Control); red = gained, blue = lost |
| `Comparison_chord_gained.pdf` | Interactions gained or strengthened in the second condition |
| `Comparison_chord_lost.pdf` | Interactions diminished or lost in the second condition |
| `Comparison_pathway_heatmap.pdf` | Pathway-level NES across all conditions |
| `Comparison_scatter_pathway.pdf` | x = flow (condition 1), y = flow (condition 2); deviations = changed pathways |
| `{pathway}_comparison.pdf` | Per-pathway: circle plots, chord diagrams, and heatmaps for all conditions |

---

# Part 3: RankCellChatPathways

`RankCellChatPathways()` goes one level deeper than `CompareCellChat`: it ranks
pathways by how much their information flow changes across conditions, then
decomposes those changes into which cell types are driving them as senders and
receivers.

**Four outputs:**

1. **Flow heatmap** - top N pathways x conditions, row-normalized so color
   shows which condition has relatively higher activity
2. **Sender tile plot** - faceted by pathway; shows how much each cell type
   sends per condition
3. **Receiver tile plot** - same for receivers
4. **Top changed links heatmap** - the sender-receiver pairs with the highest
   variance across conditions, showing exactly which cell-type interactions
   drive pathway-level changes
5. **CSV** - all underlying metrics for custom analysis

## 1. Basic call

```{r rank-basic}
rank_results <- RankCellChatPathways(
  cellchat_list = cc_results,
  output_dir    = file.path(tempdir(), "CellChat", "Ranked"))
```

---

## 2. Controlling which pathways and links to show

`top_n` controls how many pathways appear in the flow heatmap and tile plots.
`top_links` controls how many sender-receiver pairs appear in the links heatmap.
Both are ranked by differential activity (`rank_by = "range"` or `"cv"`).

```{r rank-params}
rank_results <- RankCellChatPathways(
  cellchat_list = cc_results,
  output_dir    = file.path(tempdir(), "CellChat", "Ranked"),
  top_n         = 30L,
  top_links     = 20L,
  rank_by       = "cv")    # coefficient of variation instead of range
```

---

## 3. Custom colors

Supply condition colors (for the flow heatmap strips) and cell-type colors
(for the sender/receiver tile annotation strips):

```{r rank-colors}
cond_colors <- c(Control = "#4575b4", Treated = "#d73027")
cell_colors <- GetColors(bmcite, "celltype.l1")

rank_results <- RankCellChatPathways(
  cellchat_list = cc_results,
  output_dir    = file.path(tempdir(), "CellChat", "Ranked"),
  colors        = cond_colors,
  cell_colors   = cell_colors)
```

---

## 4. Return value

```{r rank-return}
# Data frame: pathway x condition flow + ranking columns
head(rank_results$flow_table)

# Long data frame: which cell types send each pathway per condition
head(rank_results$sender_table)

# Long data frame: which cell types receive each pathway per condition
head(rank_results$receiver_table)

# Long data frame: sender-receiver pair flow per condition
head(rank_results$links_table)

# ggplot2 / pheatmap objects
rank_results$flow_heatmap
rank_results$sender_plot
rank_results$receiver_plot
rank_results$links_plot
```

---

# Part 4: RenameCellTypeInCC

If you rename a cluster or cell type after running CellChat (e.g. to correct
an annotation mistake), `RenameCellTypeInCC()` patches the CellChat object so
the new name propagates to all internal arrays: the probability matrices,
centrality scores, communication pattern tables, and LR significance tables.

Communication probabilities are **not** recomputed - only labels change. This
is safe because the underlying values are invariant to what the cell type is
called.

## 1. Rename in a single object

```{r rename-single}
cc_fixed <- RenameCellTypeInCC(cc_ctrl,
  old_name = "Immune_contam",
  new_name  = "NK_TCells")

# Save the fixed object back
saveRDS(cc_fixed,
  file.path(tempdir(), "CellChat", "CellChatObjects", "Control_CellChat.rds"))
```

---

## 2. Batch rename across all conditions

The typical workflow: patch and re-save all RDS files, then regenerate
comparison figures without any re-inference.

```{r rename-batch}
# Step 1: patch and re-save all RDS files
rds_dir   <- file.path(tempdir(), "CellChat", "CellChatObjects")
rds_files <- list.files(rds_dir, pattern = "CellChat\\.rds$", full.names = TRUE)

for (f in rds_files) {
  cc <- RenameCellTypeInCC(readRDS(f),
    old_name = "Immune_contam",
    new_name  = "NK_TCells")
  saveRDS(cc, f)
}

# Step 2: reload the fixed list and regenerate comparison figures
cclist <- setNames(
  lapply(rds_files, readRDS),
  sub("_CellChat\\.rds$", "", basename(rds_files))
)

CompareCellChat(cclist,
  output_dir = file.path(tempdir(), "CellChat", "Comparison"))

RankCellChatPathways(cclist,
  output_dir = file.path(tempdir(), "CellChat", "Ranked"))

# Step 3: regenerate per-group RunCellChat PDFs (plots only, no re-inference)
RunCellChat(bmcite,
  cell.by    = "celltype.l1",
  group.by   = "Condition",
  output_dir = file.path(tempdir(), "CellChat"),
  redo_plots = TRUE)
```

---

## Typical two-condition workflow

```{r full-workflow}
out_dir <- file.path(tempdir(), "CellChat")

# Step 1: run CellChat per condition
cc_results <- RunCellChat(bmcite,
  cell.by    = "celltype.l1",
  group.by   = "Condition",
  groups     = c("Control", "Treated"),
  colors     = GetColors(bmcite, "celltype.l1"),
  output_dir = out_dir,
  resume     = TRUE)

# Step 2: cross-condition comparison
CompareCellChat(
  cellchat_list = cc_results,
  output_dir    = file.path(out_dir, "Comparison"))

# Step 3: pathway ranking
RankCellChatPathways(
  cellchat_list = cc_results,
  output_dir    = file.path(out_dir, "Ranked"),
  top_n         = 20L)
```

---

## Parameter reference - `RunCellChat`

| Parameter | Type | Default | Description |
|-----------|------|---------|-------------|
| `seurat_object` | Seurat | required | Processed, annotated Seurat object |
| `cell.by` | character | required | Metadata column defining cell identity |
| `group.by` | character | required | Metadata column defining groups (one run per level) |
| `groups` | character vector | `NULL` | Subset or reorder levels of `group.by` |
| `species` | character | `"human"` | `"human"` or `"mouse"` |
| `colors` | named character | `NULL` | Override cell-type color palette |
| `output_dir` | character | required | Root directory for saving outputs |
| `robj_dir` | character or `NULL` | `NULL` | Directory for CellChat RDS files; defaults to `output_dir/CellChatObjects/` |
| `min.cells` | integer | `0` | Minimum cells a group must have to be included |
| `thresh.p` | numeric | `0.05` | P-value threshold for CellChat communication inference |
| `run.patterns` | logical | `TRUE` | Run communication pattern NMF analysis |
| `pattern_k_range` | integer vector | `3:10` | Range of k values for NMF pattern selection |
| `save.rds` | logical | `TRUE` | Save CellChat objects as RDS files |
| `resume` | logical | `FALSE` | Skip groups with existing RDS files |
| `redo_plots` | logical | `FALSE` | Reload saved RDS and regenerate PDFs only |
| `assay_type` | character or `NULL` | `NULL` | Seurat assay to extract counts from |
| `caffeinate` | logical | `FALSE` | Prevent system sleep during the run |

## Parameter reference - `CompareCellChat`

| Parameter | Type | Default | Description |
|-----------|------|---------|-------------|
| `cellchat_list` | named list | required | Named list of CellChat objects or paths to `.rds` files |
| `output_dir` | character | required | Directory for saving comparison plots |
| `groups.show` | character vector | `NULL` | Subset of conditions to include |
| `group.merged` | character | `NULL` | Merge these conditions before comparison |
| `colors` | named character | `NULL` | Override cell-type palette |
| `species` | character | `"human"` | `"human"` or `"mouse"` |
| `thresh` | numeric | `0.05` | P-value threshold; use `1` to include all pathways |
| `caffeinate` | logical | `FALSE` | Prevent system sleep |

## Parameter reference - `RankCellChatPathways`

| Parameter | Type | Default | Description |
|-----------|------|---------|-------------|
| `cellchat_list` | named list | required | Named list of CellChat objects or paths to `.rds` files |
| `output_dir` | character or `NULL` | `NULL` | Directory for PDFs and CSV; `NULL` = return objects only |
| `top_n` | integer | `20` | Number of top pathways in flow heatmap and tile plots |
| `top_links` | integer | `15` | Number of top sender-receiver pairs in links heatmap |
| `rank_by` | character | `"range"` | Ranking metric: `"range"` (max - min flow) or `"cv"` (coefficient of variation) |
| `colors` | named character | `NULL` | Condition colors for heatmap annotation strips |
| `cell_colors` | named character | `NULL` | Cell-type colors for sender/receiver tile annotation |

## Parameter reference - `RenameCellTypeInCC`

| Parameter | Type | Default | Description |
|-----------|------|---------|-------------|
| `cc` | CellChat | required | CellChat object to modify |
| `old_name` | character | required | Cell type label to replace |
| `new_name` | character | required | Replacement label |

---

## See also

- Vignette 01: `PrepObject` - register cell-type colors used in chord diagrams
- Vignette 06: `CellTypeAssignmentHelper` - assign cell types before CellChat
- Vignette 08: `RunGSEA` - pathway-level gene set enrichment
- Vignette 09: `RunGSEA_pseudobulk` - sample-level pseudobulk pathway analysis

---

## Session info

```{r session-info}
sessionInfo()
```
