---
title: "scSidekick - Exploring Metadata & Composition"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{scSidekick - Exploring Metadata & Composition}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 8,
  fig.height = 5
)
options(cli.width = 80)
```

## Overview

Before any fancy analysis, you usually just want to *understand your object*:
how many cells per sample, how cell-type composition shifts between conditions,
which donors carry which clinical features. The base-R answer is a pile of
nested `table()` calls and a few dozen lines of `dplyr` + `ggplot2`.

scSidekick turns that into one-liners:

- **`SummarizeMetadata()`** - collapse a giant `meta.data` into a tidy
  per-group summary table (counts, numeric means, categorical breakdowns).
- **`PlotMetaSummary()`** - composition as bar charts (counts **and**
  percentages) across one or more grouping levels, in a single call.
- **`PlotAlluvial()`, `PlotRose()`, `PlotChord()`, `PlotAtlasWheel()`** - a
  family of "at a glance" composition graphics for flows, proportions,
  relationships, and whole-atlas overviews.

```{r pkgs}
library(scSidekick)
library(Seurat)
library(SeuratData)
library(ggforce)   # PlotAlluvial uses ggforce's parallel-sets stats
```

```{r data, include=FALSE}
data("bmcite")
bmcite <- UpdateSeuratObject(bmcite)
bmcite <- PrepObject(bmcite,
  variables   = c("lane", "donor", "celltype.l1", "celltype.l2"),
  group.by    = "celltype.l1",
  split.by    = "donor",
  object_name = "bmcite")

# Most composition helpers accept either a Seurat object or a plain data frame.
# We use the metadata frame directly for the chart family below.
md <- bmcite@meta.data
```

We use `bmcite`, a bone-marrow CITE-seq dataset (~30 k cells) with rich
metadata: `lane`, `donor`, `celltype.l1`, and `celltype.l2`.

---

## 1. `SummarizeMetadata()` - one tidy table from messy metadata

Point `SummarizeMetadata()` at one or more **ID columns** and it returns one
row per ID combination with: an `n_cells` count, the mean of every numeric
column, and a compact breakdown of every categorical column - all the columns
you didn't name are summarised automatically.

```{r summarize}
donor_summary <- SummarizeMetadata(bmcite, id_columns = "donor")
donor_summary[, 1:5]   # first few summary columns
```

Numeric columns are aggregated with `numeric_func` (default `mean`; pass
`median`, `max`, a custom function, ...). Use several ID columns to summarise at
a finer grain:

```{r summarize-multi}
ds <- SummarizeMetadata(bmcite,
                        id_columns   = c("donor", "celltype.l1"),
                        numeric_func = median)
head(ds[, c("donor", "celltype.l1", "n_cells", "nFeature_RNA")])
```

This is the everyday "what's actually in my object?" answer - and it scales: on
a multi-million-cell atlas you can collapse to one row per donor in seconds.

---

## 2. `PlotMetaSummary()` - composition at a glance

The most common request - *how does cell-type composition differ between
samples?* - is one call. `variables` is what you're counting, `fill_variable`
colours the stack, and `count_unit = "cells"` counts cells (vs donors).

```{r meta-counts, fig.width=10, fig.height=4}
PlotMetaSummary(bmcite,
  variables     = c("celltype.l1", "celltype.l2"),
  fill_variable = "donor",
  count_unit    = "cells")
```

Add `percent = TRUE` for normalised composition - ideal when group sizes
differ:

```{r meta-percent, fig.width=10, fig.height=4}
PlotMetaSummary(bmcite,
  variables     = c("celltype.l1", "celltype.l2"),
  fill_variable = "donor",
  count_unit    = "cells",
  percent       = TRUE)
```

### Donor-level summaries with `id_column`

For clinical datasets you usually want **per-donor** counts, not per-cell.
Setting `id_column` collapses cells to donors first, then lays out the
variables - and `row_variable` facets into rows. This is exactly how you'd
summarise a large patient cohort (e.g. Braak / APOE / CERAD by sex):

```{r meta-donor, eval=FALSE}
# Illustrative - for a clinical atlas `md` with one row per cell:
PlotMetaSummary(md,
  id_column     = "Donor.ID",
  variables     = c("Braak", "APOE.Genotype", "CERAD.score"),
  fill_variable = "Cognitive.Status",
  row_variable  = "Sex")
```

The same function call handles a 30 k-cell demo and a 1.8 M-cell, 150-column
clinical atlas - it just collapses to the unit you ask for.

---

## 3. Composition chart family

Bar charts aren't always the clearest view. scSidekick ships four
complementary "at a glance" graphics.

### `PlotAlluvial()` - flows across categories

Alluvial plots show how cells flow between two or more categorical axes -
great for "which donors contribute to which cell types".

```{r alluvial, fig.width=8, fig.height=6}
PlotAlluvial(md,
  axes = c("donor", "celltype.l1"))
```

### `PlotRose()` - circular composition

A polar/“rose” chart of composition per group - a compact alternative to
stacked bars.

```{r rose, fig.width=8, fig.height=7}
PlotRose(md,
  stat.by = "celltype.l1",
  group.by = "donor")
```

### `PlotChord()` - relationships between two variables

Chord diagrams emphasise the *pairing* between two categoricals (here, which
cell types appear in which donors). `PlotChord()` uses **circlize** under the
hood.

```{r chord, fig.width=7, fig.height=7}
PlotChord(md,
  var1 = "donor",
  var2 = "celltype.l1")
```

### `PlotAtlasWheel()` - whole-dataset overview

For multi-study atlases, `PlotAtlasWheel()` draws a single "cover figure"
summarising cells per group across studies and patients - counts in the
centre, a wheel of wedges around it. (Shown as a template; it needs
study/patient columns that `bmcite` doesn't carry.)

```{r atlas, eval=FALSE}
PlotAtlasWheel(meta,
  group.by = "CellType",
  study_by   = "Study",
  patient_by = "Patient.ID",
  title      = "My Atlas",
  cell_label = "Cells")
```

---

## See also

- Vignette 01: `PrepObject()` - register the colours these charts reuse
- Vignette 03: `PlotComposition()` / `PlotTrendLabeled()` - composition trends
  on the UMAP
- Vignette 14: large-data workflows (BPCells / sketch / pseudobulk) for
  summarising massive objects

---

## Session info

```{r session-info}
sessionInfo()
```
