---
title: "scSidekick - Large Datasets: BPCells, Sketch & Pseudobulk"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{scSidekick - Large Datasets: BPCells, Sketch & Pseudobulk}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 8,
  fig.height = 5
)
options(cli.width = 80)
```

## Overview

Atlas-scale objects (hundreds of thousands to millions of cells) break the
usual "load everything into memory" assumption. scSidekick is built to scale in
two complementary ways:

1. **Streaming heavy functions** - feature plots, pseudobulk aggregation, and
   pathway scoring read [BPCells](https://bnprks.github.io/BPCells/) /
   on-disk and `Sketch` assays *lazily*. Only the genes (or cells) you ask for
   are pulled off disk, so the same one-liners run on a 2-million-cell object
   without materialising it.
2. **Aggregating to a tractable unit** - collapse millions of cells to
   per-sample pseudobulk profiles, then analyse those with fast, robust,
   sample-level statistics.

```{r pkgs}
library(scSidekick)
library(Seurat)
library(SeuratData)
```

```{r data, include=FALSE}
data("bmcite")
bmcite <- UpdateSeuratObject(bmcite)
bmcite <- NormalizeData(bmcite, verbose = FALSE)
bmcite <- FindVariableFeatures(bmcite, verbose = FALSE)
bmcite <- PrepObject(bmcite,
  variables   = c("donor", "celltype.l1"),
  group.by    = "celltype.l1",
  object_name = "bmcite")
```

---

## 1. BPCells & on-disk assays

With a BPCells-backed object the count matrix lives on disk, not in RAM.
scSidekick's plotting and aggregation functions detect the lazy matrix and
**stream** through it - they subset the requested rows *before* materialising,
so memory stays flat regardless of object size.

```{r bpcells, eval=FALSE}
library(BPCells)

# Write a 10x matrix to a BPCells on-disk directory (once):
mat <- open_matrix_10x_hdf5("filtered_feature_bc_matrix.h5")
write_matrix_dir(mat, dir = "bp_counts")

# Build a Seurat object whose counts are the on-disk matrix:
counts <- open_matrix_dir("bp_counts")
obj    <- CreateSeuratObject(counts) |>
  NormalizeData() |>
  FindVariableFeatures() |>
  ScaleData() |>
  RunPCA() |>
  RunUMAP(dims = 1:30)

obj <- PrepObject(obj,
  variables  = c("Sample", "Group", "seurat_clusters"),
  group.by   = "seurat_clusters",
  output_dir = "./Figures")

# These behave exactly as on an in-memory object - but only the requested
# genes are read off disk for each panel:
PlotFeaturePlots(obj, features = c("CD3E", "LYZ", "NKG7"), split.by = "Sample")
GroupHeatmap(obj, features = marker_genes)
```

## 2. The Sketch workflow

For interactive work on millions of cells, Seurat v5's `SketchData()` keeps a
representative subset "in memory" while the full data stays on disk. scSidekick
functions work on either assay - sketch for fast iteration, full for the final
figure.

```{r sketch, eval=FALSE}
obj <- SketchData(obj, ncells = 50000, method = "LeverageScore",
                  sketched.assay = "sketch")

# Iterate fast on the sketch assay ...
DefaultAssay(obj) <- "sketch"
PlotFeaturePlots(obj, features = "CD3E", split.by = "Sample")

# ... then project back to the full dataset for the publication figure.
# (For pseudobulk / pathway analysis on sketched objects, pass assay = "RNA"
#  explicitly so aggregation uses the full data, not the sketch.)
```

---

## 3. Pseudobulk: collapse cells to samples

Pseudobulk is the most robust way to analyse large multi-sample studies: sum
(or average) each gene within each sample, then treat samples as replicates.
`ComputePseudobulk()` streams the matrix once (BPCells-safe) and caches the
result in the object; `PlotPseudoBulk()` draws per-gene, per-group profiles
with one point per donor.

```{r pseudobulk-compute}
bmcite <- ComputePseudobulk(bmcite,
  group.by = c("donor", "celltype.l1"),
  genes    = c("LYZ", "CD3D", "NKG7", "CD79A", "S100A8"),
  verbose  = FALSE)
```

```{r pseudobulk-plot, fig.width=8, fig.height=4}
PlotPseudoBulk(bmcite,
  features = c("LYZ", "CD3D"),
  group.by = "celltype.l1",
  donor.by = "donor")
```

With more samples you can add cross-group statistics directly:

```{r pseudobulk-stats, eval=FALSE}
PlotPseudoBulk(obj,
  features    = "APOE",
  group.by = "Cognitive.Status",
  donor.by = "Donor.ID",
  split.by = "CellType",
  exclude     = list(Cognitive.Status = "Reference"),
  add_stats   = TRUE,
  ref_group   = "No dementia",
  min_cells   = 10)
```

---

## 4. Pathway analysis at scale

Two routes, both designed for big objects:

### Donor-level GSEA - `RunGSEA_pseudobulk()`

Pseudobulk -> per-contrast differential expression -> GSEA against Hallmark /
KEGG / Reactome / WikiPathways, plus optional ssGSEA. Sample-level statistics
mean the result isn't inflated by cell count.

```{r gsea-pb, eval=FALSE}
RunGSEA_pseudobulk(obj,
  group.by    = "CellType",
  sample.by   = "Donor.ID",
  contrast.by = "Cognitive.Status",
  contrasts   = list(Dementia_vs_Control = "Dementia - No dementia"),
  species     = "human",
  output_dir  = "./Figures/Pathways_pseudobulk")
```

### Per-cell ssGSEA - `RunSCssGSEA()`

When you want a pathway *score per cell* (e.g. to overlay on the UMAP),
`RunSCssGSEA()` computes single-cell ssGSEA and is BPCells/sketch-aware (use
`assay = "RNA"` / `"sketch"` as appropriate).

```{r ssgsea, eval=FALSE}
RunSCssGSEA(obj,
  group.by         = "CellType",
  gene_set_library = "H",          # MSigDB Hallmark
  output_dir       = "./Figures/ssGSEA")
```

---

## Tips for big objects

- Keep counts on disk with **BPCells**; scSidekick streams them.
- Use a **sketch** assay for interactive iteration, the full assay for final
  figures - and pass `assay = "RNA"` explicitly to pseudobulk/pathway
  functions so aggregation uses the full data.
- Prefer **pseudobulk** (`ComputePseudobulk` -> `PlotPseudoBulk`,
  `RunGSEA_pseudobulk`) for differential and pathway analysis across samples -
  it's both faster and statistically sounder than per-cell tests.

---

## See also

- Vignette 08: `RunGSEA()` - cell-level pathway analysis (keyword search, leading-edge visualization)
- Vignette 09: `RunGSEA_pseudobulk()` - full pseudobulk GSEA with limma-voom
- Vignette 13: `PlotMetaSummary()` - donor-level summaries for large cohorts
- Vignette 04: `PlotFeaturePlots()` - the streaming feature-map function

---

## Session info

```{r session-info}
sessionInfo()
```
