---
title: "scSidekick - Survival Analysis: TCGA & Custom Cohorts"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Survival Analysis: TCGA and Custom Cohorts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, message = FALSE, warning = FALSE)
```

## Overview

scSidekick provides a complete Kaplan-Meier survival analysis pipeline for bulk
RNA-seq expression data. Four functions cover the full workflow:

| Function | What it does |
|---|---|
| `LoadTCGA()` | Download and cache a TCGA or TARGET expression dataset |
| `EnrichTCGAMeta()` | Add GDC clinical + molecular subtype columns to metadata |
| `SurvMetaSummary()` | Explore available clinical columns before running survival analysis |
| `SurvPlot()` | Kaplan-Meier plots for genes or gene-set signatures, with optional faceting and multivariate Cox |

Two input modes:

- **TCGA mode** - pass `tcga_project = "TCGA-GBM"` and data are downloaded,
  processed, and cached automatically. Subsequent calls load instantly from cache.
- **Custom matrix mode** - pass your own `expr_mat` (genes x samples) and `meta`
  (samples x variables). Works for any bulk RNA-seq or microarray cohort.

### Required packages

```{r install}
# Core survival packages
install.packages(c("survival", "survminer"))

# For TCGA downloads (recommended backend - single consolidated download)
BiocManager::install("recount3")

# For richer clinical metadata and molecular subtypes (optional but recommended)
BiocManager::install("TCGAbiolinks")
```

---

# Part 1: Loading TCGA Data

## 1.1 Download a TCGA project

`LoadTCGA()` fetches the full expression matrix and clinical metadata for any
TCGA or TARGET project and caches the result locally. The default backend is
`recount3`, which downloads a single consolidated file per project - much faster
and more reliable than the per-patient GDC downloads.

```{r load-tcga}
library(scSidekick)

# Download TCGA-GBM (glioblastoma) - ~173 patients
# First call downloads; subsequent calls load from cache instantly
gbm <- LoadTCGA("TCGA-GBM")

# Access the expression matrix (genes x samples, log2 TPM+1)
dim(gbm$expr_mat)

# Access clinical metadata
head(gbm$meta[, c("OS", "status", "gender")])
```

The returned list contains:
- `$expr_mat` - numeric matrix, genes x samples, log2(TPM + 1)
- `$meta` - data frame of clinical metadata, one row per sample
- `$project`, `$data_type`, `$backend` - provenance fields

---

## 1.2 Available TCGA projects

Pass any `"TCGA-<CODE>"` string to `tcga_project`. Common cancer types:

| Code | Cancer type | n |
|------|-------------|---|
| `"TCGA-GBM"` | Glioblastoma | ~173 |
| `"TCGA-LGG"` | Lower grade glioma | ~516 |
| `"TCGA-BRCA"` | Breast invasive carcinoma | ~1,098 |
| `"TCGA-LUAD"` | Lung adenocarcinoma | ~585 |
| `"TCGA-LUSC"` | Lung squamous cell | ~504 |
| `"TCGA-SKCM"` | Skin cutaneous melanoma | ~473 |
| `"TCGA-PRAD"` | Prostate adenocarcinoma | ~551 |
| `"TCGA-KIRC"` | Kidney clear cell | ~611 |
| `"TCGA-OV"` | Ovarian carcinoma | ~608 |
| `"TCGA-LAML"` | Acute myeloid leukemia | ~200 |

Pediatric TARGET projects: `"TARGET-AML"`, `"TARGET-NBL"`, `"TARGET-OS"`,
`"TARGET-WT"`, `"TARGET-RT"`, `"TARGET-CCSK"`.

---

## 1.3 Forcing a fresh download

```{r load-force}
# Re-download and update the cache
gbm <- LoadTCGA("TCGA-GBM", force_reload = TRUE)
```

---

## 1.4 Microarray data

For older TCGA accessions or external microarray cohorts, set
`data_type = "microarray"`. This automatically switches to the TCGAbiolinks
backend (recount3 only carries RNA-seq).

```{r load-microarray}
gbm_array <- LoadTCGA("TCGA-GBM", data_type = "microarray")
```

---

# Part 2: Enriching Metadata

## 2.1 Add GDC clinical and molecular subtype columns

The recount3 download includes basic clinical columns (OS, status, gender).
`EnrichTCGAMeta()` adds the full GDC comprehensive clinical table and TCGA
marker-paper molecular subtype columns (`paper_*`) without re-downloading the
expression matrix.

```{r enrich-meta}
# Add IDH status, PAM50, CIMP, stage, age, treatment columns, etc.
gbm <- EnrichTCGAMeta(gbm)

# Now $meta contains paper_IDH.status, paper_Histology, etc.
SurvMetaSummary(gbm$meta)
```

Available `paper_*` columns depend on the cancer type. For GBM/LGG these
typically include `paper_IDH.status`, `paper_Histology`,
`paper_Pan.Glioma.RNA.Expression.Cluster`, and `paper_CIMP.status`.

---

# Part 3: Exploring Metadata

## 3.1 SurvMetaSummary

Always run `SurvMetaSummary()` before `SurvPlot()` to see which metadata
columns are available, which look like survival endpoints, and which are good
candidates for faceting.

```{r meta-summary}
SurvMetaSummary(gbm$meta)
```

The output is organized into sections:

- **Survival time columns** - candidates for `time_col`
- **Survival event columns** - candidates for `status_col`
- **Good facet columns** (2-10 unique values, < 50% NA) - candidates for
  `facet_row` / `facet_col`
- **Many-level columns** - consider binning before faceting
- **High-missingness columns** - use with caution

```{r meta-summary-all}
# Also show all numeric columns
SurvMetaSummary(gbm$meta, show_all = TRUE)
```

---

# Part 4: SurvPlot

## 4.1 Basic survival plot - single gene, TCGA mode

The simplest call: one gene, no faceting, median split.

```{r basic-gene}
SurvPlot(
  tcga_project = "TCGA-GBM",
  genes        = "YAP1",
  time_col     = "OS",
  status_col   = "status",
  output_file  = "~/Desktop/GBM_Survival/YAP1_survival.pdf"
)
```

Samples are split into **High** (>= median) and **Low** (< median) expression
groups. The log-rank p-value is shown on the plot. A sidecar CSV recording the
cutoff, group sizes, and p-value is written alongside the PDF.

---

## 4.2 Multiple genes

Pass a character vector to `genes`. Each gene produces a separate PDF in the
output directory.

```{r multi-gene}
SurvPlot(
  tcga_project = "TCGA-GBM",
  genes        = c("YAP1", "CD276", "CD47", "S100A4", "CCL5"),
  time_col     = "OS",
  status_col   = "status",
  output_file  = "~/Desktop/GBM_Survival/"   # directory -> one PDF per gene
)
```

To put all genes in a single multi-page PDF, use a path ending in `.pdf`:

```{r multi-gene-pdf}
SurvPlot(
  tcga_project = "TCGA-GBM",
  genes        = c("YAP1", "CD276", "CD47"),
  time_col     = "OS",
  status_col   = "status",
  output_file  = "~/Desktop/GBM_Survival/combined.pdf"  # single file
)
```

---

## 4.3 Faceting by clinical variables

`facet_row` and `facet_col` arrange panels into a grid. This is one of
`SurvPlot`'s most powerful features: see whether the survival association holds
within each molecular subtype or sex.

```{r facet}
# Sex (rows) x IDH status (columns)
SurvPlot(
  tcga_project = "TCGA-GBM",
  genes        = c("YAP1", "S100A4"),
  time_col     = "OS",
  status_col   = "status",
  facet_row    = "gender",
  facet_col    = "paper_IDH.status",
  output_file  = "~/Desktop/GBM_Survival/"
)
```

One page is produced per gene. The grid is sized automatically
(~4.5 in per panel). Facet cells with fewer than `2 * min_group_n` samples are
left blank to preserve alignment.

---

## 4.4 Splitting strategies

### Median split (default)

```{r split-median}
SurvPlot(
  tcga_project = "TCGA-GBM",
  genes        = "YAP1",
  split_method = "median",
  time_col     = "OS",
  status_col   = "status",
  output_file  = "~/Desktop/GBM_Survival/"
)
```

`split_per_group = TRUE` (default) computes the median independently within
each facet cell - so a subgroup with uniformly low expression is still split
50/50. This is the approach used in most published survival analyses. Set
`split_per_group = FALSE` for a single global threshold.

### Quartile split

Produces four curves (Q1-Q4). Useful for detecting non-linear dose-response
relationships.

```{r split-quartile}
SurvPlot(
  tcga_project     = "TCGA-GBM",
  genes            = c("YAP1", "CD276", "CD47"),
  split_method     = "quartile",
  quartile_palette = c("#2166AC", "#92C5DE", "#F4A582", "#D6604D"),
  time_col         = "OS",
  status_col       = "status",
  output_file      = "~/Desktop/GBM_Survival/quartiles.pdf"
)
```

### Fixed value split

Split at a user-defined threshold. Use `split_value = 0` for gene-set
enrichment scores where 0 means no enrichment.

```{r split-value}
SurvPlot(
  tcga_project = "TCGA-GBM",
  genes        = "YAP1",
  split_method = "value",
  split_value  = 4.0,    # log2 TPM+1 threshold
  time_col     = "OS",
  status_col   = "status",
  output_file  = "~/Desktop/GBM_Survival/"
)
```

### Optimal split

Finds the threshold that maximizes the log-rank statistic (survminer).
Best used for validation cohorts; can overfit on small datasets.

```{r split-optimal}
SurvPlot(
  tcga_project = "TCGA-GBM",
  genes        = "YAP1",
  split_method = "optimal",
  time_col     = "OS",
  status_col   = "status",
  output_file  = "~/Desktop/GBM_Survival/"
)
```

---

## 4.5 Gene-set signatures

Instead of individual genes, score each sample as the mean expression of a
named gene set. Pass a named list to `gene_lists`. Each element becomes one
survival panel.

```{r gene-sets}
SurvPlot(
  tcga_project = "TCGA-LGG",
  gene_lists   = list(
    YAP_Activated = c("PRC1", "HPCA", "SRSF7", "TIMM17A", "PIM1"),
    YAP_Repressed = c("ZIC1", "ZIC2", "CHD7", "KIF5C", "DPYSL2")
  ),
  split_method  = "median",
  time_col      = "OS",
  status_col    = "status",
  facet_col     = "paper_Pan.Glioma.RNA.Expression.Cluster",
  output_file   = "~/Desktop/LGG_Signatures/"
)
```

For enrichment-style interpretation, use `split_value = 0` (score >= 0 means
enriched):

```{r gene-sets-value}
SurvPlot(
  tcga_project = "TCGA-LGG",
  gene_lists   = list(
    YAP_Activated = c("PRC1", "HPCA", "SRSF7"),
    YAP_Repressed = c("ZIC1", "ZIC2", "CHD7")
  ),
  split_method = "value",
  split_value  = 0,
  time_col     = "OS",
  status_col   = "status",
  output_file  = "~/Desktop/LGG_Signatures/"
)
```

---

## 4.6 Custom cohort (non-TCGA)

Pass your own expression matrix and metadata. Rownames of `meta` must match
colnames of `expr_mat`.

```{r custom-cohort}
# Cavalli medulloblastoma dataset (example)
Cav.EXP  <- t(read.table("Cavalli_expression.txt", header = TRUE, row.names = 1))
Cav.meta <- read.table("Cavalli_pheno.txt", header = TRUE, row.names = 1)

res <- SurvPlot(
  expr_mat     = Cav.EXP,
  meta         = Cav.meta,
  genes        = c("YAP1", "CD276", "CD47", "CCL5"),
  time_col     = "survival",   # column name in your metadata
  status_col   = "status",
  facet_row    = "Gender",
  facet_col    = "Subgroup",
  split_method = "median",
  output_file  = "~/Desktop/Cavalli_Survival/"
)
```

---

## 4.7 Multivariate Cox regression

Supply `multivariate_covariates` to run a multivariate Cox proportional hazards
model alongside the KM analysis. A forest plot (hazard ratios + 95% CI) is
saved as a companion `_cox.pdf` file. The Cox model uses continuous expression
values, not the High/Low labels.

```{r cox}
SurvPlot(
  tcga_project            = "TCGA-GBM",
  genes                   = c("YAP1", "S100A4"),
  time_col                = "OS",
  status_col              = "status",
  facet_col               = "paper_IDH.status",
  multivariate_covariates = c("gender", "age_at_index", "paper_IDH.status"),
  output_file             = "~/Desktop/GBM_Survival/"
)
```

---

## 4.8 TARGET (pediatric) datasets

```{r target}
SurvPlot(
  tcga_project = "TARGET-NBL",
  genes        = c("MYCN", "ALK"),
  time_col     = "OS",
  status_col   = "status",
  split_method = "optimal",
  output_file  = "~/Desktop/NBL_Survival/"
)
```

---

## 4.9 Return value and sidecar CSV

`SurvPlot` returns a list with `$plots` and `$sidecar`. The sidecar records
every panel's cutoff, group sizes, and log-rank p-value - the complete audit
trail for the stratification used.

```{r return}
res <- SurvPlot(
  tcga_project = "TCGA-GBM",
  genes        = c("YAP1", "CD276"),
  time_col     = "OS",
  status_col   = "status"
)

# Inspect the sidecar
head(res$sidecar)

# Access a ggsurvplot object for further customization
res$plots[["YAP1"]][[1]]$plot
```

The sidecar CSV columns:

| Column | Description |
|--------|-------------|
| `feature` | Gene name or signature name |
| `facet_row`, `facet_col` | Facet level values (`"all"` when not faceted) |
| `split_method` | Method used (may differ if fallback occurred) |
| `cutoff` | Threshold applied |
| `n_total` | Samples in this panel |
| `n_High`, `n_Low` | Samples per group (or `n_Q1`-`n_Q4` for quartile) |
| `pval` | Log-rank p-value |

---

## 4.10 Typical full workflow

```{r full-workflow}
library(scSidekick)

# Step 1: download and cache
gbm <- LoadTCGA("TCGA-GBM")

# Step 2: enrich with molecular subtypes and full clinical data
gbm <- EnrichTCGAMeta(gbm)

# Step 3: explore available metadata columns
SurvMetaSummary(gbm$meta)

# Step 4: run survival analysis
SurvPlot(
  expr_mat                = gbm$expr_mat,
  meta                    = gbm$meta,
  genes                   = c("YAP1", "S100A4", "CD276", "CD47"),
  time_col                = "OS",
  status_col              = "status",
  facet_row               = "gender",
  facet_col               = "paper_IDH.status",
  split_method            = "median",
  multivariate_covariates = c("gender", "age_at_index", "paper_IDH.status"),
  output_file             = "~/Desktop/GBM_Survival/"
)
```

---

## Parameter reference - `LoadTCGA`

| Parameter | Type | Default | Description |
|-----------|------|---------|-------------|
| `project` | character | required | Project ID, e.g. `"TCGA-GBM"`, `"TARGET-NBL"` |
| `data_type` | character | `"rnaseq"` | `"rnaseq"` or `"microarray"` |
| `backend` | character | `"recount3"` | `"recount3"` (fast, single file) or `"TCGAbiolinks"` (per-patient) |
| `cache_dir` | character or `NULL` | `NULL` | Cache directory; defaults to OS user cache |
| `force_reload` | logical | `FALSE` | Delete cache and re-download |

## Parameter reference - `EnrichTCGAMeta`

| Parameter | Type | Default | Description |
|-----------|------|---------|-------------|
| `tcga_data` | list or data frame | required | List from `LoadTCGA()`, or a bare metadata data frame |
| `project` | character or `NULL` | `NULL` | Required only when passing a bare data frame |

## Parameter reference - `SurvMetaSummary`

| Parameter | Type | Default | Description |
|-----------|------|---------|-------------|
| `meta` | data frame | required | Clinical metadata |
| `max_levels` | integer | `10` | Max unique values for a column to be flagged as "good facet" |
| `na_threshold` | numeric | `0.5` | Fraction of NAs above which a column is flagged "high missingness" |
| `show_all` | logical | `FALSE` | Also show numeric columns |

## Parameter reference - `SurvPlot`

| Parameter | Type | Default | Description |
|-----------|------|---------|-------------|
| `expr_mat` | matrix or `NULL` | `NULL` | Genes x samples expression matrix (log scale) |
| `meta` | data frame or `NULL` | `NULL` | Sample-level clinical metadata |
| `tcga_project` | character or `NULL` | `NULL` | TCGA/TARGET project ID (alternative to `expr_mat` + `meta`) |
| `tcga_data_type` | character | `"rnaseq"` | `"rnaseq"` or `"microarray"` |
| `tcga_backend` | character | `"recount3"` | Download engine |
| `tcga_cache_dir` | character or `NULL` | `NULL` | Cache directory |
| `genes` | character vector or `NULL` | `NULL` | Gene symbols for single-gene mode |
| `gene_lists` | named list or `NULL` | `NULL` | Named gene sets for signature-score mode |
| `time_col` | character | `"OS"` | Survival time column in `meta` |
| `status_col` | character | `"status"` | Event indicator column (1 = event, 0 = censored) |
| `split_method` | character | `"median"` | `"median"`, `"quartile"`, `"value"`, or `"optimal"` |
| `split_value` | numeric or `NULL` | `NULL` | Fixed threshold for `split_method = "value"` |
| `split_per_group` | logical | `TRUE` | Compute threshold independently within each facet group |
| `facet_row` | character or `NULL` | `NULL` | Column defining panel rows |
| `facet_col` | character or `NULL` | `NULL` | Column defining panel columns |
| `output_file` | character or `NULL` | `NULL` | Path ending in `.pdf` (single file) or directory (one PDF per gene) |
| `pdf_width` | numeric or `NULL` | `NULL` | PDF width in inches (auto-sized when `NULL`) |
| `pdf_height` | numeric or `NULL` | `NULL` | PDF height in inches (auto-sized when `NULL`) |
| `palette` | character(2) | `c("Red","Blue")` | Colors for High and Low groups |
| `quartile_palette` | character(4) | blue-to-red | Colors for Q1-Q4 |
| `show_pval` | logical | `TRUE` | Show log-rank p-value on each panel |
| `show_n` | logical | `TRUE` | Append group n to legend labels |
| `write_sidecar` | logical | `TRUE` | Write companion sidecar CSV |
| `min_group_n` | integer | `5` | Minimum samples per group; panels below threshold are skipped |
| `multivariate_covariates` | character vector or `NULL` | `NULL` | Covariates for multivariate Cox; forest plot saved as `_cox.pdf` |

---

## See also

- Vignette 08: `RunGSEA` - pathway enrichment analysis from single-cell data
- Vignette 09: `RunGSEA_pseudobulk` - sample-level pseudobulk GSEA
- Vignette 10: `PlotPathwayButterfly` - uses the same mean-expression scorer for single-cell pathway state mapping

---

## Session info

```{r session-info}
sessionInfo()
```
