---
title: "scSidekick - Co-expression Analysis"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Co-expression Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 10,
  fig.height = 6
)
options(cli.width = 80)
```

## Overview

Three functions cover the co-expression workflow end-to-end:

- **`CorrelateGene()`** - correlate one gene of interest against every other
  gene, within each cell type, at single-cell, pseudobulk, or group-average
  resolution.
- **`PlotCorrelation()`** - visualize the results as per-cell-type
  volcano-style plots (r vs. -log10 adjusted p-value).
- **`SplitByGene()`** - label each cell as high / low (or expressed / zero)
  based on a gene's expression, per cell type. Feeds directly into
  `RunGSEA_pseudobulk()` or `FindMarkers()`.

```{r pkgs, eval=T, message=FALSE, warning=FALSE, include=T}
library(scSidekick)
library(Seurat)
library(SeuratData)
```

```{r data, message=FALSE, warning=FALSE, include=T}
#for this we will load an object with multiple donors/samples
#InstallData("panc8")
data("panc8")
panc8 <- UpdateSeuratObject(panc8)
panc8 <- NormalizeData(panc8, verbose = FALSE)
panc8$donor=panc8$orig.ident
panc8 <- PrepObject(panc8,
  variables   = c("donor", "celltype"),
  group.by    = "celltype",
  object_name = "panc8")
```

---

## 1. `CorrelateGene()` - single-cell level

The default `level = "sc"` computes Spearman correlation across individual
cells within each cell type. Only genes expressed in at least `min_pct`% of
cells in a given cell type are tested (`expressed_only = TRUE` by default).

```{r correlate-sc, eval=T}
cor_sc <- CorrelateGene(panc8,
  gene     = "STAT1",
  group.by = "celltype",
  level    = "sc")

head(cor_sc)
```

Each row is one gene in one cell type. Key columns:

| Column | Description |
|--------|-------------|
| `gene` | Correlated gene |
| `cell_type` | Cell-type group |
| `r` | Spearman (or Pearson) correlation with the gene of interest |
| `p` | Raw p-value |
| `padj` | BH-adjusted p-value |
| `n` | Number of cells tested |
| `goi` | Gene of interest |
| `level` | Resolution used (`"sc"`, `"pseudobulk"`, or `"group"`) |

---

## 2. `PlotCorrelation()` - visualize results

`PlotCorrelation()` draws one panel per cell type showing the top `n_pos`
positively and top `n_neg` negatively correlated genes. Significant genes
(`padj < padj_cutoff`) are colored; the top `label_n` in each direction are
labeled. Panels are assembled with `patchwork::wrap_plots()`.

```{r plot-cor, eval=T, fig.width=12, fig.height=8}
PlotCorrelation(cor_sc,
  n_pos       = 20,
  n_neg       = 20,
  label_n     = 5,
  padj_cutoff = 0.05,
  r_cutoff    = 0.3,     # draws ±0.3 dotted reference lines
  ncol        = 4)
```

Restrict to specific cell types with `cell.type`:

```{r plot-cor-subset, eval=T, fig.width=8, fig.height=4}
PlotCorrelation(cor_sc,
  cell.type = c("acinar", "ductal"),
  n_pos     = 15,
  n_neg     = 15,
  label_n   = 6)
```

Save to PDF (a `.legend` sidecar is written automatically). `width`/`height`
override the auto-calculated PDF size, and `timestamp = TRUE` appends a
`_YYYYMMDD-HHMMSS` stamp to the saved file name so repeated runs are
versioned instead of overwritten:

```{r plot-cor-save, eval=FALSE}
PlotCorrelation(cor_sc,
  n_pos      = 20,
  n_neg      = 20,
  output_dir = "./Figures",
  file_name  = "STAT1_coexpression",
  timestamp  = TRUE)
```

---

## 3. `CorrelateGene()` - pseudobulk level

For multi-donor datasets, pseudobulk correlation (`level = "pseudobulk"`)
aggregates counts to the donor x cell-type level first. This removes
within-donor cell-count inflation and is more appropriate for multi-sample
comparisons. Requires `sample.by`.

```{r correlate-pb, eval=T}
cor_pb <- CorrelateGene(panc8,
  gene       = "STAT1",
  group.by   = "celltype",
  level      = "pseudobulk",
  sample.by  = "donor",
  min_donors = 3)
```

### Group-average level

`level = "group"` averages expression within each combination of cell type and
`condition.by`, then correlates across those group-level averages. Useful when
donors are few but groups are many (e.g. comparing many cell-type x timepoint
combinations).

```{r correlate-group, eval=T}

cor_grp <- CorrelateGene(panc8,
  gene         = "STAT1",
  group.by     = "celltype",
  level        = "sc",
  condition.by = "donor",
  min_groups   = 3)

head(cor_grp)
```

---

## 4. Stratified analysis with `compare.by`

Pass `compare.by` to run the correlation separately within each level of a
grouping variable (e.g. sex, condition, batch). The result gains a
`compare_group` column, and `PlotCorrelation()` automatically lays out one
column per comparison group.

```{r compare-by, eval=T, fig.height=8, fig.width=8}
cor_by_dataset <- CorrelateGene(panc8,
  gene       = "STAT1",
  group.by   = "celltype",
  level      = "sc",
  compare.by = "dataset")

# ncol defaults to the number of comparison groups
PlotCorrelation(cor_by_dataset,
  cell.type = "acinar",
  n_pos     = 15,
  n_neg     = 15,ncol = 3)
```

---

## 5. `SplitByGene()` - label cells for downstream analysis

`SplitByGene()` adds a new metadata column splitting cells into expression
groups within each cell type. The column plugs straight into
`RunGSEA_pseudobulk()` (as `contrast.by`) or `Seurat::FindMarkers()`.

### Default: zero vs. expressed

```{r split-zero, eval=T}
panc8 <- SplitByGene(panc8,
  gene     = "STAT1",
  group.by = "celltype",
  split    = "zero_vs_expressed")

# New column "STAT1_split" is now in meta.data:
table(panc8$STAT1_split, panc8$celltype, useNA = "ifany")
```

### Quantile split

```{r split-quantile, eval=FALSE}
panc8 <- SplitByGene(panc8,
  gene           = "STAT1",
  group.by       = "celltype",
  split          = "quantile",
  quantile_range = c(0.25, 0.75))  # bottom 25% = "low", top 25% = "high"
```

Middle cells (between the two quantiles) receive `NA` and are excluded from
downstream tests, keeping only the contrast extremes.

### Feeding into `RunGSEA_pseudobulk()`

```{r split-to-gsea, eval=FALSE}
RunGSEA_pseudobulk(panc8,
  group.by    = "celltype",
  sample.by   = "donor",
  contrast.by = "STAT1_split",
  pathway_sets = list( Hallmark = list(category = "H")),
  species     = "human",
  output_dir  = "./Figures/STAT1_pseudobulk_GSEA")
```

### Feeding into `RunWilcoxAUC()`

```{r split-to-markers, eval=T}

markers <- RunWilcoxAUC(panc8,
  group.by  = "STAT1_split",split.by="celltype",
  min.pct   = 0.1,
  logfc.threshold = 0.25)
head(markers)
```


---

## See also

- Vignette 05: `GroupHeatmap()`, `SplitDotPlot()` - expression summaries per cell type
- Vignette 08: `RunGSEA()` - pathway analysis on a gene list
- Vignette 09: `RunGSEA_pseudobulk()` - full pseudobulk GSEA pipeline (`contrast.by` accepts the column from `SplitByGene`)

---

## Session info

```{r session-info}
sessionInfo()
```
