---
title: "scSidekick - Copy Number Variation Inference"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Copy Number Variation Inference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  eval    = FALSE,
  message = FALSE,
  warning = FALSE
)
options(cli.width = 80)
```

## Overview

Distinguishing malignant cells from normal stromal and immune infiltrates is a
common challenge in tumor single-cell studies. scSidekick provides two
complementary wrappers:

- **`RunInferCNV()`** - full inferCNV workflow with automatic gene-order
  generation, downsampling, HMM-optional inference, and a Tirosh-style
  reference-anchored CNV score threshold.
- **`RunCopyKAT()`** - fast Bayesian CNV calling via CopyKAT; can run in
  unsupervised mode (no normal reference required).

Both functions add new metadata columns to the Seurat object and save standard
figures to `output_dir`.

### When to use which

| | `RunInferCNV()` | `RunCopyKAT()` |
|---|---|---|
| **Accuracy** | Higher (gene-order smoothing + denoising) | Moderate (fast Bayesian model) |
| **Speed** | Slow (minutes to hours) | Fast (minutes) |
| **Reference required** | Yes | Optional (unsupervised mode) |
| **Best use** | Publication-grade tumor/normal separation | Quick first pass; no normal cells available |

Run both and compare `CNV_Prediction` vs. `copykat_prediction` for maximum
confidence.

---

## Installation

```{r install}
# inferCNV (Bioconductor)
if (!requireNamespace("BiocManager")) install.packages("BiocManager")
BiocManager::install("infercnv")

# CopyKAT (GitHub)
devtools::install_github("navinlabcode/copykat")
```

---

## 1. `RunInferCNV()` - reference-guided CNV inference

### Basic call

```{r infercnv-basic}
library(scSidekick)
library(Seurat)

# Typical tumor dataset: immune and stromal cells serve as the normal reference
obj <- RunInferCNV(obj,
  cell_type_col   = "Assignment",      # metadata column with cell-type labels
  ref_group_names = c("T cell", "B cell", "Macrophage", "Endothelial"),
  output_dir      = "./Figures/inferCNV",
  species         = "human")
```

`RunInferCNV()` automatically runs `AnnotateFeatures()` to generate the gene
order table if the object has not been annotated yet.

### Output metadata columns

| Column | Description |
|--------|-------------|
| `CNV_score` | Per-cell CNV score: `mean((expr - 1)^2)` across all genes (Tirosh 2016) |
| `CNV_Prediction` | `"Malignant"` or `"Normal"` based on `cnv_score_threshold` |
| `infercnv_subcluster` | Cell subcluster assigned by inferCNV's internal clustering |
| `proportion_cnv_chrN` | Fraction of chromosome N genes called as CNV-affected |

### Output figures

Three PDFs are saved to `output_dir`:

- `CNV_proportion_heatmap.pdf` - ComplexHeatmap of per-chromosome CNV
  proportions, split by cell type.
- `CNV_dimplot.pdf` - UMAP colored by `CNV_Prediction` (left) and cell type
  (right).
- `CNV_score_featureplot.pdf` - Continuous `CNV_score` overlaid on the UMAP.

---

## 2. Tuning the malignancy threshold

The default `cnv_score_threshold = "auto"` anchors the threshold to the
**reference cells**:

```
threshold = mean(ref_scores) + cnv_k * sd(ref_scores)
```

`cnv_k = 2` (default) is a good starting point. Increase it to call fewer
cells malignant (more conservative); decrease it to be more sensitive.

```{r threshold-auto}
# More conservative: require 3 SD above reference mean
obj <- RunInferCNV(obj,
  cell_type_col   = "Assignment",
  ref_group_names = c("T cell", "B cell", "Macrophage"),
  output_dir      = "./Figures/inferCNV",
  species         = "human",
  cnv_k           = 3)
```

Override with a fixed numeric value when you have a biological reason:

```{r threshold-manual}
obj <- RunInferCNV(obj,
  cell_type_col       = "Assignment",
  ref_group_names     = c("T cell", "B cell"),
  output_dir          = "./Figures/inferCNV",
  species             = "human",
  cnv_score_threshold = 0.02)
```

Inspect the score distribution to choose a threshold:

```{r threshold-inspect}
hist(obj$CNV_score,
  breaks = 50,
  main   = "CNV score distribution",
  xlab   = "CNV score",
  col    = "steelblue")
abline(v = 0.02, col = "red", lty = 2)

# Tabulate predictions
table(obj$CNV_Prediction, obj$Assignment)
```

---

## 3. `resume = TRUE` - skip re-running

inferCNV can take hours on large datasets. Use `resume = TRUE` after the first
run to skip the inference step and jump straight to post-processing (useful
when tuning `cnv_k` or `cnv_score_threshold`):

```{r resume}
# First run - full inference (~2 hours)
obj <- RunInferCNV(obj,
  cell_type_col   = "Assignment",
  ref_group_names = c("T cell", "B cell"),
  output_dir      = "./Figures/inferCNV",
  species         = "human")

# Tune threshold without re-running - post-processing only (~seconds)
obj <- RunInferCNV(obj,
  cell_type_col       = "Assignment",
  ref_group_names     = c("T cell", "B cell"),
  output_dir          = "./Figures/inferCNV",
  species             = "human",
  cnv_score_threshold = 0.015,
  resume              = TRUE)
```

---

## 4. HMM mode and advanced parameters

```{r infercnv-hmm}
RunInferCNV(obj,
  cell_type_col   = "Assignment",
  ref_group_names = c("T cell", "B cell"),
  output_dir      = "./Figures/inferCNV_HMM",
  species         = "human",
  HMM             = TRUE,     # adds subclonal CNV state calls; much slower
  denoise         = TRUE,
  downsample      = 300L,     # cells per group (NULL = use all)
  cutoff          = 0.1,      # 0.1 for 10x; use 1 for Smart-seq2
  num_threads     = 8L,
  caffeinate      = TRUE)     # keep laptop awake during long run
```

### Limiting chromosomes

By default all autosomes + X/Y are analyzed. Restrict to a subset when you
have a prior reason (e.g., known focal amplification on chr8q in MYC-driven
tumors):

```{r infercnv-chromosomes}
RunInferCNV(obj,
  cell_type_col   = "Assignment",
  ref_group_names = c("T cell", "B cell"),
  output_dir      = "./Figures/inferCNV",
  species         = "human",
  chromosomes     = as.character(c(1:8, 17)))
```

---

## 5. `RunCopyKAT()` - fast Bayesian CNV calling

### Reference-guided mode

```{r copykat-ref}
obj <- RunCopyKAT(obj,
  cell_type_col = "Assignment",
  ref_cells     = c("T cell", "B cell", "Macrophage"),
  output_dir    = "./Figures/CopyKAT",
  n_cores       = 8L)
```

### Unsupervised mode (no normal reference)

When the sample contains no confidently normal cells, omit `ref_cells`:

```{r copykat-unsupervised}
obj <- RunCopyKAT(obj,
  cell_type_col = "Assignment",
  output_dir    = "./Figures/CopyKAT",
  n_cores       = 8L)
```

CopyKAT will infer the diploid baseline automatically.

### Output metadata columns

| Column | Description |
|--------|-------------|
| `copykat_prediction` | `"aneuploid"`, `"diploid"`, or `"not.defined"` |
| `copykat_score` | Sum of absolute CNA values across all genomic windows (continuous) |

`"not.defined"` cells are those CopyKAT could not confidently classify -
typically very sparse cells or cells at the boundary of the two populations.

---

## 6. Combining both tools

Use both and require agreement for high-confidence malignancy calls:

```{r combine}
# Run both (output_dir can be the same or different)
obj <- RunInferCNV(obj,
  cell_type_col   = "Assignment",
  ref_group_names = c("T cell", "B cell"),
  output_dir      = "./Figures/inferCNV",
  species         = "human")

obj <- RunCopyKAT(obj,
  cell_type_col = "Assignment",
  ref_cells     = c("T cell", "B cell"),
  output_dir    = "./Figures/CopyKAT")

# Cells called malignant/aneuploid by BOTH tools
obj$CNV_consensus <- with(obj@meta.data,
  ifelse(CNV_Prediction == "Malignant" & copykat_prediction == "aneuploid",
         "Malignant", "Normal/Uncertain"))

table(obj$CNV_consensus, obj$Assignment)
```

Visualize agreement:

```{r combine-plot}
library(ggplot2)

ggplot(obj@meta.data, aes(x = CNV_score, y = copykat_score,
                            color = CNV_consensus)) +
  geom_point(size = 0.5, alpha = 0.6) +
  scale_color_manual(values = c(Malignant = "#ab3000",
                                 "Normal/Uncertain" = "#2f4b7c")) +
  labs(x = "inferCNV score", y = "CopyKAT score",
       title = "CNV score concordance") +
  theme_classic()
```

---

## See also

- `AnnotateFeatures()` / `GetGeneOrder()` - gene-order annotation used
  internally by `RunInferCNV()` (vignette 01)
- `PlotDimPlots()` - visualize `CNV_Prediction` on the UMAP after inference
  (vignette 03)
- `RunWilcoxAUC()` - find markers distinguishing malignant from normal clusters
  after labeling

---

## Session info

```{r session-info, eval=TRUE}
sessionInfo()
```
