---
title: "scSidekick - Gene Expression Programs with cNMF"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Gene Expression Programs with cNMF}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  eval    = FALSE,
  message = FALSE,
  warning = FALSE
)
options(cli.width = 80)
```

## Overview

Consensus non-negative matrix factorization (cNMF; Kotliar et al. 2019, eLife)
discovers **gene expression programs (GEPs)**: recurrent modules of co-expressed
genes that can correspond to cell-type identity or to shared activities such as
cell cycle, stress, or interferon response. Unlike clustering, a cell can use
several programs at once, which makes cNMF a natural complement to the
annotation, pathway, and co-expression tools elsewhere in scSidekick.

scSidekick wraps the Python [cNMF package](https://github.com/dylkot/cNMF)
through `reticulate` with three functions:

- **`RunCNMF()`** - prepares the data, runs the heavy factorization (optionally
  across parallel workers and with Harmony batch correction), and writes the
  K-selection diagnostic.
- **`GetCNMFPrograms()`** - extracts the consensus result for a chosen K and
  attaches per-cell **usages** as a dimensional reduction (and metadata) plus
  gene **spectra** to the object.
- **`GetCNMFTopGenes()`** - pulls the top genes per program for heatmaps and
  downstream enrichment.

cNMF is **interactive by design**: you run the factorization, inspect the
K-selection plot, choose the number of programs, then extract. It cannot be a
single one-shot call without compromising the science.

> **cNMF is a Python tool.** It is not an R dependency. Section 1 walks through
> a one-time environment setup that captures every compatibility fix needed to
> drive it from R; this is the part most likely to trip you up, so do it
> carefully once.

---

## 1. One-time environment setup

cNMF and its dependencies are Python packages with compiled extensions. Driving
them from R through `reticulate` (which runs Python *embedded* in the R process)
exposes several version-compatibility issues. The recipe below is a known-good
combination. Run it once.

### 1.1 Create the conda environment

```bash
# Python 3.9 from conda-forge
conda create -n cnmf_env python=3.9 -c conda-forge -y
conda activate cnmf_env
```

### 1.2 Install cNMF and pin the compatible versions

Each line addresses a specific failure mode (noted in the comment). Run them in
this order.

```bash
# (a) Provide libc++ INSIDE the env. numba/llvmlite (pulled in by scanpy) load a
#     compiled dylib that needs libc++.1.dylib; under R/reticulate the system
#     copy is not on the loader path, so install it into the env.
conda install -c conda-forge libcxx -y

# (b) Install cNMF itself.
pip install cnmf

# (c) Pin NumPy < 2. Two things require this: (i) reticulate's R<->NumPy bridge
#     (older reticulate cannot convert arrays from NumPy 2.0), and (ii) the
#     scikit-misc binary used by scanpy's seurat_v3 highly-variable-gene step.
pip install "numpy<2"

# (d) ONLY needed for batch correction (Section 7): harmonypy with the API that
#     cNMF's Preprocess expects. Newer harmonypy (0.0.10+) removed the internal
#     attribute cNMF reads, so pin 0.0.9.
pip install "harmonypy==0.0.9"
```

On the R side, make sure `reticulate` is installed (and reasonably current):

```{r reticulate-install}
install.packages("reticulate")
```

> **Alternative to pinning NumPy:** a current `reticulate` (>= ~1.39) can handle
> NumPy 2.0, but scanpy's `seurat_v3` HVG step (used by batch correction) still
> needs its compiled `scikit-misc` to match the installed NumPy. Pinning
> `numpy<2` keeps the whole stack consistent and is the simplest reliable path.

### 1.3 Point R at the environment

`reticulate` can only initialize **one** Python interpreter per R session, and
the binding must happen **before** any other Python-touching code runs. The most
robust approach is to set `RETICULATE_PYTHON` at the very top of a fresh
session:

```{r bind-python}
# FIRST line of a fresh R session, before library() calls:
Sys.setenv(RETICULATE_PYTHON = "/path/to/anaconda3/envs/cnmf_env/bin/python")

library(scSidekick)
library(Seurat)
```

(Find the path with `conda run -n cnmf_env python -c "import sys; print(sys.executable)"`,
or it is typically `~/anaconda3/envs/cnmf_env/bin/python` /
`~/miniconda3/envs/cnmf_env/bin/python`.)

### 1.4 Verify before a long run

```{r verify-env}
reticulate::py_config()                  # should show .../cnmf_env/bin/python, numpy 1.x
reticulate::py_module_available("cnmf")  # must be TRUE
```

If `py_config()` shows the wrong interpreter, **restart R** and set
`RETICULATE_PYTHON` before anything else. See the Troubleshooting table at the
end for the specific errors you may hit and their fixes.

---

## 2. Running cNMF

`RunCNMF()` writes a raw-counts AnnData, then runs cNMF `prepare` ->
`factorize` -> `combine` -> `k_selection_plot`. Point it at your environment
with `conda_env` (or rely on `RETICULATE_PYTHON` from Section 1.3).

```{r run-basic}
Immune <- RunCNMF(Immune,
  output_dir = "./Figures/cNMF",
  k_range    = 5:15,     # candidate numbers of programs
  n_iter     = 100,      # NMF replicates per K (robustness)
  conda_env  = "cnmf_env")
```

**Runtime scales with `n_iter * length(k_range)`, not object size** (the
defaults are 100 x 11 = 1100 NMF fits). For a quick end-to-end test, shrink
both:

```{r run-quick}
Immune <- RunCNMF(Immune, output_dir = "./Figures/cNMF",
                  n_iter = 5, k_range = 5:8, conda_env = "cnmf_env")
```

### Parallel factorization

The factorize step parallelizes across worker **processes**. Set `n_workers` to
the number of cores you can spare:

```{r run-parallel}
Immune <- RunCNMF(Immune, output_dir = "./Figures/cNMF",
                  k_range = 5:15, n_iter = 100,
                  n_workers = 6,           # 6 parallel cNMF factorize workers
                  conda_env = "cnmf_env")
```

Each worker logs to `<output_dir>/<name>.factorize.worker_<i>.log` - `tail -f`
one of them to watch progress.

### Key parameters

| Parameter | Default | Description |
|---|---|---|
| `k_range` | `5:15` | Candidate numbers of programs to factorize across |
| `n_iter` | `100L` | NMF replicates per K |
| `num_highvar_genes` | `2000L` | Overdispersed genes cNMF selects |
| `n_workers` | `1L` | Parallel factorize worker processes |
| `seed` | `14L` | Random seed for reproducibility |
| `name` | `"cnmf"` | Run name; output goes under `output_dir/name/` |
| `resume` | `FALSE` | Reuse a completed factorization on re-run |
| `batch.by` | `NULL` | Harmony batch correction (Section 7) |
| `caffeinate` | `FALSE` | Keep the machine awake during long runs |

> Reusing the same `name` in the same `output_dir` makes cNMF treat existing
> factorization files as done. For a clean run with different parameters, pass a
> fresh `name` (e.g. `name = "Immune_k5to15"`).

---

## 3. Choosing the number of programs (K)

`RunCNMF()` writes a **K-selection plot** (stability vs. reconstruction error)
and, when `show_plot = TRUE` (default), renders it into the RStudio Plots pane.
Pick the K at the stability peak / error elbow.

```{r kselect}
# The plot is at <output_dir>/cnmf/cnmf.k_selection.png and is shown
# automatically. To view it again later:
magick::image_read("./Figures/cNMF/cnmf/cnmf.k_selection.png")
```

There is no honest way to automate this choice; inspect the diagnostic and
decide.

---

## 4. Extracting programs for a chosen K

`GetCNMFPrograms()` runs the consensus step for your chosen `k` and a
`density_threshold` (the outlier filter; lower is stricter, `2.0` keeps
everything), then attaches the results to the object.

```{r get-programs}
Immune <- GetCNMFPrograms(Immune,
  k                 = 10,      # chosen from the K-selection plot
  density_threshold = 0.1,
  conda_env         = "cnmf_env")
```

This adds:

- A **dimensional reduction** `Immune[["cnmf"]]` with embeddings `cNMF_1 ...
  cNMF_k` (per-cell program usages), so you can overlay programs on the UMAP.
- **Metadata columns** `cNMF_1 ... cNMF_k` (when `add_metadata = TRUE`).
- `Immune@misc$cnmf$results` holding `usage`, `spectra_scores` (gene loadings),
  `spectra_tpm`, and `top_genes`.

The consensus **clustergram** is written next to the other outputs and shown in
the Plots pane.

Both `RunCNMF()` and `GetCNMFPrograms()` also write an `analysis_params.json`
into `output_dir` (the same provenance sidecar produced by `RunGSEA()`,
`RunCellChat()`, etc.). It records the cNMF parameters and a draft methods
paragraph, and is picked up automatically by `create_analysis_pptx()`:

```{r methods-json}
jsonlite::read_json(file.path("./Figures/cNMF", "analysis_params.json"))$methods_text
```

---

## 5. Visualizing programs

Because usages live as numeric metadata, the standard scSidekick plotting
functions accept them as "features" directly.

### Usages on the UMAP

```{r plot-usages-umap}
PlotFeaturePlots(Immune,
  features   = paste0("cNMF_", 1:10),
  output_dir = "./Figures/cNMF/usage_maps")
```

### Usage heatmap across conditions

```{r plot-usages-heatmap}
GroupHeatmap(Immune,
  features         = paste0("cNMF_", 1:10),
  group.by         = c("Group", "Sex"),
  column_split_by  = "genotype")
```

### Top genes per program

`GetCNMFTopGenes()` returns a tidy `program / gene / rank` table that feeds
straight into a blocked `GroupHeatmap` (or `SplitDotPlot`, or `RunGSEA`):

```{r top-genes}
tg <- GetCNMFTopGenes(Immune, n = 15)        # top 15 genes per program
head(tg)

# Heatmap of program marker genes, blocked by program:
GroupHeatmap(Immune,
  features      = tg$gene,
  group.by      = "CellType",
  feature_split = stats::setNames(tg$program, tg$gene))
```

Restrict to specific programs with `programs`:

```{r top-genes-subset}
tg <- GetCNMFTopGenes(Immune, n = 20, programs = c(1, 3, 5))
```

(`n` is capped by `top_n_genes` set in `GetCNMFPrograms()`, default 50.)

---

## 6. A complete run

```{r full-workflow}
# 1. Factorize (parallel), then inspect the K-selection plot
Immune <- RunCNMF(Immune, output_dir = "./Figures/cNMF",
                  k_range = 5:15, n_iter = 100, n_workers = 6,
                  conda_env = "cnmf_env", caffeinate = TRUE)

# 2. Choose K from the plot, extract the consensus programs
Immune <- GetCNMFPrograms(Immune, k = 10, density_threshold = 0.1,
                          conda_env = "cnmf_env")

# 3. Visualize
PlotFeaturePlots(Immune, features = paste0("cNMF_", 1:10),
                 output_dir = "./Figures/cNMF/usage_maps")
tg <- GetCNMFTopGenes(Immune, n = 15)
GroupHeatmap(Immune, features = tg$gene, group.by = "CellType",
             feature_split = stats::setNames(tg$program, tg$gene))
```

---

## 7. Batch correction (optional)

By default cNMF can learn programs that reflect batch rather than biology. cNMF's
recommended remedy (Kotliar et al., Figure 5) is to **Harmony-correct the counts
before factorization**, learn usages on the corrected data, and re-fit the
spectra on the uncorrected TP10K data. Supply one or more metadata columns to
`batch.by` and `RunCNMF()` runs this workflow via cNMF's `Preprocess`:

```{r batch}
Immune <- RunCNMF(Immune,
  output_dir = "./Figures/cNMF",
  k_range    = 5:15,
  n_iter     = 100,
  batch.by   = "Sample",          # or c("Sex", "Sample")
  conda_env  = "cnmf_env")
```

**Batch correction needs the extra dependencies from Section 1.2:** a
`numpy<2`-consistent `scikit-misc` (for the seurat_v3 HVG step) and
`harmonypy==0.0.9` (for the correction itself). If you skipped those, install
them and restart R before using `batch.by`.

When `batch.by` is set, `genes_use` is ignored (Preprocess selects
high-variance genes on the corrected data), and cNMF writes
`<name>.batchcorrect.Corrected.HVG.Varnorm.h5ad`, `*.TP10K.h5ad`, and
`*.Corrected.HVGs.txt`, which `prepare` then consumes.

Harmony correction is slow. If a run crashes after correction but before
factorization, re-run with `resume = TRUE` to reuse the corrected files and skip
straight to factorization:

```{r batch-resume}
Immune <- RunCNMF(Immune, output_dir = "./Figures/cNMF",
                  batch.by = "Sample", k_range = 5:15,
                  resume = TRUE, conda_env = "cnmf_env")
```

---

## 8. Troubleshooting the Python environment

Every error below is an environment / binary-compatibility issue, not a problem
with the data or the wrapper. They are listed in the order you are likely to hit
them.

| Symptom | Cause | Fix |
|---|---|---|
| `cnmf not available ... Active Python: unknown`, with a "use_python ... will be ignored" warning | Python already bound to the wrong interpreter this session | Restart R; set `RETICULATE_PYTHON` to the env's python **before** anything else |
| `py_config()` shows `numpy: [NOT FOUND]` / wrong env | Bound to an env that lacks the packages | Point at the env that actually has cnmf (`conda list -n <env> \| grep cnmf`) |
| `OSError ... libllvmlite.dylib ... Library not loaded: @rpath/libc++.1.dylib` | numba/llvmlite can't find libc++ under R's loader path | `conda install -n <env> -c conda-forge libcxx` |
| `Required version of NumPy not available: incompatible NumPy binary version` | reticulate's R<->NumPy bridge too old for NumPy 2.0 | `install.packages("reticulate")`, or `pip install "numpy<2"` |
| `ValueError: numpy.dtype size changed ... Expected 96 ... got 88` (only with `batch.by`) | `scikit-misc` binary built against a different NumPy than installed | `pip install "numpy<2"` (align the stack), then restart R |
| `AttributeError: 'Harmony' object has no attribute 'Phi_moe'` (only with `batch.by`) | harmonypy version newer than cNMF's Preprocess expects | `pip install "harmonypy==0.0.9"` |
| `cp: ... is not a directory` then `FileNotFoundError: ... .tpm.h5ad` | A **space in `output_dir`** (e.g. Box "Yun lab projects") breaks cNMF's unquoted internal `cp` | Handled automatically: scSidekick runs cNMF through a space-free symlink. If symlink creation fails, use an `output_dir` without spaces |

General rule: bind the env **first** in a fresh R session, and keep the whole
Python stack consistent at `numpy < 2`.

---

## See also

- Vignette 06: `CellTypeAssignmentHelper()` - annotate the cell types you can
  then relate programs to
- Vignette 16: `CorrelateGene()` / `PlotCorrelation()` - gene co-expression, a
  complementary view of program structure
- Vignette 08: `RunGSEA()` - feed `GetCNMFTopGenes()` output in as a gene set to
  characterize each program

---

## Session info

```{r session-info, eval=TRUE}
sessionInfo()
```
