---
title: "scSidekick - Upstream Regulator Analysis"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Upstream Regulator Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  eval    = FALSE,
  message = FALSE,
  warning = FALSE
)
options(cli.width = 80)
```

## Overview

Differential expression tells you *which genes* change; upstream regulator
analysis (URA) asks *which transcription factors* drive those changes.
`RunURA()` infers per-sample TF activity from gene expression using
[decoupleR](https://saezlab.github.io/decoupleR/) and the
[CollecTRI](https://github.com/saezlab/CollecTRI) TF-target network, then
(optionally) tests for **differential TF activity** between conditions.

The workflow:

1. Collapse cells to **pseudobulk** profiles (per group, donor, and any
   stratifier) - statistically sound replicates for downstream testing.
2. Score TF activity per sample with decoupleR's univariate linear model
   (`ulm`, the default; `viper` and `mlm` are also available) against CollecTRI.
3. If a contrast is given, test for differential activity (limma when there are
   at least 3 donors per side, t-test otherwise).
4. Save a TF-activity heatmap and a differential bubble plot, each with a
   `.legend` sidecar.

`RunURA()` complements the pathway tools (Vignettes 08-10): GSEA finds enriched
gene sets, URA finds the regulators upstream of them.

> **decoupleR is an optional dependency.** Install it once (Section 1). The
> CollecTRI network is fetched from OmniPath on first use and cached locally.

---

## 1. Installation

```{r install}
if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
BiocManager::install("decoupleR")
# Optional but recommended for the differential heatmap and limma-based testing:
BiocManager::install(c("limma", "ComplexHeatmap"))
```

```{r libs}
library(scSidekick)
library(Seurat)
```

The first `RunURA()` call downloads CollecTRI from OmniPath and caches it to
`~/.scSidekick/collectri_<organism>.rds`, so subsequent runs are offline and
instant. See the note on network access at the end if OmniPath is blocked on
your network.

---

## 2. Basic TF activity (no contrast)

With just `group.by`, `RunURA()` infers mean TF activity per group and saves a
z-scored activity heatmap of the most variable TFs.

```{r basic}
ura <- RunURA(obj,
  group.by   = "CellType",
  organism   = "human",
  output_dir = "./Figures/URA")
```

The return value is a list:

| Element | Description |
|---|---|
| `activity` | TF x group mean-activity matrix |
| `scores` | Long-format decoupleR output (all pseudobulk samples) |
| `de` | Differential table (`NULL` when no contrast) |
| `network` | The TF-target network actually used |

```{r basic-inspect}
ura$activity[1:5, 1:4]      # TFs x cell types
```

---

## 3. Differential TF activity between conditions

Supply `contrast.by` (the condition column) and `contrast` (a
`c("test", "reference")` pair), plus `donor.by` to enable replicate-based
testing. This produces a differential heatmap (activity difference, with dots
marking `p.adj < 0.05`) and a bubble plot.

```{r contrast}
ura <- RunURA(obj,
  group.by    = "CellType",
  donor.by    = "Donor.ID",            # biological replicates -> limma
  contrast.by = "Diagnosis",
  contrast    = c("AD", "Control"),    # test vs reference
  organism    = "human",
  output_dir  = "./Figures/URA")

head(ura$de)   # TF, cell_type, activity_diff, p.value, p.adj, contrast ...
```

### Multiple contrasts at once

```{r contrast-multi}
ura <- RunURA(obj,
  group.by    = "CellType",
  donor.by    = "Donor.ID",
  contrast.by = "Diagnosis",
  contrast    = list(c("AD", "Control"), c("MCI", "Control")),
  output_dir  = "./Figures/URA")
```

### Stratify with `split.by`

`split.by` (e.g. `"Sex"`) adds a second dimension: the contrast is computed
within each level, columns of the heatmap are grouped by it, and the bubble
plot is faceted.

```{r split}
ura <- RunURA(obj,
  group.by    = "CellType",
  donor.by    = "Donor.ID",
  split.by    = "Sex",
  contrast.by = "Diagnosis",
  contrast    = c("AD", "Control"),
  output_dir  = "./Figures/URA")
```

---

## 4. Useful options

```{r options}
RunURA(obj,
  group.by       = "CellType",
  contrast.by    = "Diagnosis",
  contrast       = c("AD", "Control"),
  method         = "ulm",        # "ulm" (default), "viper", or "mlm"
  n_top          = 30,           # TFs shown in plots
  min_cells      = 10,           # minimum cells per pseudobulk group
  group.levels   = c("Microglia", "Astrocyte"),  # restrict to some groups
  use_pseudobulk = TRUE,         # FALSE = score per cell, then average
  organism       = "human",      # or "mouse"
  output_dir     = "./Figures/URA")
```

- **`use_pseudobulk = TRUE`** (default) is the rigorous route: TF activity is
  inferred from per-sample pseudobulk averages. With `use_pseudobulk = FALSE`,
  decoupleR scores every cell directly and activities are averaged per group
  (simpler, no `donor.by` needed, but cells are pseudoreplicates so interpret DE
  p-values with caution; use `max_cells` to subsample large objects).
- **`force = TRUE`** recomputes pseudobulk even if a cached result is stored in
  `obj@misc$pseudobulk`.
- A custom `network` (a data frame with `source`, `target`, `mor`) can be passed
  instead of CollecTRI.

---

## 5. Outputs

When `output_dir` is set, `RunURA()` writes (file names encode `group.by`,
`split.by`, and the contrast):

- `<object>_URA_by_<group>..._heatmap.pdf` - one per contrast; rows are top TFs,
  columns are cell types (grouped by `split.by`), color is activity difference,
  black dots mark `p.adj < 0.05`.
- `<object>_URA_by_<group>..._bubble.pdf` - matrix bubble plot (cell type x TF,
  color = activity difference, size = -log10 p.adj), faceted by contrast and
  `split.by`.
- A `.legend` sidecar next to each PDF, so `create_analysis_pptx()` picks the
  description up automatically.

---

## 6. Note on network access (OmniPath)

CollecTRI is fetched from the OmniPath web service on first use. `RunURA()`
tries several fallbacks (bundled copy, alternate OmniPathR options, disk cache,
direct REST download) and caches a working network to
`~/.scSidekick/collectri_<organism>.rds`. If your institution blocks OmniPath,
fetch the network once from an unrestricted machine and drop it in the cache:

```{r offline}
net <- decoupleR::get_collectri(organism = "human")
saveRDS(net, "~/.scSidekick/collectri_human.rds")
```

Then every `RunURA()` call uses the cached network offline. You can also pass it
directly via the `network` argument.

---

## See also

- Vignette 08-10: `RunGSEA()` and the pathway family - the enriched gene sets
  that URA finds the regulators for
- Vignette 14: `ComputePseudobulk()` - the donor-level aggregation URA builds on
- Vignette 12: `create_analysis_pptx()` - assembles the URA figures (with their
  `.legend` text) into a deck

---

## Session info

```{r session-info, eval=TRUE}
sessionInfo()
```
