---
title: "scSidekick - Pseudotime & Trajectory Analysis"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Pseudotime & Trajectory Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 10,
  fig.height = 6
)
options(cli.width = 80)
```

## Overview

Six functions cover the pseudotime workflow end-to-end, built directly on
Seurat objects with no `SingleCellExperiment` conversion in sight:

- **`RunSlingshot()`** - fits [Slingshot](https://bioconductor.org/packages/slingshot/)
  directly on any Seurat reduction (PCA, Harmony, or another integrated
  embedding) plus a cluster metadata column. Only the reduction matrix and a
  metadata column are read, so it doesn't matter whether the object has
  BPCells, sketch, or multiple v5 assays, and the expression matrix is never
  touched or re-normalized. Results are written back onto the object so they
  can be reused without recomputation.
- **`SlingshotLineages()`** - a readable summary of what was found: one row
  per lineage, with its cluster path and cell count.
- **`PlotTrajectory()`** - overlays the trajectory on a DimPlot, either as
  straight cluster-to-cluster segments (the clean, DDRTree-like look) or as
  Slingshot's own smoothed principal curves.
- **`PlotPseudotime()`** - density/ridge distribution of pseudotime across
  clusters; works with any numeric pseudotime column, not just Slingshot's.
- **`PlotFeatureTrend()`** - smoothed gene expression (or any numeric
  feature) vs. pseudotime, one facet per feature.
- **`PlotFeatureDistribution()`** - a more general density/ridge plot for any
  single gene or metadata feature, with `group.by`/`split.by`/`facet.by` -
  useful well beyond pseudotime (e.g. checking whether replicate samples
  agree).

We'll walk through all six on `panc8`, the classic multi-technology pancreas
dataset, tracing the ductal-to-endocrine differentiation trajectory.

```{r pkgs, eval=T, message=FALSE, warning=FALSE, include=T}
library(scSidekick)
library(Seurat)
library(SeuratData)
library(harmony)
```

## Preparing the data

`panc8` combines pancreatic islet cells profiled across eight different
technologies (`dataset`) - exactly the kind of batch structure Harmony
integration is for, and a good stress test for whether a trajectory holds up
across technologies rather than being an artifact of one.

We restrict to the ductal and endocrine cell types (`ductal`, `alpha`,
`beta`, `gamma`, `delta`, `epsilon`) so the trajectory reflects the known
endocrine differentiation biology, rather than an uninterpretable path
through unrelated immune/stromal/vascular populations that also happen to be
in this dataset (`activated_stellate`, `endothelial`, `macrophage`, `mast`,
`quiescent_stellate`, `schwann`).

```{r data, eval=T, message=FALSE, warning=FALSE, include=T}
#InstallData("panc8")
data("panc8")
panc8 <- UpdateSeuratObject(panc8)
panc8$donor <- panc8$orig.ident

endocrine_types <- c("ductal", "alpha", "beta", "gamma", "delta", "epsilon")
panc8 <- subset(panc8, subset = celltype %in% endocrine_types)

panc8 <- NormalizeData(panc8, verbose = FALSE)
panc8 <- FindVariableFeatures(panc8, nfeatures = 2000, verbose = FALSE)
panc8 <- ScaleData(panc8, verbose = FALSE)
panc8 <- RunPCA(panc8, npcs = 30, verbose = FALSE)
panc8 <- RunHarmony(panc8, group.by.vars = "dataset", verbose = FALSE)
panc8 <- RunUMAP(panc8, reduction = "harmony", dims = 1:30, verbose = FALSE)

panc8 <- PrepObject(panc8,
  variables   = c("celltype", "dataset", "tech"),
  group.by    = "celltype",
  object_name = "panc8")
```

---

## 1. `RunSlingshot()` - fit the trajectory

We fit on the **Harmony** embedding (not PCA or UMAP) so the trajectory
reflects batch-corrected biology rather than technology-driven structure, and
root it at `"ductal"` - the progenitor-like population endocrine cells
differentiate from.

```{r run-slingshot, eval=T}
panc8 <- RunSlingshot(panc8,
  cluster.by = "celltype",
  reduction  = "harmony",
  start.clus = "ductal",
  progress   = FALSE)
```

A few things worth noting from the message output:

- Harmony here has 30 components; `RunSlingshot()` automatically capped
  fitting to the first 20 (`max_dims`, default `20L`) because principal-curve
  fitting cost scales with dimensionality far more than with cell count -
  pass `dims = 1:30` to use them all, or tune `max_dims`.
- `Slingshot_Pseudotime_1`, `Slingshot_Pseudotime_2`, ... are now columns in
  `meta.data`; the fitted model, cluster MST, and lineage paths live in
  `panc8@misc$slingshot` for the plotting functions below to reuse without
  recomputation.
- For a real (much larger) object, set `progress = TRUE` (the default) to get
  live "still fitting" heartbeat messages from a background process instead
  of a silent wait, and `caffeinate = TRUE` to keep the machine awake for a
  long run.

---

## 2. `SlingshotLineages()` - which lineage is which

```{r lineages, eval=T}
SlingshotLineages(panc8)
```

Two lineages were found, both rooted at `ductal` and both routing through
`alpha` and the very rare `epsilon` population (30 cells in the whole
dataset) before diverging to `delta` -> `beta` or to `gamma`. Take the
`epsilon` branch point as an MST artifact of how few cells that population
has, not a strong biological claim - always sanity-check a lineage's path
against what you know of the biology, not just the algorithm's output.

---

## 3. `PlotTrajectory()` - overlay the trajectory

### Straight segments (default)

The default `style = "straight"` draws the cluster-level minimum spanning
tree as straight, arrowed segments between cluster centroids - legible even
with many clusters, and reusable on *any* plot reduction since only the
MST's cluster connectivity is reused (node positions come from whichever
reduction you plot on).

```{r traj-straight, eval=T, fig.width=8, fig.height=7}
PlotTrajectory(panc8, reduction = "umap", style = "straight")
```

Line *type* (not color) distinguishes the two lineages, so it doesn't
collide with the cluster color legend - see "Lineage" in the legend below
the plot.

### Smoothed curves

`style = "curves"` draws Slingshot's own principal curves instead. These
live in whichever reduction Slingshot was *fit* on (Harmony here), so
`reduction` must match `panc8@misc$slingshot$reduction` - `PlotTrajectory()`
stops with an actionable message rather than silently drawing a curve in the
wrong coordinate space if it doesn't.

```{r traj-curves, eval=T, fig.width=8, fig.height=7}
PlotTrajectory(panc8, reduction = "harmony", style = "curves")
```

### Splitting and faceting

Like every other scSidekick plotting function, `PlotTrajectory()` accepts
`split.by` (and `row.by`, for a two-way grid) - each panel is drawn from an
independent subset of the object, so the trajectory overlay and cluster
centroids are recomputed per panel rather than reusing the pooled fit's
positions. Here, one panel per sequencing technology checks whether the
overlay looks consistent across platforms rather than being driven by one of
them:

```{r traj-split, eval=T, fig.width=14, fig.height=5}
PlotTrajectory(panc8, reduction = "umap", style = "straight",
  split.by = "tech")
```

### Numbered labels

`number_labels = TRUE` swaps the legend and centroid labels for short numeric
codes (`1`, `2`, ...) instead of full level names - handy once names get long
enough to crowd a legend, or when you'd rather label panels by number in a
figure and spell out the key separately in a caption:

```{r traj-numbered, eval=T, fig.width=8, fig.height=7}
PlotTrajectory(panc8, reduction = "umap", style = "straight",
  number_labels = TRUE)
```

---

## 4. `PlotPseudotime()` - distribution across cell types

`plot.type = "density"` (default) draws a density + rug plot of pseudotime
per cell type, with peak labels marking each level's modal pseudotime:

```{r plot-pseudotime, eval=T, fig.width=6, fig.height=5}
PlotPseudotime(panc8, pseudotime.by = "Slingshot_Pseudotime_1")
```

`plot.type = "ridge"` shows the same distributions as ridges instead:

```{r plot-pseudotime-ridge, eval=T, fig.width=6, fig.height=5}
PlotPseudotime(panc8,
  pseudotime.by = "Slingshot_Pseudotime_1",
  plot.type     = "ridge")
```

`split.by` (and `row.by`, for a two-way grid) facets either style into
column/row panels - here, one panel per sequencing technology, to check
whether the pseudotime ordering holds up across platforms:

```{r plot-pseudotime-split, eval=T, fig.width=14, fig.height=4}
PlotPseudotime(panc8,
  pseudotime.by = "Slingshot_Pseudotime_1",
  split.by      = "tech")
```

---

## 5. `PlotFeatureTrend()` - hormone genes across pseudotime

Endocrine cell types are defined by the hormone they secrete: `GCG`
(glucagon, alpha), `INS` (insulin, beta), `SST` (somatostatin, delta), and
`PPY` (pancreatic polypeptide, gamma). Coloring by `dataset` shows whether
each technology traces a similar expression trend along pseudotime.

```{r feature-trend, eval=T, fig.width=10, fig.height=8}
PlotFeatureTrend(panc8,
  pseudotime.by = "Slingshot_Pseudotime_1",
  features      = c("GCG", "INS", "SST", "PPY"),
  color.by      = "dataset")
```

All four hormone genes rise together around the same point of pseudotime
(where the endocrine populations sit) for most technologies - except
`fluidigmc1`, which stays flat. That's a real, known property of this
dataset: Fluidigm C1 capture has much lower sensitivity for these
transcripts than the other platforms, not a biological difference. This is
exactly the kind of batch artifact `color.by` is meant to surface before you
draw a biological conclusion from a pseudotime trend.

`features` accepts genes and numeric `meta.data` columns interchangeably
(module scores, cNMF usages, QC metrics, ...), and extraction goes through
the same assay/version-safe path (`.nk_feature_matrix()`/`.get_layer_data()`)
as every other feature-plotting function in the package.

When `color.by` is left `NULL`, it falls back to `group.by` for coloring -
matching the rest of the package's default - while `group.by` still doubles
as a cell-subset filter (paired with `groups`) regardless of which variable
ends up coloring the plot. `split.by`/`row.by` facet the whole thing into
column/row panels, one independent trend fit per panel:

```{r feature-trend-split, eval=T, fig.width=10, fig.height=8}
PlotFeatureTrend(panc8,
  pseudotime.by = "Slingshot_Pseudotime_1",
  features      = "INS",
  group.by      = "celltype",
  split.by      = "tech")
```

---

## 6. `PlotFeatureDistribution()` - ridge/density for any feature

A more general version of Seurat's `RidgePlot()` - one ridge per
`group.by` level, for any single gene or numeric feature - but with a
`split.by` fill *within* each ridge and a `facet.by` panel split, which
`RidgePlot()` doesn't support. Here: one ridge per cell type, one panel per
technology, to see whether `INS` (insulin) distinguishes `beta` cells
consistently across all eight technologies.

```{r feature-dist, eval=T, fig.width=14, fig.height=5}
PlotFeatureDistribution(panc8,
  feature  = "INS",
  group.by = "celltype",
  facet.by = "dataset",
  style    = "ridge")
```

`beta` separates cleanly from the other cell types in every technology,
which is reassuring - the `INS` signal isn't a single-platform artifact. Use
`style = "density"` for overlaid curves instead of ridges, or swap
`facet.by` for `"donor"` to check individual-sample agreement rather than
whole-technology agreement.

---

## Saving figures

Every plotting function above follows the same save convention as the rest
of the package: pass `output_dir` (or store one with `PrepObject()`) and a
`.legend` sidecar is written automatically alongside the PDF, describing
cell counts, the trajectory rooting, and what's plotted - so a figure never
travels without its own methods description.

```{r save-example, eval=FALSE}
PlotTrajectory(panc8,
  reduction  = "umap",
  style      = "straight",
  output_dir = "./Figures")
```

`file_name`, `width`, and `height` override the auto-built name and
auto-calculated PDF size when you need one; `timestamp = TRUE` appends a
`_YYYYMMDD-HHMMSS` stamp to the saved file name so repeated runs are versioned
instead of overwriting the previous PDF.

`RunSlingshot()` additionally writes `analysis_params.json` documenting the
fit itself (reduction, dims used, root cluster, lineages found) whenever
`output_dir` is available - merged into any existing json rather than
overwriting it, so it sits alongside whatever earlier steps (QC,
clustering, GSEA, ...) already logged there.

---

## See also

- Vignette 02: `RunHarmony`-based integration workflow this trajectory builds on
- Vignette 03: `PlotDimPlots()` - the UMAP grid `PlotTrajectory()` matches visual style with
- Vignette 16: `CorrelateGene()` - co-expression analysis, a complementary way to explore gene relationships within a cell type

---

## Session info

```{r session-info}
sessionInfo()
```
