---
title: "scSidekick - Pseudotime & Trajectory Analysis"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Pseudotime & Trajectory Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width  = 10,
  fig.height = 6
)
options(cli.width = 80)
```

## Overview

Six functions cover the pseudotime workflow end-to-end, built directly on
Seurat objects with no `SingleCellExperiment` conversion in sight:

- **`RunSlingshot()`** - fits [Slingshot](https://bioconductor.org/packages/slingshot/)
  directly on any Seurat reduction (PCA, Harmony, or another integrated
  embedding) plus a cluster metadata column. Only the reduction matrix and a
  metadata column are read, so it doesn't matter whether the object has
  BPCells, sketch, or multiple v5 assays, and the expression matrix is never
  touched or re-normalized. Results are written back onto the object so they
  can be reused without recomputation.
- **`SlingshotLineages()`** - a readable summary of what was found: one row
  per lineage, with its cluster path and cell count.
- **`PlotTrajectory()`** - overlays the trajectory on a DimPlot, either as
  straight cluster-to-cluster segments (the clean, DDRTree-like look) or as
  Slingshot's own smoothed principal curves.
- **`PlotPseudotime()`** - density/ridge distribution of pseudotime across
  clusters; works with any numeric pseudotime column, not just Slingshot's.
- **`PlotFeatureTrend()`** - smoothed gene expression (or any numeric
  feature) vs. pseudotime, one facet per feature.
- **`PlotFeatureDistribution()`** - a more general density/ridge plot for any
  single gene or metadata feature, with `group.by`/`split.by`/`facet.by` -
  useful well beyond pseudotime (e.g. checking whether replicate samples
  agree).

We'll walk through all six on `panc8`, the classic multi-technology pancreas
dataset, tracing the ductal-to-endocrine differentiation trajectory.

```{r pkgs, eval=T, message=FALSE, warning=FALSE, include=T}
library(scSidekick)
library(Seurat)
library(SeuratData)
library(harmony)
```

## Preparing the data

`panc8` combines pancreatic islet cells profiled across eight different
technologies (`dataset`) - exactly the kind of batch structure Harmony
integration is for, and a good stress test for whether a trajectory holds up
across technologies rather than being an artifact of one.

We restrict to the ductal and endocrine cell types (`ductal`, `alpha`,
`beta`, `gamma`, `delta`, `epsilon`) so the trajectory reflects the known
endocrine differentiation biology, rather than an uninterpretable path
through unrelated immune/stromal/vascular populations that also happen to be
in this dataset (`activated_stellate`, `endothelial`, `macrophage`, `mast`,
`quiescent_stellate`, `schwann`).

```{r data, eval=T, message=FALSE, warning=FALSE, include=T}
#InstallData("panc8")
data("panc8")
panc8 <- UpdateSeuratObject(panc8)
panc8$donor <- panc8$orig.ident

endocrine_types <- c("ductal", "alpha", "beta", "gamma", "delta", "epsilon")
panc8 <- subset(panc8, subset = celltype %in% endocrine_types)

panc8 <- NormalizeData(panc8, verbose = FALSE)
panc8 <- FindVariableFeatures(panc8, nfeatures = 2000, verbose = FALSE)
panc8 <- ScaleData(panc8, verbose = FALSE)
panc8 <- RunPCA(panc8, npcs = 30, verbose = FALSE)
panc8 <- RunHarmony(panc8, group.by.vars = "dataset", verbose = FALSE)
panc8 <- RunUMAP(panc8, reduction = "harmony", dims = 1:30, verbose = FALSE)

panc8 <- PrepObject(panc8,
  variables   = c("celltype", "dataset"),
  group.by    = "celltype",
  object_name = "panc8")
```

---

## 1. `RunSlingshot()` - fit the trajectory

We fit on the **Harmony** embedding (not PCA or UMAP) so the trajectory
reflects batch-corrected biology rather than technology-driven structure, and
root it at `"ductal"` - the progenitor-like population endocrine cells
differentiate from.

```{r run-slingshot, eval=T}
panc8 <- RunSlingshot(panc8,
  cluster.by = "celltype",
  reduction  = "harmony",
  start.clus = "ductal",
  progress   = FALSE)
```

A few things worth noting from the message output:

- Harmony here has 30 components; `RunSlingshot()` automatically capped
  fitting to the first 20 (`max_dims`, default `20L`) because principal-curve
  fitting cost scales with dimensionality far more than with cell count -
  pass `dims = 1:30` to use them all, or tune `max_dims`.
- `Slingshot_Pseudotime_1`, `Slingshot_Pseudotime_2`, ... are now columns in
  `meta.data`; the fitted model, cluster MST, and lineage paths live in
  `panc8@misc$slingshot` for the plotting functions below to reuse without
  recomputation.
- For a real (much larger) object, set `progress = TRUE` (the default) to get
  live "still fitting" heartbeat messages from a background process instead
  of a silent wait, and `caffeinate = TRUE` to keep the machine awake for a
  long run.

---

## 2. `SlingshotLineages()` - which lineage is which

```{r lineages, eval=T}
SlingshotLineages(panc8)
```

Two lineages were found, both rooted at `ductal` and both routing through
`alpha` and the very rare `epsilon` population (30 cells in the whole
dataset) before diverging to `delta` -> `beta` or to `gamma`. Take the
`epsilon` branch point as an MST artifact of how few cells that population
has, not a strong biological claim - always sanity-check a lineage's path
against what you know of the biology, not just the algorithm's output.

---

## 3. `PlotTrajectory()` - overlay the trajectory

### Straight segments (default)

The default `style = "straight"` draws the cluster-level minimum spanning
tree as straight, arrowed segments between cluster centroids - legible even
with many clusters, and reusable on *any* plot reduction since only the
MST's cluster connectivity is reused (node positions come from whichever
reduction you plot on).

```{r traj-straight, eval=T, fig.width=8, fig.height=7}
PlotTrajectory(panc8, reduction = "umap", style = "straight")
```

Line *type* (not color) distinguishes the two lineages, so it doesn't
collide with the cluster color legend - see "Lineage" in the legend below
the plot.

### Smoothed curves

`style = "curves"` draws Slingshot's own principal curves instead. These
live in whichever reduction Slingshot was *fit* on (Harmony here), so
`reduction` must match `panc8@misc$slingshot$reduction` - `PlotTrajectory()`
stops with an actionable message rather than silently drawing a curve in the
wrong coordinate space if it doesn't.

```{r traj-curves, eval=T, fig.width=8, fig.height=7}
PlotTrajectory(panc8, reduction = "harmony", style = "curves")
```

---

## 4. `PlotPseudotime()` - distribution across cell types

Density + rug plot of pseudotime per cell type (with peak labels), plus an
optional ridge plot split by a second variable - here, `dataset`, to check
whether the pseudotime ordering holds up across technologies.

```{r plot-pseudotime, eval=T, fig.width=12, fig.height=5}
PlotPseudotime(panc8,
  pseudotime.by = "Slingshot_Pseudotime_1",
  split.by      = "dataset")
```

---

## 5. `PlotFeatureTrend()` - hormone genes across pseudotime

Endocrine cell types are defined by the hormone they secrete: `GCG`
(glucagon, alpha), `INS` (insulin, beta), `SST` (somatostatin, delta), and
`PPY` (pancreatic polypeptide, gamma). Coloring by `dataset` shows whether
each technology traces a similar expression trend along pseudotime.

```{r feature-trend, eval=T, fig.width=10, fig.height=8}
PlotFeatureTrend(panc8,
  pseudotime.by = "Slingshot_Pseudotime_1",
  features      = c("GCG", "INS", "SST", "PPY"),
  color.by      = "dataset")
```

All four hormone genes rise together around the same point of pseudotime
(where the endocrine populations sit) for most technologies - except
`fluidigmc1`, which stays flat. That's a real, known property of this
dataset: Fluidigm C1 capture has much lower sensitivity for these
transcripts than the other platforms, not a biological difference. This is
exactly the kind of batch artifact `color.by` is meant to surface before you
draw a biological conclusion from a pseudotime trend.

`features` accepts genes and numeric `meta.data` columns interchangeably
(module scores, cNMF usages, QC metrics, ...), and extraction goes through
the same assay/version-safe path (`.nk_feature_matrix()`/`.get_layer_data()`)
as every other feature-plotting function in the package.

---

## 6. `PlotFeatureDistribution()` - ridge/density for any feature

A more general version of Seurat's `RidgePlot()` - one ridge per
`group.by` level, for any single gene or numeric feature - but with a
`split.by` fill *within* each ridge and a `facet.by` panel split, which
`RidgePlot()` doesn't support. Here: one ridge per cell type, one panel per
technology, to see whether `INS` (insulin) distinguishes `beta` cells
consistently across all eight technologies.

```{r feature-dist, eval=T, fig.width=14, fig.height=5}
PlotFeatureDistribution(panc8,
  feature  = "INS",
  group.by = "celltype",
  facet.by = "dataset",
  style    = "ridge")
```

`beta` separates cleanly from the other cell types in every technology,
which is reassuring - the `INS` signal isn't a single-platform artifact. Use
`style = "density"` for overlaid curves instead of ridges, or swap
`facet.by` for `"donor"` to check individual-sample agreement rather than
whole-technology agreement.

---

## Saving figures

Every plotting function above follows the same save convention as the rest
of the package: pass `output_dir` (or store one with `PrepObject()`) and a
`.legend` sidecar is written automatically alongside the PDF, describing
cell counts, the trajectory rooting, and what's plotted - so a figure never
travels without its own methods description.

```{r save-example, eval=FALSE}
PlotTrajectory(panc8,
  reduction  = "umap",
  style      = "straight",
  output_dir = "./Figures")
```

`RunSlingshot()` additionally writes `analysis_params.json` documenting the
fit itself (reduction, dims used, root cluster, lineages found) whenever
`output_dir` is available - merged into any existing json rather than
overwriting it, so it sits alongside whatever earlier steps (QC,
clustering, GSEA, ...) already logged there.

---

## See also

- Vignette 02: `RunHarmony`-based integration workflow this trajectory builds on
- Vignette 03: `PlotDimPlots()` - the UMAP grid `PlotTrajectory()` matches visual style with
- Vignette 16: `CorrelateGene()` - co-expression analysis, a complementary way to explore gene relationships within a cell type

---

## Session info

```{r session-info}
sessionInfo()
```
