---
title: "scSidekick - AskSidekick(): the AI Documentation Assistant"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{AskSidekick - AI Documentation Assistant}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  eval    = FALSE,   # every call here hits a live API and needs a key -
                      # nothing in this vignette is meant to run during
                      # R CMD build/check.
  message = FALSE,
  warning = FALSE
)
options(cli.width = 80)
```

## Overview

`AskSidekick()` is a small, self-contained help assistant for scSidekick
itself. Instead of digging through `?FunctionName` or the vignette index,
you can just ask:

```r
library(scSidekick)
AskSidekick("Which function do I use to split a UMAP by two variables at once?")
```

It answers four kinds of questions particularly well:

1. **"How do I recreate this figure?"** - point it at a saved PDF or a
   screenshot with `image_path` and it reads off the plot type and axis/legend
   labels, matches the file's own path against how scSidekick names its
   outputs, and proposes the call that likely made it.
2. **"Which column in my object is Age?"** - pass your `seurat_object` and it
   reads the real `meta.data` column names (and any colors already saved by
   `PrepObject()`) instead of guessing from a generic naming convention.
3. **"How do I change these colors?"** - grounded answers about `PrepObject()`,
   `custom_colors`, `GetColors()`, etc., pulled from the actual installed
   documentation rather than a fixed FAQ that can drift out of date.
4. **"Which covariates should I adjust for?"** - general how-to questions about
   any exported function, answered from that function's real man page.

Every answer is grounded in **your installed copy's own documentation** - man
pages plus vignette sources - so it never quotes an argument name or default
that doesn't actually exist in the version you have installed.

---

## Setup

`AskSidekick()` uses the [ellmer](https://ellmer.tidyverse.org/) package to
talk to a large language model. `ellmer` is a `Suggests`-only dependency -
scSidekick loads and works normally without it - so install it once:

```r
install.packages("ellmer")
```

By default `AskSidekick()` talks to Google Gemini, which has a genuinely free
tier (no credit card required). Get a key at
<https://aistudio.google.com/apikey>, then add one line to `~/.Renviron`
(`usethis::edit_r_environ()` opens it for you) and restart R:

```
GEMINI_API_KEY=your_key_here
```

That's the whole setup. No account tier to choose, no billing to configure.

---

## 1. A plain question

No Seurat object needed for a general "how do I..." question:

```r
AskSidekick("How do I split a UMAP by donor and color by cell type in the same figure?")
```

```
Use PlotDimPlots() with `group.by` for color and `split.by` for the panel split:

scSidekick::PlotDimPlots(seurat_object,
  group.by = "cell_type",
  split.by = "donor")
```

Each call is independent - `AskSidekick()` doesn't remember earlier questions
in the same session. That's deliberate: Gemini's free tier has a fairly tight
per-minute token budget, and re-sending a growing conversation history would
burn through it in one or two calls. Ask one self-contained question at a
time.

---

## 2. Passing your own object: "which column is Age?"

Supply `seurat_object` and the assistant sees your actual `meta.data` column
names (and any colors already registered via `PrepObject()`) instead of
guessing from common naming conventions:

```r
AskSidekick("Which metadata column represents the donor's age?",
            seurat_object = SeuratObj)
```

```
Looking at SeuratObj's meta.data, "Age.at.Death" (numeric) is the column that
represents age:

scSidekick::PlotFeature(SeuratObj, features = "Age.at.Death",
  group.by = "Diagnosis")
```

Notice the reply uses `SeuratObj` - your own variable's name - in the code
snippet, not a generic placeholder like `seurat_obj`.

---

## 3. Recreating a figure from an image

Point `image_path` at a saved PDF or a screenshot of a figure someone sent
you, and `AskSidekick()` reads off the plot type, axis/legend labels, and the
file's own path (scSidekick's output folders and filenames are deterministic,
not just cosmetic - e.g. only `RunSCssGSEA()` creates a folder starting `"SC
ssGSEA "`) to identify the function that most likely made it:

```r
AskSidekick(image_path = "Figures/bmcite_celltype.l1_donor_GroupHeatmap.pdf",
            seurat_object = SeuratObj)
```

```
The filename matches GroupHeatmap()'s own naming convention (<object>_<group
.by>_<split.by>_GroupHeatmap.pdf). Based on the visible row/column labels:

scSidekick::GroupHeatmap(SeuratObj,
  group.by        = "celltype.l1",
  column_split_by = "donor")
```

You can add your own question alongside the image if you want something more
specific than "recreate this":

```r
AskSidekick("What are the top marker genes labeled on this heatmap?",
            image_path = "Figures/markers_heatmap.pdf")
```

### `same_dataset`: is this figure actually from this object?

When you pass both `image_path` and `seurat_object`, `same_dataset` tells the
assistant how confident to be about matching the figure's grouping variables
to your object's real columns:

- **`same_dataset = TRUE`** - "this figure was made from this exact object."
  The assistant is told an exact column match *must* exist for every
  grouping/coloring variable shown, and to search carefully among
  similar-but-not-identical columns (`Cluster` vs. `seurat_clusters` vs.
  `Assignment`) rather than settling for the first plausible name.
- **`same_dataset = FALSE`** (or the default `NA`) - the figure may be from a
  different dataset, an earlier package version, or a collaborator's file.
  The assistant gives its best estimate and says explicitly that it's an
  estimate, rather than implying a confirmed match.

```r
AskSidekick(image_path    = "Figures/old_project_umap.pdf",
            seurat_object = SeuratObj,
            same_dataset  = TRUE)
```

Either way, after the model replies, `AskSidekick()` independently
double-checks anything that looks like a gene list (`c("Cd3e", "Cd8a", ...)`)
against `rownames(SeuratObj)` and anything that looks like a metadata column
(`group.by = "..."`, `stat.by = "..."`, etc.) against
`colnames(SeuratObj@meta.data)` - deterministically, not by asking the model
to grade its own work. If either check comes back thin (e.g. a species
mismatch between Title-case mouse genes and ALL-CAPS human genes), a
**"Heads up"** note is appended to the reply automatically, telling you
exactly which names didn't match and the closest real alternatives.

---

## 4. The first call after install (or after an update)

The assistant's documentation context - man pages plus vignette sources - is
built from your **installed** copy of scSidekick and cached under
`tools::R_user_dir("scSidekick", "cache")`, keyed by package version. The very
first `AskSidekick()` call after installing or updating the package rebuilds
that cache automatically:

```
scSidekick: setting up AskSidekick() for the first time with this package
version (v1.2.0) - hang on, this can take a minute or two. It only happens
once; every call after this is instant.
scSidekick: ready (120 documentation chunks cached).
```

Every call after that reuses the cache silently - no bundled asset to ship
with the package, and no manual "rebuild the docs" step on release day.

If you're actively developing scSidekick itself (editing a man page or
vignette without bumping the version number), force a rebuild directly:

```r
RebuildSidekickContext()
```

---

## 5. Model fallback

Google Gemini's free tier gives each model its own separate daily quota
(RPD), and the allowance varies a lot by model - "Flash Lite" models
typically get a much larger daily allowance than flagship Flash models on the
same account (check <https://aistudio.google.com/rate-limit> for yours).
`AskSidekick()` tries the models in `models` (a "Flash Lite" model first, by
default) and automatically moves to the next one on a quota or server error,
rather than making you wait out a 429:

```
scSidekick bot: gemini-3.5-flash-lite unavailable right now (RESOURCE_EXHAUSTED),
trying the next model...
scSidekick bot: answered using gemini-3.1-flash-lite (earlier model(s) were
over today's quota).
```

That fallback is always printed, never silent. Pass your own `models` vector
to reorder or restrict which models are tried:

```r
AskSidekick("...", models = c("gemini-2.5-flash"))
```

### Using a different provider entirely

`AskSidekick()` is built on [ellmer](https://ellmer.tidyverse.org/), which is
provider-agnostic - Gemini's free tier is simply a convenient default that
doesn't require billing setup. There's no built-in argument to swap providers
today, but because the whole assistant is under 700 lines in
`R/ask_sidekick.R`, pointing it at Anthropic, OpenAI, or a local Ollama model
instead is a matter of swapping the single `ellmer::chat_google_gemini(...)`
call for `ellmer::chat_anthropic(...)`, `ellmer::chat_openai(...)`, or
`ellmer::chat_ollama(...)` - see the `ellmer` documentation for each
provider's exact arguments.

---

## Parameter reference

| Parameter | Type | Default | Description |
|-----------|------|---------|-------------|
| `question` | character | `NULL` | What you want to ask. Optional when `image_path` is supplied. |
| `seurat_object` | Seurat object | `NULL` | Gives the assistant your real `meta.data` columns and any `PrepObject()`-saved colors; also enables the post-reply gene/column match checks. |
| `image_path` | character | `NULL` | Path to a PNG/JPEG/PDF figure to read and reverse-engineer. |
| `same_dataset` | logical | `NA` | `TRUE` = demand an exact metadata-column match (the figure is confirmed to be from `seurat_object`); `FALSE`/`NA` = best estimate, flagged as such. |
| `models` | character vector | Flash Lite + flagship Flash mix | Gemini models to try in order; automatic fallback on quota/server errors. |
| `rebuild_context` | logical | `FALSE` | Force a documentation-cache rebuild before asking. Rarely needed - see `RebuildSidekickContext()`. |

---

## See also

- `RebuildSidekickContext()` - force a documentation-cache rebuild
- Vignette 01: Getting Started - `PrepObject()` is what `AskSidekick()` reads
  columns and colors from when you pass a `seurat_object`
- Vignette 12: Reporting & Utilities - the rest of scSidekick's
  reproducibility layer (`.legend` sidecars, `ExtractMethods()`, the session
  logger)

---

## Session info

```{r session-info, eval=TRUE}
sessionInfo()
```
